---
title: "Module - Clustering"
description: | 
  Description
execute:
  freeze: auto
  cache: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: 'Test'
  seurat_object: './data/project_reduction_object-5.RDS'
  input_features_plot: 'LYZ;CCL5;IL32;PTPRCAP;FCGR3A;PF4;PTPRC'
  input_group_plot: 'patient_id;timepoint'
  input_integration_dimension: 'auto'
  thr_quantile: 'q01'
  thr_npc: 'auto'
  thr_resolution: 0.25
  thr_proportion: 0.25
  n_threads: 8
  n_memory: 8
  work_directory: !expr here::here()
  auto_save: TRUE
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE, message = FALSE, warning = FALSE, echo = FALSE}
#| include: false

# Project parameters 
project_name <- params$project_name
seurat_object <- params$seurat_object

# Inputs and thresholds
input_integration_dimension <- params$input_integration_dimension
input_features_plot <- strsplit(
  params$input_features_plot, split = ';')[[1]]
input_group_plot <- strsplit(
  params$input_group_plot, split = ';')[[1]]

thr_npc <- params$thr_npc
thr_resolution <- params$thr_resolution
thr_proportion <- params$thr_proportion
thr_quantile <- params$thr_quantile

# Dataflow/Computational parameters
n_threads <- params$n_threads
n_memory <- params$n_memory

# Output parameters
work_directory <- params$work_directory
timestamp <- params$timestamp
auto_save <- params$auto_save

# Knitr options
knitr::opts_knit$set(
  root.dir = work_directory
  )

```

## Loading library

```{r library}
#| echo: false

library(readr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(Seurat)
library(BPCells)
library(ROGUE)
library(SCP)

```

## Folder structure

```{r project_directory}

if(!dir.exists(work_directory)) {
  dir.create(work_directory, recursive = T)
}

# showWarnings = FALSE: data/ already exists (the BPCells store is staged there).
for(sub_directory in c('data', 'figures')) {
  dir.create(paste0(work_directory, '/', sub_directory), showWarnings = FALSE)
}

```

## Loading Seurat object

```{r loading_rds}
#| code-fold: true

# Reading the reduction object. It is BPCells-backed: its counts/data layers are
# lazy transforms over the on-disk store staged at data/bpcells_counts, so
# BPCells must be loaded (above) for the object to deserialize and resolve.
seurat_object <- readRDS(file = seurat_object)

# Cell number
ncell <- dim(seurat_object)[2]

# Knowledge-based parameters
if(thr_npc == 'auto') {
  thr_npc <- ifelse(ncell < 50000, 25, 50)
}

# Setting random seed
random_seed <- 2203
set.seed(random_seed)

```

## Unifying sample layers

```{r seurat_rejoin}
#| code-fold: true

if(length(seurat_object@assays$RNA@layers) > 1) {
  
  seurat_object <- JoinLayers(seurat_object)  

}

```

## Input validation

```{r basic_input_validation}

# Extracting gene names and metadata columns
metadata_columns <- colnames(seurat_object@meta.data)
features_names <- row.names(seurat_object)

# Checking input variables
input_features_plot <- intersect(input_features_plot, features_names)
input_group_plot <- intersect(input_group_plot, metadata_columns)

# Setting reduction variable
integration_method <- function(
    method = c('pca', 'integrated.mnn', 'integrated.rpca', 'integrated.cca', 'harmony', 'auto')) {

  # Default reduction
  reduction_variable <- 'pca'
  
  # Programmatic choices
  if(method == 'auto') {
    if(file.exists('./batch_method.selected.txt')) {
      reduction_variable <- read_lines(file = "./batch_method.selected.txt")
    } 
  }
  
  if(method %in% names(seurat_object@reductions)) {
    reduction_variable <- method
  } 
  
  return(reduction_variable)
  
}

reduction_variable <- integration_method(method = input_integration_dimension)

```

::: {.callout-warning}
The clustering step will be performed over the **`r reduction_variable`** reduction.
:::

## Finding neighbors and clusters

```{r finding_neighbors}
#| code-fold: true
#| warning: false

# Finding clusters
seurat_object <- FindNeighbors(
  object = seurat_object, 
  reduction = reduction_variable, 
  dims = 1:thr_npc,
  graph.name = paste0('snn_npc', thr_npc)
)

seurat_object <- FindClusters(
  object = seurat_object,
  graph.name = paste0('snn_npc', thr_npc),
  resolution = thr_resolution
)

# Running UMAP reduction
seurat_object <- RunUMAP(
  object = seurat_object,
  reduction = reduction_variable, 
  dims = 1:thr_npc,
  seed.use = random_seed
)

```

```{r umap_main_plot, fig.height = 12}
#| label: fig-umap-plot
#| fig-cap: 'UMAP'
#| fig-cap-location: top
#| fig-height: 10
#| fig-width: 10

# Plotting UMAP based on Seurat clusters
SCP::CellDimPlot(
  srt = seurat_object, 
  group.by = "seurat_clusters",
  reduction = "umap",
  label = TRUE, 
  label_repel = TRUE,
  show_stat = TRUE,
  theme_use = "theme_blank"
)

```

:::{.callout-tip}
The @fig-umap-plot presentind the overall clustering after running Leiden algorithm.
:::

```{r clusters_composition}

seurat_cluster_composition <- seurat_object@meta.data %>%
  group_by(seurat_clusters, orig.ident) %>%
  summarise(
    n_cell = n()
  )

seurat_cluster_composition <- seurat_cluster_composition %>%
  group_by(seurat_clusters) %>%
  mutate(
    total_cell = sum(n_cell),
    percentage = round((n_cell / total_cell) * 100, digits = 2)
  )

```

```{r cluster-purity-compute, include = FALSE}
# seurat_cluster_composition (computed above) already holds the per-cluster,
# per-sample percentages. Nothing in this notebook ever drew a conclusion from
# them, so a cluster that is 96% one sample -- i.e. a batch artefact rather
# than a cell type -- rendered as an unremarkable bar in the stacked plot.
suppressPackageStartupMessages({library(ggplot2); library(data.table)})
.cc <- as.data.table(seurat_cluster_composition)
.thr <- 90   # % of a cluster coming from one sample before it looks sample-driven

.n_samples <- length(unique(.cc$orig.ident))
.purity <- .cc[, .(top_sample = orig.ident[which.max(percentage)],
                   top_pct    = max(percentage),
                   cells      = total_cell[1]), by = seurat_clusters][order(-top_pct)]

# A cluster being mostly one sample is only suspicious relative to how big that
# sample is overall: if a sample contributes 80% of all cells, an 80% cluster is
# exactly what you would expect. Enrichment corrects for that.
.cohort <- .cc[, .(cohort_pct = 100 * sum(n_cell) / sum(.cc$n_cell)), by = orig.ident]
.purity <- merge(.purity, .cohort, by.x = "top_sample", by.y = "orig.ident", all.x = TRUE)
.purity[, enrichment := round(top_pct / pmax(1e-9, cohort_pct), 2)]
setorder(.purity, -top_pct)
.skewed <- .purity[top_pct >= .thr & enrichment >= 1.25]
```

```{r cluster-purity-finding, results = 'asis'}
if (.n_samples < 2) {
  scratch_finding(
    "Only one sample is present, so cluster composition cannot distinguish biology from batch.",
    type = "note", title = "Cluster composition")
} else {
  msg <- sprintf(paste("Across **%d clusters** from **%d samples**, the most",
                       "sample-dominated cluster is **cluster %s** at **%.1f%%** from",
                       "`%s` (%.2fx that sample's cohort-wide share)."),
                 nrow(.purity), .n_samples, as.character(.purity$seurat_clusters[1]),
                 .purity$top_pct[1], .purity$top_sample[1], .purity$enrichment[1])
  if (nrow(.skewed)) {
    msg <- paste(msg, sprintf(
      paste("**%d cluster(s) (%s) are at least %d%% single-sample and enriched above",
            "their share of the cohort.** Clusters like these usually reflect batch or",
            "donor identity rather than a distinct cell state, and any downstream",
            "result keyed on cluster identity -- markers, composition, cell-cell",
            "communication -- will inherit that. Check them against the batch-corrected",
            "embedding before interpreting them biologically."),
      nrow(.skewed), paste(.skewed$seurat_clusters, collapse = ", "), .thr))
  } else {
    msg <- paste(msg, sprintf(
      "No cluster exceeds %d%% single-sample membership with enrichment, so clustering does not appear to be driven by sample identity.",
      .thr))
  }
  scratch_finding(msg, type = if (nrow(.skewed)) "caution" else "note",
                  title = "Is clustering driven by sample identity?")
  scratch_table(.purity[, .(cluster = seurat_clusters, cells,
                            `top sample` = top_sample,
                            `% from top sample` = round(top_pct, 1),
                            `sample's cohort share %` = round(cohort_pct, 1),
                            enrichment)],
                caption = "Cluster composition, most sample-dominated first")
}
```

```{r cluster-purity-figure, message = FALSE, warning = FALSE}
#| fig-width: 9
#| fig-height: 5
if (.n_samples >= 2) {
  .hm <- copy(.cc)
  .hm[, tip := sprintf("cluster %s / %s\n%.1f%% of the cluster (%s cells)",
                       seurat_clusters, orig.ident, percentage, format(n_cell, big.mark = ","))]
  ggplot(.hm, aes(factor(seurat_clusters), orig.ident, fill = percentage, text = tip)) +
    geom_tile(colour = "white", linewidth = .4) +
    scale_fill_gradient(low = "#eef4fd", high = "#0d366b", limits = c(0, 100),
                        name = "% of\ncluster") +
    labs(title = "Sample composition of each cluster",
         subtitle = sprintf("a column dominated by one row is a cluster made of one sample (flagged at %d%%)", .thr),
         x = "cluster", y = NULL) +
    theme_scratch()
}
```

```{r barplot_composition}
#| label: fig-bar-composition
#| fig-cap: 'Barplot'
#| fig-width: 18
#| fig-height: 10
#| fig-align: center

# Boxplot
SCP::CellStatPlot(
  seurat_object,
  stat.by = "orig.ident", 
  group.by = "seurat_clusters", 
  label = TRUE
)

```

:::{.callout-tip}
The @fig-bar-composition provides an overall view of cell composition across distinct clusters on the dataset.
:::

```{r printing_cluster_table, message = FALSE, warning = FALSE, echo = FALSE}

seurat_cluster_composition %>%
  distinct(seurat_clusters, total_cell) %>%
  tidyr::pivot_wider(names_from = seurat_clusters, values_from = total_cell)

rogue_value <- "Skipped"

```

## Clustering performance {.tabset}

```{r clustering_performance}
#| code-fold: true
#| warning: false
#| eval: false

# Sampling cells per cluster
sample_clusters_members <- seurat_object@meta.data %>%
  mutate(barcode = row.names(seurat_object@meta.data)) %>%
  group_by(seurat_clusters) %>%
  slice_sample(prop = thr_proportion)

sample_clusters_members <- sample_clusters_members %>%
  dplyr::rename(
    Patient = patient_id,
    clusters = seurat_clusters
  )

# Extracting expression counts
expression_counts <- GetAssayData(seurat_object, layer = "counts")

# Sampling counts
expression_counts <- expression_counts[, sample_clusters_members$barcode] %>%
  as.matrix()

expression_counts <- matr.filter(expression_counts, min.cells = 10, min.genes = 10)

# Computing entropy background
entropy_result <- SE_fun(expression_counts)

# Avg. ROGUE score across clusters
rogue_value <- CalculateRogue(entropy_result, platform = "UMI")

# Calculating ROGUE score
rogue_result <- rogue(expression_counts, 
                      labels = sample_clusters_members$clusters,
                      samples = sample_clusters_members$Patient,
                      platform = "UMI",
                      span = 0.6
                      )

rogue_result

```

::: {.callout-warning}
The ROGUE score for the clustering composition (resolution = **`r thr_resolution`**) is
**`r rogue_value`**: the block above is `eval: false`, so no purity score was computed
for this run.

ROGUE calls `as.matrix()` on the sampled counts, which materialises a dense matrix —
the one thing the on-disk BPCells path exists to avoid. Re-enabling it needs a
cluster-wise downsample small enough to densify safely, not a straight `eval: true`.
:::

```{r rogue_removed_note, include = FALSE}
# The SEplot and ROGUE-boxplot sections were removed with the BPCells port. Their
# chunks were `eval: false` alongside the block above, so they emitted no figure
# while their prose still wrote @fig-se-plot / @fig-rogue-plot -- unresolvable
# cross-references in every rendered clustering report. The plotting code lives
# with ROGUE itself; restore both together.
#
# Kept as an include:FALSE chunk, not a markdown comment: an HTML comment in the
# prose is passed through to the rendered page and shows up in its text.
```

## Data visualization {.tabset}

### Sample-based UMAP

```{r umap_sample_plot, fig.height = 12, fig.width = 14}
#| label: fig-umap-sample
#| fig-cap: 'UMAP'
#| fig-width: 15
#| fig-height: 10
#| fig-align: center

SCP::CellDimPlot(
  srt = seurat_object, 
  group.by = "orig.ident",
  reduction = "umap",
  label = TRUE, 
  label_repel = TRUE,
  show_stat = TRUE,
  theme_use = "theme_blank"
)

```

:::{.callout-tip}
The @fig-umap-sample sample-based UMAP can be used to evaluate cluster composition. Clusters composed of a single sample can be potentially associated with malignant cells.
:::



### Batch-based UMAP

```{r umap_batch_plot}
#| label: fig-umap-batch
#| fig-cap: 'UMAP'
#| fig-width: 15
#| fig-height: 10
#| fig-align: center

# SCP::CellDimPlot(
#   srt = seurat_object, 
#   group.by = "batch",
#   reduction = "umap",
#   label = TRUE, 
#   label_repel = TRUE,
#   show_stat = TRUE,
#   theme_use = "theme_blank"
# )

# Check if "batch" is in metadata before running the batch-based UMAP
if ("batch" %in% colnames(seurat_object@meta.data)) {
  SCP::CellDimPlot(
    srt = seurat_object, 
    group.by = "batch",
    reduction = "umap",
    label = TRUE, 
    label_repel = TRUE,
    show_stat = TRUE,
    theme_use = "theme_blank"
  )
} else {
  message("Skipping batch-based UMAP: 'batch' column not found in metadata.")
}

```

:::{.callout-tip}
The @fig-umap-batch displays the batch effect over the cell clusters.
:::

### Group-based UMAP

```{r}
#| label: fig-umap-multi
#| eval: false
#| column: screen-inset-shaded
#| fig-align: center
#| fig-cap: UMAP
#| fig-height: 8
#| fig-width: 12
#| include: false
#| layout-nrow: 1

# SCP::CellDimPlot(
#   srt = seurat_object, 
#   group.by = input_group_plot,
#   reduction = "umap",
#   label = TRUE, 
#   label_repel = TRUE,
#   show_stat = TRUE,
#   theme_use = "theme_blank"
# )

# Split input_group_plot (it contains "patient_id;timepoint")
group_vars <- unlist(strsplit(input_group_plot, ";"))

# Check if all group_vars exist in metadata before running the group-based UMAP
if (all(group_vars %in% colnames(seurat_object@meta.data))) {
  SCP::CellDimPlot(
    srt = seurat_object, 
    group.by = input_group_plot,
    reduction = "umap",
    label = TRUE, 
    label_repel = TRUE,
    show_stat = TRUE,
    theme_use = "theme_blank"
  )
} else {
  message("Skipping group-based UMAP: Some columns in 'patient_id;timepoint' not found in metadata.")
}


```

:::{.callout-note}
The group-based UMAP above is `eval: false`, so it draws nothing here. It facets on
the other study/design features (`patient_id;timepoint`). Set `eval: true` when the
metadata carries those columns — the chunk already falls back to a message when it
does not.
:::

## Saving Seurat object

```{r materialise_layers}
#| code-fold: true

# CLUSTER is the last stage that runs in a BPCells-aware image. Everything
# downstream -- annotation, CNV, stratify, metaprograms, batch correction,
# trajectory, cell communication, TME -- runs in its own container, and none of
# them ship BPCells. A BPCells-backed object handed to any of them dies at load
# with "unable to load required package 'BPCells'" from deep inside
# as.SingleCellExperiment(), which names the missing package but not the reason.
#
# So the on-disk store stops here: the layers are realised into ordinary sparse
# matrices before the object is published. BPCells still does the job it was
# brought in for -- the merge and the doublet pass over every raw cell, where
# JoinLayers used to hit dgCMatrix's ~2.1 billion non-zero ceiling. By this
# point the object has been through QC filtering and doublet removal.
#
# It does not always fit, which is why this is measured rather than assumed.
.nnz_ceiling <- 2^31 - 1   # dgCMatrix indexes non-zeros with a 32-bit signed int

.layer_names <- SeuratObject::Layers(seurat_object, assay = 'RNA')
.bp_layers <- .layer_names[vapply(.layer_names, function(ln) {
  inherits(SeuratObject::LayerData(seurat_object, assay = 'RNA', layer = ln),
           'IterableMatrix')
}, logical(1))]

.materialised <- character(0)
.kept_ondisk  <- character(0)

if (length(.bp_layers)) {

  # Counted once, on the counts layer, and applied to all of them: normalisation
  # transforms the stored values but not which entries are stored, so every layer
  # shares one sparsity pattern. Running matrix_stats() on the `data` layer would
  # instead force the whole lazy log-normalisation to compute just to be counted.
  .counts_layer <- if ('counts' %in% .bp_layers) 'counts' else .bp_layers[1]
  .nnz <- tryCatch(
    sum(as.numeric(BPCells::matrix_stats(
      SeuratObject::LayerData(seurat_object, assay = 'RNA', layer = .counts_layer),
      row_stats = 'nonzero')$row_stats['nonzero', ])),
    error = function(e) NA_real_)

  for (.ln in .bp_layers) {
    .lay <- SeuratObject::LayerData(seurat_object, assay = 'RNA', layer = .ln)
    if (!is.na(.nnz) && .nnz > .nnz_ceiling) {
      .kept_ondisk <- c(.kept_ondisk, .ln)
      next
    }
    # scale.data must be a DENSE base matrix: Seurat's v5->v4 assay coercion
    # (which Azimuth performs before mapping) validates that slot as
    # "matrix", and a dgCMatrix there aborts with `invalid class "Assay"
    # object: invalid object for slot "scale.data"`. It is scaled over the
    # variable features only, so densifying it is bounded.
    SeuratObject::LayerData(seurat_object, assay = 'RNA', layer = .ln) <-
      if (startsWith(.ln, 'scale.data')) as.matrix(.lay) else as(.lay, 'dgCMatrix')
    .materialised <- c(.materialised, .ln)
  }

  message(sprintf("BPCells layers: %s | non-zeros: %s | materialised: %s",
                  paste(.bp_layers, collapse = ','),
                  format(.nnz, big.mark = ',', scientific = FALSE),
                  paste(c(.materialised, '(none)')[1], collapse = ',')))
}

```

```{r materialise-finding, results = 'asis'}
if (length(.bp_layers)) {
  if (length(.kept_ondisk)) {
    scratch_finding(sprintf(paste(
      "Layer(s) **%s** exceed dgCMatrix's 2.1 billion non-zero ceiling and were",
      "left as on-disk BPCells matrices. **Every stage after clustering runs in a",
      "container without BPCells and will fail to load this object.** Subset the",
      "cohort, or add BPCells to the downstream images, before running annotation."),
      paste(.kept_ondisk, collapse = ', ')),
      type = "caution", title = "Object left on disk")
  } else {
    scratch_finding(sprintf(paste(
      "Layer(s) **%s** were realised from the on-disk BPCells store into in-memory",
      "sparse matrices, so the published object loads in the downstream containers",
      "(none of which ship BPCells). The store itself is unchanged."),
      paste(.materialised, collapse = ', ')),
      type = "note", title = "Counts materialised for downstream stages")
  }
}
```

```{r object_dump}

if(auto_save) {
  
  saveRDS(seurat_object, file = paste0('./data/', project_name, '_cluster_object.RDS'))
  
}

```

------------------------------------------------------------------------

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

params

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```
