---
title: "SCRATCH-E2E — Malignant / non-malignant stratification"
execute:
  freeze: false
  cache: false
  warning: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    embed-resources: true
    code-fold: show
    fig-format: png
    fig-dpi: 300
    fig-align: center
params:
  seurat_object: null
  project_name: "project"
  cnv_evidence_dir: "cnv_evidence"
  work_directory: "."

  # consensus | cnv | annotation
  method: "consensus"

  # How many CNV callers must agree before their verdict is taken.
  min_callers: 1

  # Annotation labels treated as the tumour lineage (tie-break / fallback).
  tumor_labels: "Epithelial"
  celltype_col: "azimuth_labels"
  sample_col: "sample_id"

  # Lineages that are, a priori, never malignant. Used to build the null
  # distribution for inferCNV's continuous score.
  reference_labels: "T_Cells;NK_Cells;Myeloid;B_Plasma_Cells;Endothelial_Cells;Fibroblast"

  # inferCNV score threshold, in SDs above the reference-lineage mean.
  infercnv_sd_threshold: 3

  # Cell type defines the compartment; CNV adjudicates only within the
  # compartment that could plausibly be malignant.
  gate_by_lineage: true

  # exclude | malignant | non_malignant
  ambiguous: "exclude"

  min_cells_per_compartment: 50
  seed: 1234
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE}
suppressPackageStartupMessages({
  library(Seurat)
  library(data.table)
  library(ggplot2)
  library(scales)
})

set.seed(params$seed)
if (!is.null(params$work_directory) && dir.exists(params$work_directory)) {
  setwd(params$work_directory)
}
dir.create("data",    showWarnings = FALSE, recursive = TRUE)
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

# Palette: two compartments plus an ambiguous class. Validated categorical
# slots; every figure also carries text labels, so colour is never the only
# channel distinguishing them.
COL <- c(malignant = "#eb6834", non_malignant = "#2a78d6", ambiguous = "#8a9095")
INK <- "#0f1417"; INK2 <- "#4e5a60"; MUTED <- "#898781"
GRID <- "#e1e0d9"; SURFACE <- "#fcfcfb"

theme_scratch <- function(base = 12) {
  theme_minimal(base_size = base) +
    theme(plot.background  = element_rect(fill = SURFACE, colour = NA),
          panel.background = element_rect(fill = SURFACE, colour = NA),
          panel.grid.major = element_line(colour = GRID, linewidth = .3),
          panel.grid.minor = element_blank(),
          axis.text  = element_text(colour = MUTED),
          axis.title = element_text(colour = INK2),
          plot.title = element_text(colour = INK, face = "bold"),
          plot.subtitle = element_text(colour = INK2, size = base * .9),
          legend.text = element_text(colour = INK2),
          legend.title = element_text(colour = INK2))
}

split_list <- function(x) {
  if (is.null(x)) return(character(0))
  trimws(unlist(strsplit(as.character(x), "[;,|]")))
}
```

## What this step decides

Every cell is assigned to one of three classes:

| class | meaning |
|---|---|
| `malignant` | aneuploid by CNV evidence (or tumour-lineage when no CNV evidence exists) |
| `non_malignant` | diploid by CNV evidence, or a reference immune/stromal lineage |
| `ambiguous` | the available evidence conflicts |

**CNV evidence decides; the annotation label is the fallback.** Aneuploidy is a
direct measurement of malignancy, whereas a label such as `Epithelial` is a
lineage proxy that calls normal epithelium a tumour.

Under `method = "consensus"` (the default) a *majority* CNV verdict wins; when
the callers split evenly, or when no caller scored the cell, the annotation
label decides. `ambiguous` therefore means "no CNV verdict and no usable label",
not "the callers disagreed". Use `method = "cnv"` if you would rather exclude
conflicting cells than let the label resolve them.

`reference_labels` must match the vocabulary your annotation actually emits —
they define the null distribution for inferCNV's continuous score, and a name
that matches nothing silently shrinks it.

```{r load}
stopifnot(!is.null(params$seurat_object))
obj <- readRDS(params$seurat_object)
cat("Input object:", ncol(obj), "cells x", nrow(obj), "features\n")

meta <- as.data.table(obj@meta.data, keep.rownames = "cell_id")
meta[, cell_id := as.character(cell_id)]

# Resolve the annotation column, tolerating naming drift between stages.
ct_candidates <- c(params$celltype_col, "azimuth_labels", "celltype", "cell_type",
                   "annotation", "sctype", "sctype_agg", "sctype_labels", "cell_label")
# NOTE: "sctype" is what THIS pipeline's own annotation stage writes. Its absence
# from this list made a real run complete with no labels at all — gate dead,
# no tie-break, 22.6% of cells dropped to ambiguous, exit code 0.
col_ct <- ct_candidates[ct_candidates %in% names(meta)][1]
if (is.na(col_ct)) {
  msg <- paste0("No annotation column found. Tried: ", paste(ct_candidates, collapse=", "),
                "\n  Object has: ", paste(head(names(meta), 40), collapse=", "),
                "\n  Set --stratify_celltype_col to the right column.")
  # Under `consensus` the label drives BOTH the lineage gate and the tie-break,
  # so continuing without it silently changes the algorithm. Fail loudly.
  if (params$method == "consensus" || isTRUE(as.logical(params$gate_by_lineage))) {
    stop(msg)
  }
  warning(msg, "\n  method=", params$method, " does not require it; continuing.")
  meta[, .__ct := NA_character_]
  col_ct <- ".__ct"
}
cat("Annotation column:", col_ct, "\n")

# Sample column drives the per-sample inferCNV thresholds.
smp_candidates <- c(params$sample_col, "sample_id", "orig.ident", "sample", "patient_id")
col_smp <- smp_candidates[smp_candidates %in% names(meta)][1]
cat("Sample column:", if (is.na(col_smp)) "none (one cohort-wide threshold)" else col_smp, "\n")

tumor_labels <- split_list(params$tumor_labels)
ref_labels   <- split_list(params$reference_labels)
cat("Tumour-lineage labels:", paste(tumor_labels, collapse = ", "), "\n")
```

## Collecting CNV evidence

```{r evidence}
ev_dir <- params$cnv_evidence_dir
ev_files <- if (dir.exists(ev_dir)) {
  list.files(ev_dir, recursive = TRUE, full.names = TRUE)
} else character(0)

cat("Staged CNV artefacts:", length(ev_files), "\n")
if (length(ev_files)) cat(paste(" -", basename(ev_files)), sep = "\n")

evidence <- data.table(cell_id = meta$cell_id)
callers  <- character(0)

# ---- inferCNV: a continuous genome-wide CNV load ---------------------------
# There is no hard call, so the cutoff is set from the reference lineages: the
# score distribution of cells that cannot be malignant is the null, and cells
# more than `infercnv_sd_threshold` SDs above its mean are called aneuploid.
f <- ev_files[grepl("infercnv_meta_object\\.RDS$", ev_files)]
if (length(f)) {
  ic <- tryCatch(readRDS(f[1]), error = function(e) NULL)
  ic_meta <- if (inherits(ic, "Seurat")) as.data.table(ic@meta.data, keep.rownames = "cell_id")
             else if (is.data.frame(ic))  as.data.table(ic, keep.rownames = "cell_id")
             else NULL

  score_col <- if (!is.null(ic_meta)) {
    intersect(c("genomwide_cnv_proportion", "genomewide_cnv_proportion",
                "cnv_score", "cnv_load"), names(ic_meta))[1]
  } else NA_character_

  if (!is.null(ic_meta) && !is.na(score_col)) {
    ic_meta[, cell_id := as.character(cell_id)]
    sc <- ic_meta[, .(cell_id, infercnv_score = as.numeric(get(score_col)))]
    evidence <- merge(evidence, sc, by = "cell_id", all.x = TRUE, sort = FALSE)

    # --- Per-sample thresholds -------------------------------------------
    # CNV-load distributions shift between samples: a CSF aspirate with a
    # handful of tumour cells and a resection with thousands do not share a
    # scale. One global cutoff over-calls in some samples and under-calls in
    # others, so the null is rebuilt per sample from that sample's own
    # reference lineages. Samples with too few reference cells fall back to
    # the cohort-wide null, and only then to a quantile.
    ref_ids <- meta[get(col_ct) %in% ref_labels, cell_id]
    ev <- merge(evidence[, .(cell_id, infercnv_score)],
                meta[, .(cell_id, smp = if (!is.na(col_smp)) as.character(get(col_smp)) else "ALL")],
                by = "cell_id", all.x = TRUE, sort = FALSE)
    ev[is.na(smp), smp := "ALL"]

    global_null <- ev[cell_id %in% ref_ids & is.finite(infercnv_score), infercnv_score]
    k <- as.numeric(params$infercnv_sd_threshold)

    global_cut <- if (length(global_null) >= 30) {
      mean(global_null) + k * stats::sd(global_null)
    } else {
      warning(sprintf("Only %d reference cells cohort-wide; falling back to the 75th percentile. Check that --stratify_reference_labels matches your annotation vocabulary.",
                      length(global_null)))
      stats::quantile(ev$infercnv_score, .75, na.rm = TRUE)
    }

    cuts <- ev[, {
      nulls <- infercnv_score[cell_id %in% ref_ids & is.finite(infercnv_score)]
      if (length(nulls) >= 30) .(cutoff = mean(nulls) + k * stats::sd(nulls),
                                 n_ref = length(nulls), basis = "per_sample")
      else                     .(cutoff = global_cut, n_ref = length(nulls), basis = "cohort")
    }, by = smp]

    cat(sprintf("inferCNV cutoffs (mean + %g SD of reference lineages):\n", k))
    print(cuts[order(smp)])
    fwrite(cuts, "data/infercnv_thresholds.csv")

    ev <- merge(ev, cuts[, .(smp, cutoff)], by = "smp", all.x = TRUE, sort = FALSE)
    ev[, call := fifelse(is.na(infercnv_score), NA_character_,
                  fifelse(infercnv_score > cutoff, "malignant", "non_malignant"))]

    evidence <- merge(evidence, ev[, .(cell_id, infercnv_call = call)],
                      by = "cell_id", all.x = TRUE, sort = FALSE)
    callers <- c(callers, "infercnv")
  } else {
    cat("inferCNV object staged but no recognised score column.\n")
  }
} else cat("No inferCNV evidence.\n")

# ---- SCEVAN: a per-cell tumour/normal class -------------------------------
f <- ev_files[grepl("scevan_meta_object\\.RDS$", ev_files)]
if (length(f)) {
  sv <- tryCatch(readRDS(f[1]), error = function(e) NULL)
  sv_meta <- if (inherits(sv, "Seurat")) as.data.table(sv@meta.data, keep.rownames = "cell_id")
             else if (is.data.frame(sv))  as.data.table(sv, keep.rownames = "cell_id")
             else NULL
  cls <- if (!is.null(sv_meta)) intersect(c("class", "scevan_class", "pred"), names(sv_meta))[1] else NA_character_

  if (!is.null(sv_meta) && !is.na(cls)) {
    sv_meta[, cell_id := as.character(cell_id)]
    sc <- sv_meta[, .(cell_id, raw = tolower(as.character(get(cls))))]
    sc[, scevan_call := fifelse(grepl("tumou?r|malig|aneupl", raw), "malignant",
                         fifelse(grepl("normal|diplo", raw), "non_malignant", NA_character_))]
    evidence <- merge(evidence, sc[, .(cell_id, scevan_call)], by = "cell_id", all.x = TRUE, sort = FALSE)
    callers <- c(callers, "scevan")
  }
} else cat("No SCEVAN evidence.\n")

# ---- CopyKAT: aneuploid/diploid -------------------------------------------
f <- ev_files[grepl("copykat", ev_files, ignore.case = TRUE) & grepl("\\.(csv|tsv|txt|RDS)$", ev_files)]
if (length(f)) {
  ck <- tryCatch(
    if (grepl("\\.RDS$", f[1])) as.data.table(readRDS(f[1]), keep.rownames = "cell_id") else fread(f[1]),
    error = function(e) NULL)
  if (!is.null(ck) && nrow(ck)) {
    idc <- intersect(c("cell.names", "cell_id", "cell", "barcode", "V1"), names(ck))[1]
    prd <- intersect(c("copykat.pred", "copykat_pred", "prediction", "pred"), names(ck))[1]
    if (!is.na(idc) && !is.na(prd)) {
      setnames(ck, c(idc, prd), c("cell_id", "raw"))
      ck[, cell_id := as.character(cell_id)]
      ck[, copykat_call := fifelse(grepl("aneuploid", tolower(raw)), "malignant",
                            fifelse(grepl("diploid", tolower(raw)), "non_malignant", NA_character_))]
      evidence <- merge(evidence, ck[, .(cell_id, copykat_call)], by = "cell_id", all.x = TRUE, sort = FALSE)
      callers <- c(callers, "copykat")
    }
  }
} else cat("No CopyKAT evidence.\n")

# ---- Numbat: allele-aware, already a hard per-cell call --------------------
# Emitted keyed by the pipeline's own cell_id (numbat_run.R does the join via
# the explicit barcode map), so unlike the other callers no id translation is
# needed here.
f <- ev_files[grepl("_numbat_calls\\.tsv$", ev_files)]
if (length(f)) {
  nb <- rbindlist(lapply(f, function(x) tryCatch(fread(x), error = function(e) NULL)),
                  use.names = TRUE, fill = TRUE)
  if (!is.null(nb) && nrow(nb) && all(c("cell_id", "numbat_call") %in% names(nb))) {
    nb[, cell_id := as.character(cell_id)]
    nb <- unique(nb[, .(cell_id, numbat_call)], by = "cell_id")
    evidence <- merge(evidence, nb, by = "cell_id", all.x = TRUE, sort = FALSE)
    callers <- c(callers, "numbat")
    cat(sprintf("Numbat: %d cells called across %d sample file(s)\n", nrow(nb), length(f)))
  }
} else cat("No Numbat evidence.\n")

cat("\nCNV callers contributing:", if (length(callers)) paste(callers, collapse = ", ") else "none", "\n")
```

```{r annotation-evidence}
# Annotation-derived call: tumour lineage vs known-reference lineage.
ann <- meta[, .(cell_id, lab = as.character(get(col_ct)))]
ann[, annotation_call := fifelse(lab %in% tumor_labels, "malignant",
                          fifelse(lab %in% ref_labels, "non_malignant", NA_character_))]
evidence <- merge(evidence, ann[, .(cell_id, annotation_call)], by = "cell_id", all.x = TRUE, sort = FALSE)
```

## Combining the evidence

```{r combine}
call_cols <- intersect(c("infercnv_call", "scevan_call", "copykat_call", "numbat_call"),
                       names(evidence))

n_mal <- if (length(call_cols)) rowSums(evidence[, ..call_cols] == "malignant", na.rm = TRUE) else rep(0L, nrow(evidence))
n_non <- if (length(call_cols)) rowSums(evidence[, ..call_cols] == "non_malignant", na.rm = TRUE) else rep(0L, nrow(evidence))

evidence[, n_malignant     := n_mal]
evidence[, n_non_malignant := n_non]
evidence[, n_cnv_calls     := n_mal + n_non]

method      <- params$method
min_callers <- as.integer(params$min_callers)

if (method == "annotation" || !length(call_cols)) {
  if (method != "annotation") {
    warning("No CNV evidence available; falling back to annotation-only stratification.")
  }
  evidence[, malignant_status := fifelse(is.na(annotation_call), "ambiguous", annotation_call)]
  evidence[, evidence_basis := "annotation"]

} else if (method == "cnv") {
  evidence[, malignant_status := fifelse(
      n_cnv_calls == 0L, "ambiguous",
      fifelse(n_malignant >= min_callers & n_malignant > n_non_malignant, "malignant",
      fifelse(n_non_malignant >= min_callers & n_non_malignant > n_malignant, "non_malignant",
              "ambiguous")))]
  evidence[, evidence_basis := fifelse(n_cnv_calls == 0L, "none", "cnv")]

} else {  # consensus (default)
  # CNV decides when it reaches a majority verdict. Otherwise — including when
  # two callers split 1-1 — the annotation label decides. `ambiguous` is
  # therefore reserved for cells with NO CNV verdict AND no usable label; a
  # caller disagreement does not by itself produce one. Use method = "cnv" if
  # you want conflicts to be excluded rather than resolved by the label.
  evidence[, malignant_status := fifelse(
      n_malignant >= min_callers & n_malignant > n_non_malignant, "malignant",
      fifelse(n_non_malignant >= min_callers & n_non_malignant > n_malignant, "non_malignant",
      fifelse(!is.na(annotation_call), annotation_call, "ambiguous")))]
  evidence[, evidence_basis := fifelse(
      n_malignant >= min_callers & n_malignant > n_non_malignant, "cnv",
      fifelse(n_non_malignant >= min_callers & n_non_malignant > n_malignant, "cnv",
      fifelse(!is.na(annotation_call), "annotation", "none")))]
}

cat("Method:", method, "| min_callers:", min_callers, "\n\n")
print(evidence[, .N, by = .(malignant_status, evidence_basis)][order(malignant_status, -N)])
```

## Lineage gate

```{r lineage-gate}
# Cell type defines the biological compartment; CNV adjudicates malignancy only
# WITHIN the compartment that could plausibly be malignant.
#
# Without this gate the procedure is internally inconsistent: the reference
# lineages define the null distribution for inferCNV's score, yet nothing stops
# a cell of those same lineages from landing above the cutoff. With a one-sided
# mean + k*SD threshold a fixed fraction of them will by construction — on a
# large immune infiltrate that is tens of cells entering the tumour compartment
# and flowing into the metaprogram stage.
#
# Cells NOT in a reference lineage — including unlabelled and "Unknown" cells —
# are left to the CNV evidence. That matters for tumours whose malignant cells
# have no entry in the marker database (a glioma's astrocyte-like cells are not
# "Epithelial", and this database has no glial lineage at all): gating on a
# positive tumour-lineage list would discard them, gating on the negative
# reference list keeps them.
gate_on <- isTRUE(as.logical(params$gate_by_lineage))

evidence <- merge(evidence, meta[, .(cell_id, .lab = as.character(get(col_ct)))],
                  by = "cell_id", all.x = TRUE, sort = FALSE)
evidence[, in_reference_lineage := !is.na(.lab) & .lab %in% ref_labels]

if (gate_on) {
  # Empirical false-positive estimate: cells the CNV evidence called malignant
  # that a known non-malignant lineage overrides. This is the single most useful
  # number for judging whether the CNV calls can be trusted.
  fp <- evidence[in_reference_lineage == TRUE & malignant_status == "malignant"]
  n_ref_total <- evidence[in_reference_lineage == TRUE, .N]

  if (nrow(fp)) {
    cat(sprintf("\nLineage gate overrode %d of %d reference-lineage cells (%.2f%%) that CNV had called malignant.\n",
                nrow(fp), n_ref_total, 100 * nrow(fp) / max(n_ref_total, 1)))
    cat("Treat that percentage as an empirical false-positive rate for the CNV calls.\n")
    print(fp[, .N, by = .lab][order(-N)])
    fwrite(fp[, .N, by = .lab][order(-N)], "data/cnv_false_positive_by_lineage.csv")
  } else {
    cat("\nLineage gate overrode no cells: no reference-lineage cell was called malignant.\n")
  }

  evidence[in_reference_lineage == TRUE,
           `:=`(malignant_status = "non_malignant", evidence_basis = "lineage_gate")]
} else {
  cat("\nLineage gate disabled (--stratify_gate_by_lineage false): CNV calls stand on every lineage.\n")
}

cat("\nAfter gating:\n")
print(evidence[, .N, by = .(malignant_status, evidence_basis)][order(malignant_status, -N)])
evidence[, .lab := NULL]
```

```{r concordance}
# Where the callers disagree, and by how much. This is the number to look at
# before trusting the split.
if (length(call_cols) >= 2) {
  conc <- evidence[n_cnv_calls >= 2]
  if (nrow(conc)) {
    agree <- conc[n_malignant == 0L | n_non_malignant == 0L, .N]
    cat(sprintf("\nCaller concordance: %d of %d cells with >=2 calls agree (%.1f%%)\n",
                agree, nrow(conc), 100 * agree / nrow(conc)))
  }
  pair <- evidence[, .N, by = call_cols]
  fwrite(pair, "data/cnv_caller_agreement.csv")
  knitr::kable(pair)
} else {
  cat("\nFewer than two CNV callers ran; no concordance to report.\n")
}
```

## Assignment

```{r assign}
setkey(evidence, cell_id)
assign_dt <- evidence[meta$cell_id, .(cell_id, malignant_status, evidence_basis,
                                      n_malignant, n_non_malignant, annotation_call)]
assign_dt[is.na(malignant_status), malignant_status := "ambiguous"]

status <- setNames(assign_dt$malignant_status, assign_dt$cell_id)
obj$malignant_status <- unname(status[colnames(obj)])
obj$malignancy_evidence <- unname(setNames(assign_dt$evidence_basis, assign_dt$cell_id)[colnames(obj)])

tab <- assign_dt[, .N, by = malignant_status][order(-N)]
tab[, pct := round(100 * N / sum(N), 1)]
knitr::kable(tab)

fwrite(assign_dt, file.path("data", paste0(params$project_name, "_malignancy_assignment.csv")))
```

```{r composition-plot, fig.width = 9, fig.height = 5}
if (!is.na(col_ct) && col_ct != ".__ct") {
  comp <- merge(assign_dt[, .(cell_id, malignant_status)],
                meta[, .(cell_id, lab = as.character(get(col_ct)))],
                by = "cell_id")
  comp <- comp[!is.na(lab)][, .N, by = .(lab, malignant_status)][, frac := N / sum(N), by = lab]

  p <- ggplot(comp, aes(x = reorder(lab, -N), y = frac, fill = malignant_status)) +
    geom_col(width = .72, colour = SURFACE, linewidth = 1) +
    geom_text(data = comp[frac > .12],
              aes(label = percent(frac, accuracy = 1)),
              position = position_stack(vjust = .5), size = 3, colour = "#ffffff") +
    scale_fill_manual(values = COL, name = "compartment") +
    scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, .02))) +
    labs(title = "Malignancy assignment within each annotated lineage",
         subtitle = "A tumour-lineage label that is largely diploid indicates normal epithelium",
         x = NULL, y = "share of cells") +
    theme_scratch() + theme(axis.text.x = element_text(angle = 35, hjust = 1))
  print(p)
  ggsave("figures/malignancy_by_lineage.png", p, width = 9, height = 5.5, dpi = 400)
}
```

```{r score-plot, fig.width = 9, fig.height = 4}
if ("infercnv_score" %in% names(evidence)) {
  d <- merge(evidence[, .(cell_id, infercnv_score)],
             assign_dt[, .(cell_id, malignant_status)], by = "cell_id")
  d <- d[is.finite(infercnv_score)]
  if (nrow(d) > 20) {
    p <- ggplot(d, aes(x = infercnv_score, fill = malignant_status)) +
      geom_density(alpha = .55, colour = NA) +
      scale_fill_manual(values = COL, name = "compartment") +
      labs(title = "inferCNV genome-wide CNV load by assigned compartment",
           subtitle = "Separation here is the visual check on the cutoff",
           x = "CNV load", y = "density") +
      theme_scratch()
    print(p)
    ggsave("figures/infercnv_score_distribution.png", p, width = 9, height = 4, dpi = 400)
  }
}
```

## Writing the two compartments

```{r split}
ambiguous_rule <- params$ambiguous

tumor_cells <- colnames(obj)[obj$malignant_status == "malignant"]
normal_cells <- colnames(obj)[obj$malignant_status == "non_malignant"]
amb_cells   <- colnames(obj)[obj$malignant_status == "ambiguous"]

if (ambiguous_rule == "malignant") {
  tumor_cells <- c(tumor_cells, amb_cells)
} else if (ambiguous_rule == "non_malignant") {
  normal_cells <- c(normal_cells, amb_cells)
}
# 'exclude' (default): ambiguous cells enter neither compartment, but they are
# still recorded in the assignment table and in `malignant_status` on the object.

cat(sprintf("malignant:     %d cells\nnon-malignant: %d cells\nambiguous:     %d cells (%s)\n",
            length(tumor_cells), length(normal_cells), length(amb_cells), ambiguous_rule))

min_cells <- as.integer(params$min_cells_per_compartment)

if (length(tumor_cells) >= min_cells) {
  saveRDS(subset(obj, cells = tumor_cells),
          file.path("data", paste0(params$project_name, "_malignant_object.RDS")))
} else {
  warning(sprintf("Only %d malignant cells (< min_cells_per_compartment=%d); no tumour object written. The metaprogram stage will be skipped.",
                  length(tumor_cells), min_cells))
}

if (length(normal_cells) >= min_cells) {
  saveRDS(subset(obj, cells = normal_cells),
          file.path("data", paste0(params$project_name, "_nonmalignant_object.RDS")))
} else {
  warning(sprintf("Only %d non-malignant cells (< min_cells_per_compartment=%d); no non-malignant object written. Batch correction and its downstream stages will be skipped.",
                  length(normal_cells), min_cells))
}

# The full object, carrying the assignment, for the integration layer.
saveRDS(obj, file.path("data", paste0(params$project_name, "_stratified_object.RDS")))

fwrite(data.table(compartment = c("malignant", "non_malignant", "ambiguous"),
                  cells = c(length(tumor_cells), length(normal_cells), length(amb_cells)),
                  ambiguous_rule = ambiguous_rule),
       "data/compartment_sizes.csv")
```

## Parameters

```{r params_log, echo = FALSE}
print(params)
```

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
