---
title: "Module - Doublet Detection"
description: | 
  Description
execute:
  freeze: auto
  cache: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: 'Project'
  seurat_object: './data/Project_merged_object.RDS'
  n_threads: 8
  n_memory: 16
  work_directory: !expr here::here()
  auto_save: TRUE
---

# Project Name: `r params$project_name`

```{r setup, include = FALSE, message = FALSE, warning = FALSE, echo = FALSE}
#| include: false

# Project parameters 
project_name <- params$project_name
seurat_object <- params$seurat_object

# Inputs and thresholds

# Optional parameters

# Dataflow/Computational parameters
n_threads <- params$n_threads
n_memory <- params$n_memory

# Output parameters
work_directory <- params$work_directory
auto_save <- params$auto_save

# Rmarkdown options

knitr::opts_knit$set(
  root.dir = work_directory
  )

```

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```


## Loading library

```{r library}
#| output: false

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(BPCells)
library(DT)
library(SingleCellExperiment)
library(scDblFinder)
library(SCP)

# Expanding memory
options(future.globals.maxSize = n_memory*1024^3)

```

## Folder structure

```{r project_directory}

if(!dir.exists(work_directory)) {
  dir.create(work_directory, recursive = T)
}

# showWarnings = FALSE: data/ already exists (the BPCells store is staged there).
for(sub_directory in c('data', 'figures', 'figures/doublet')) {
  dir.create(paste0(work_directory, '/', sub_directory), showWarnings = FALSE)
}

```

## Loading Seurat object

```{r loading_rds}

# The merged object is BPCells-backed. Its counts live in the on-disk store
# staged at data/bpcells_counts. Rebuild the object from that store plus the
# saved per-cell metadata so we never depend on the (possibly stale) matrix
# path baked into the RDS by the upstream merge task.
loaded_object <- readRDS(file = seurat_object)

disk_counts <- BPCells::open_matrix_dir(dir = file.path('data', 'bpcells_counts'))

keep_meta <- loaded_object@meta.data
keep_meta <- keep_meta[, setdiff(colnames(keep_meta),
                                 c('nCount_RNA', 'nFeature_RNA')), drop = FALSE]

seurat_object <- CreateSeuratObject(
  counts    = disk_counts,
  project   = project_name,
  meta.data = keep_meta
)

rm(loaded_object); gc()

# Setting random seed
random_seed <- 2203
set.seed(random_seed)

```

## Per-sample doublet detection

Doublets form **within a single 10x capture**, so scDblFinder is run
independently on each sample (`orig.ident`) — the same partitioning
`scDblFinder(..., samples=)` performs internally, but done explicitly so the
full matrix is never realized at once. The merged object is BPCells-backed, so
each sample's columns are subset lazily and only that one small sample is
materialized in memory at a time. This never approaches the 2.1B non-zero
`dgCMatrix` ceiling a full-cohort realization would hit, so it scales to
arbitrarily large cohorts.

Results are reproducible: a fixed RNG seed is used, samples are processed in
sorted order, and `bplapply` returns results in input order regardless of
execution order.

```{r scdblfinder_per_sample}
#| warning: false

# BPCells-backed counts for the whole merged object (lazy; nothing realized yet).
counts_mat <- SeuratObject::LayerData(seurat_object, assay = 'RNA', layer = 'counts')

# Map every cell to its sample and iterate samples in a fixed (sorted) order.
cell_sample <- as.character(seurat_object$orig.ident)
names(cell_sample) <- colnames(seurat_object)
sample_ids  <- sort(unique(cell_sample))

detect_one <- function(s) {
  cells <- names(cell_sample)[cell_sample == s]

  # Realize ONLY this sample's counts (a small dgCMatrix, never near the ceiling
  # that realizing the full cohort at once would hit).
  sub <- methods::as(counts_mat[, cells, drop = FALSE], 'dgCMatrix')

  tryCatch({
    tbl <- as.data.frame(scDblFinder(sub, returnType = 'table'))
    # returnType='table' includes the artificial doublets used internally; keep
    # only the real cells so rownames line up with the object's barcodes.
    tbl <- tbl[tbl$type == 'real', , drop = FALSE]
    data.frame(
      row.names        = rownames(tbl),
      class            = as.character(tbl$class),
      scdbl_score      = tbl$score,
      stringsAsFactors = FALSE
    )
  }, error = function(e) {
    # Degenerate/too-few-cells sample: keep every cell (conservative) instead of
    # dropping the sample, and record why.
    message(sprintf("scDblFinder failed for '%s' (%s); keeping all its cells as singlet",
                    s, conditionMessage(e)))
    data.frame(
      row.names        = cells,
      class            = 'singlet',
      scdbl_score      = NA_real_,
      stringsAsFactors = FALSE
    )
  })
}

# Parallel across samples but reproducible: BiocParallel seeds each worker's RNG
# stream from RNGseed. (Serial + set.seed per sample would be equally
# reproducible; this is just faster for large cohorts.)
bp <- BiocParallel::MulticoreParam(workers = max(1L, as.integer(n_threads)),
                                   RNGseed = random_seed)
calls_list <- BiocParallel::bplapply(sample_ids, detect_one, BPPARAM = bp)

dbl_sample <- do.call(rbind, calls_list)
# Align to the object's cell order (barcodes are already unique across samples).
dbl_sample <- dbl_sample[colnames(seurat_object), , drop = FALSE]

# Per-sample singlet/doublet tally.
print(table(sample = cell_sample, class = dbl_sample$class))

```

## Adding metadata

```{r double_combine_metadata}
#| echo: true

seurat_object <- AddMetaData(
  seurat_object,
  metadata = dbl_sample
)

```

```{r meta_inspection}
#| column: screen-inset-shaded

# Cap the interactive table: embedding every cell would bloat the HTML for
# large cohorts. Show a reproducible random sample when the object is big.
meta_preview <- seurat_object@meta.data
if (nrow(meta_preview) > 2000) {
  set.seed(random_seed)
  meta_preview <- meta_preview[sample(nrow(meta_preview), 2000), , drop = FALSE]
}

# Registered as well as rendered, so the agent can compute over these cells.
# See the note in notebook_quality_table_report.qmd.
scratch_table(meta_preview, caption = "Per-cell doublet metadata (sampled)", page = 12)

```

## Doublet composition

```{r}
#| label: fig-doublet-composition
#| fig-cap: 'Barplot'
#| fig-subcap: 
#| - 'Cell composition across all cells'
#| - 'Cell composition by cell clusters'
#| column: screen-inset-shaded
#| layout-nrow: 1

meta_cols <- colnames(seurat_object@meta.data)

# Group by patient_id when the metadata carries it; otherwise fall back to the
# sample id (orig.ident, always present) so the composition still renders.
group_col <- if ("patient_id" %in% meta_cols) "patient_id" else "orig.ident"

if ("class" %in% meta_cols) {
  print(SCP::CellStatPlot(
      seurat_object,
      stat.by = "class",
      group.by = group_col,
      legend.position = "bottom",
      legend.direction = "horizontal",
      label = TRUE
  ))
}

if ("cluster_class" %in% meta_cols) {
  print(SCP::CellStatPlot(
      seurat_object,
      stat.by = "cluster_class",
      group.by = group_col,
      legend.position = "bottom",
      legend.direction = "horizontal",
      label = TRUE
  ))
}

# SubCellType / Phase are produced by downstream annotation + cell-cycle scoring
# and are not present at the QC/doublet stage; only draw this when they exist.
if (all(c("SubCellType", "Phase") %in% meta_cols)) {
  df <- seurat_object@meta.data %>%
    dplyr::mutate(
      SubCellType = as.character(SubCellType),
      Phase = as.character(Phase)
    ) %>%
    dplyr::filter(!is.na(SubCellType), !is.na(Phase)) %>%
    dplyr::count(SubCellType, Phase) %>%
    dplyr::group_by(SubCellType) %>%
    dplyr::mutate(percent = 100 * n / sum(n))

  print(ggplot(df, aes(x = SubCellType, y = percent, fill = Phase)) +
    geom_col() +
    geom_text(aes(label = paste0(round(percent, 1), "%")),
              position = position_stack(vjust = 0.5), size = 3, color = "white") +
    scale_y_continuous(labels = scales::percent_format(scale = 1), expand = c(0, 0)) +
    labs(x = "SubCellType", y = "Percentage", fill = "Phase") +
    theme_minimal() +
    theme(legend.position = "right", legend.title = element_text(face = "bold")) +
    scale_fill_manual(values = c("G1" = "#A6CEE3", "S" = "#1F78B4", "G2M" = "#B2DF8A")))
}



```
:::{.callout-tip}
The @fig-doublet-composition provides doublets/singlets distribution across all cells and clusters per patient.
:::

## Doublet call validation {#doublet-validation}

```{r dbl-validate-compute, include = FALSE}
suppressPackageStartupMessages({library(ggplot2); library(data.table)})
.dm <- as.data.table(seurat_object@meta.data, keep.rownames = "cell_id")
.has_cls <- "class" %in% names(.dm)
.cnt <- intersect(c("nCount_RNA","nCount"), names(.dm))[1]
.ftr <- intersect(c("nFeature_RNA","nFeature"), names(.dm))[1]

if (.has_cls) {
  .dm[, class := as.character(class)]
  .n_dbl  <- sum(.dm$class == "doublet", na.rm = TRUE)
  .rate   <- 100 * .n_dbl / nrow(.dm)
  # A doublet is two cells in one droplet, so it should carry MORE RNA than a
  # singlet. If it does not, the caller is flagging something else and the
  # filtering is removing real cells. This is the check that makes the call
  # believable rather than merely reported.
  .ratio <- NA_real_; .pval <- NA_real_
  if (!is.na(.cnt) && .n_dbl > 2) {
    .md_d <- median(.dm[class == "doublet"][[.cnt]], na.rm = TRUE)
    .md_s <- median(.dm[class == "singlet"][[.cnt]], na.rm = TRUE)
    .ratio <- .md_d / max(1, .md_s)
    .pval <- tryCatch(suppressWarnings(
      wilcox.test(.dm[class == "doublet"][[.cnt]],
                  .dm[class == "singlet"][[.cnt]])$p.value),
      error = function(e) NA_real_)
  }
}
```


```{r dbl-validate-finding, results = 'asis'}
if (.has_cls) {
  msg <- sprintf("scDblFinder flagged **%s of %s cells (%.1f%%)** as doublets.",
                 format(.n_dbl, big.mark = ","), format(nrow(.dm), big.mark = ","), .rate)
  if (!is.na(.ratio)) {
    msg <- paste(msg, sprintf(
      "Median library size is **%.2fx higher** in flagged doublets than singlets%s.",
      .ratio, if (!is.na(.pval)) sprintf(" (Wilcoxon p = %s)",
                format.pval(.pval, digits = 2, eps = 1e-300)) else ""))
    if (.ratio < 1.15) msg <- paste(msg,
      "**A ratio near 1 means the flagged cells are not obviously double-loaded.**",
      "Two cells in one droplet should carry visibly more RNA; without that",
      "signal the caller may be removing real cells, so check the figure below",
      "before trusting the filtered object.")
  }
  # 10x doublet rate is ~0.8% per 1,000 cells recovered.
  # 10x multiplet rate is ~0.8% per 1,000 cells recovered -> a PERCENTAGE.
  .expected <- 0.8 * nrow(.dm) / 1000
  if (.rate > 2 * .expected) msg <- paste(msg, sprintf(
    "The observed rate is well above the ~%.1f%% expected for %s cells at 10x loading densities.",
    .expected, format(nrow(.dm), big.mark = ",")))
  scratch_finding(msg, type = if (!is.na(.ratio) && .ratio < 1.15) "caution" else "note",
                  title = "Doublet detection")

  # --- the two callers are never compared anywhere else -----------------
  if ("cluster_class" %in% names(.dm)) {
    .cc <- .dm[!is.na(class) & !is.na(cluster_class)]
    if (nrow(.cc)) {
      .agree <- sum(.cc$class == as.character(.cc$cluster_class))
      .pct <- 100 * .agree / nrow(.cc)
      m2 <- sprintf(paste("The sample-based and cluster-based callers agree on",
                          "**%.1f%%** of cells (%s of %s); they disagree on **%s**."),
                    .pct, format(.agree, big.mark = ","),
                    format(nrow(.cc), big.mark = ","),
                    format(nrow(.cc) - .agree, big.mark = ","))
      m2 <- paste(m2, sprintf("Only the **%s** mode is applied downstream (--qc_doublet_filter).",
                              if (is.null(params$doublet_filter)) "sample" else params$doublet_filter))
      scratch_finding(m2, type = if (.pct < 90) "caution" else "note",
                      title = "Caller agreement")
      scratch_table(dcast(.cc[, .N, by = .(class, cluster_class)],
                          class ~ cluster_class, value.var = "N", fill = 0),
                    caption = "sample-based (rows) vs cluster-based (columns) calls")
    }
  }
}
```


```{r dbl-validate-figure, message = FALSE, warning = FALSE}
#| fig-width: 8.5
#| fig-height: 4
if (.has_cls && !is.na(.cnt)) {
  .pl <- rbindlist(list(
    data.table(class = .dm$class, metric = "UMIs per cell",  value = as.numeric(.dm[[.cnt]])),
    if (!is.na(.ftr)) data.table(class = .dm$class, metric = "genes per cell",
                                 value = as.numeric(.dm[[.ftr]]))), use.names = TRUE)
  .pl <- .pl[is.finite(value) & value > 0 & !is.na(class)]
  # Violin, not violin+jitter: overlaying 18,000 points reproduces exactly the
  # overplotting the density plots exist to avoid. The quartile lines carry the
  # same information the jitter was there to suggest.
  # Quartiles come from a narrow boxplot inside the violin rather than
  # geom_violin's quantile argument: that argument was renamed in ggplot2 4.0,
  # so either spelling warns or errors depending on the installed version.
  # This idiom is stable across versions and reads more clearly anyway.
  ggplot(.pl, aes(class, value, fill = class)) +
    geom_violin(scale = "width", linewidth = .3, colour = "grey25") +
    geom_boxplot(width = .12, outlier.shape = NA, linewidth = .3,
                 colour = "grey20", fill = "white", alpha = .9) +
    facet_wrap(~ metric, scales = "free_y") +
    scale_y_log10(labels = scales::comma) +
    scale_fill_manual(values = c(singlet = "#8fb4dd", doublet = "#8c1d18"), guide = "none") +
    labs(title = "Library size by doublet call",
         subtitle = "doublets should sit visibly higher; quartiles drawn",
         x = NULL, y = NULL) +
    theme_scratch()
}
```


## Removing doublets

```{r remove_doublets}

n_before <- ncol(seurat_object)

singlet_object <- subset(
  seurat_object,
  subset = class == "singlet"
)

n_after <- ncol(singlet_object)

# Inline `r n_after` printed as "2.4415^{4}" -- knitr renders a
# scientific-notation number as LaTeX. Format once, here, as in notebook_seurat_qc.
.fmt_n     <- function(x) format(x, big.mark = ",", scientific = FALSE)
n_before_fmt  <- .fmt_n(n_before)
n_after_fmt   <- .fmt_n(n_after)
n_removed_fmt <- .fmt_n(n_before - n_after)

```

The singlet subset is still a lazy BPCells view over the full merged store. So
that downstream steps (which rebuild the object from `data/bpcells_counts`)
never see the removed doublets, we write a fresh store containing **only** the
singlet counts and point the saved object at it.

```{r write_singlet_store}

store_dir <- file.path('data', 'bpcells_counts')
tmp_store <- file.path('data', '.bpcells_counts_singlets')

# Lazy singlet counts (still reference the staged full store).
singlet_counts <- SeuratObject::LayerData(singlet_object, assay = 'RNA', layer = 'counts')

# Write the singlet-only store to a temp dir FIRST — this reads from the staged
# full store, so it must exist until the write completes.
if (dir.exists(tmp_store)) unlink(tmp_store, recursive = TRUE, force = TRUE)
BPCells::write_matrix_dir(mat = singlet_counts, dir = tmp_store)

# Now replace the staged store (a symlink to the merge task's output) with the
# singlet store at the path downstream steps expect.
unlink(store_dir, recursive = TRUE, force = TRUE)
file.rename(tmp_store, store_dir)
singlet_counts <- BPCells::open_matrix_dir(dir = store_dir)

# Rebuild the object from the singlet store so its counts layer references the
# new on-disk store (drop auto-computed count columns; they are regenerated).
keep_meta <- singlet_object@meta.data
keep_meta <- keep_meta[, setdiff(colnames(keep_meta),
                                 c('nCount_RNA', 'nFeature_RNA')), drop = FALSE]

singlet_object <- CreateSeuratObject(
  counts    = singlet_counts,
  project   = project_name,
  meta.data = keep_meta
)

```

::: {.callout-warning}
Total cells after doublet removal **`r n_after_fmt`** (removed **`r n_removed_fmt`** of **`r n_before_fmt`**).
:::

## Saving Seurat object

```{r object_dump}

if(auto_save) {

  saveRDS(singlet_object, file = paste0('data/', project_name, '_qc_dbl_singlet_object.RDS'))

}

```

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

print(params)

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```