---
title: "Module - Normalization and Dimensionality Reduction"
description: | 
  Description
execute:
  freeze: auto
  cache: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: 'Test'
  seurat_object: './data/ovarian_qc_merged_object.RDS'
  thr_n_features: 2000
  thr_n_dimensions: 100
  n_threads: 8
  n_memory: 8
  work_directory: !expr here::here()
  auto_save: TRUE
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE, message = FALSE, warning = FALSE, echo = FALSE}

# Project parameters 
project_name <- params$project_name
seurat_object <- params$seurat_object

# Inputs and thresholds
thr_n_features <- params$thr_n_features
thr_n_dimensions <- params$thr_n_dimensions

# Optional parameters

# Dataflow/Computational parameters
n_threads <- params$n_threads
n_memory <- params$n_memory

# Output parameters
work_directory <- params$work_directory
timestamp <- params$timestamp
auto_save <- params$auto_save

# Knitr options
knitr::opts_knit$set(
  root.dir = work_directory
  )

```

## Loading library

```{r library}
#| output: false
#| code-fold: true

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(BPCells)

```

## Folder structure

```{r project_directory}
#| output: false
#| code-fold: true

if(!dir.exists(work_directory)) {
  dir.create(work_directory, recursive = T)
}

# showWarnings = FALSE: data/ already exists (the BPCells store is staged there).
for(sub_directory in c('data', 'figures')) {
  dir.create(paste0(work_directory, '/', sub_directory), showWarnings = FALSE)
}

```

## Loading Seurat object

```{r loading_rds}

# The merged object is BPCells-backed. Its counts live in the on-disk store
# staged at data/bpcells_counts. Rebuild the object from that store plus the
# saved per-cell metadata so we never depend on the (possibly stale) matrix
# path baked into the RDS by the upstream task.
loaded_object <- readRDS(file = seurat_object)

disk_counts <- BPCells::open_matrix_dir(dir = file.path('data', 'bpcells_counts'))

keep_meta <- loaded_object@meta.data
keep_meta <- keep_meta[, setdiff(colnames(keep_meta),
                                 c('nCount_RNA', 'nFeature_RNA')), drop = FALSE]

seurat_object <- CreateSeuratObject(
  counts    = disk_counts,
  project   = project_name,
  meta.data = keep_meta
)

rm(loaded_object); gc()

# Setting random seed
random_seed <- 2203
set.seed(random_seed)

```

## Unifying sample layers

```{r seurat_rejoin}
#| code-fold: true

if(length(seurat_object@assays$RNA@layers) > 1) {
  
  seurat_object <- JoinLayers(seurat_object)  

}

```

## Counts normalization

```{r normalization, message = FALSE, warning = FALSE, echo = TRUE}

seurat_object <- NormalizeData(
  object = seurat_object, 
  normalization.method = "LogNormalize",
  scale.factor = 10000)

seurat_object <- FindVariableFeatures(
  seurat_object, selection.method = "vst", nfeatures = thr_n_features)

seurat_object <- ScaleData(seurat_object)

```

## Dimensionality Reduction

```{r dimensionality_reduction}
#| warning: false

# Running PCA and printing out dimensions
seurat_object <- RunPCA(seurat_object, verbose = FALSE, npcs = thr_n_dimensions)
print(seurat_object[["pca"]], dims = 1:5, nfeatures = 5)

```

### PCA Loadings

```{r pca_loading_plot, fig.height = 14}
#| label: fig-loadings
#| fig-cap: 'PCA Loadings'
#| fig-cap-location: top
#| fig-height: 14
#| fig-width: 10

VizDimLoadings(seurat_object, dims = 1:5, reduction = "pca")

```

:::{.callout-tip}
The @fig-loadings visualizes dimensionality reduction loadings in a Seurat object. It displays the contribution of each gene to the principal components or other reduced dimensions, aiding in the interpretation of underlying biological variation.
:::

### Features Plot

```{r variable_plot}
#| label: fig-variable-plot
#| fig-cap: 'VariableFeaturePlot (TOP 10)'
#| fig-cap-location: top
#| warning: false

# Identify the 10 most highly variable genes
variable_top10 <- head(VariableFeatures(seurat_object), 10)

# plot variable features with and without labels
variable_plot <- VariableFeaturePlot(seurat_object)
variable_plot <- LabelPoints(
  plot = variable_plot, points = variable_top10, repel = TRUE)

variable_plot

```

:::{.callout-tip}
The @fig-variable-plot displays the variability of features across cells. It highlights genes with high variance, providing insights into the heterogeneity of gene expression within a dataset.
:::

### Elbow Plot

```{r elbow_plot}
#| label: fig-elbow
#| fig-cap: 'ElbowPlot'
#| fig-cap-location: top

elbow_plot <- ElbowPlot(object = seurat_object, ndims = 100)
elbow_plot

```

:::{.callout-tip}
The @fig-elbow illustrates the elbow point of the variance explained by principal components. It helps in determining the optimal number of principal components to retain for downstream analysis, such as dimensionality reduction or clustering.
:::

## Reduction diagnostics {#reduction-diagnostics}

```{r reduction-compute, include = FALSE}
suppressPackageStartupMessages({library(ggplot2); library(data.table)})

# --- variance explained, and where it actually plateaus ---------------------
sdev <- tryCatch(Seurat::Stdev(seurat_object, reduction = "pca"),
                 error = function(e) seurat_object[["pca"]]@stdev)
varexp <- sdev^2 / sum(sdev^2)
cumvar <- cumsum(varexp)

# The "elbow" chosen by eye is not reproducible. Take the knee as the point of
# maximum distance from the straight line joining the first and last PC of the
# scree curve -- the standard geometric definition, so the number is defensible
# and identical on a re-run.
knee_of <- function(y) {
  n <- length(y); if (n < 3) return(n)
  x <- seq_len(n)
  x1 <- 1; y1 <- y[1]; x2 <- n; y2 <- y[n]
  d <- abs((y2 - y1) * x - (x2 - x1) * y + x2 * y1 - y2 * x1) /
       sqrt((y2 - y1)^2 + (x2 - x1)^2)
  which.max(d)
}
n_knee <- knee_of(varexp)
n_used <- thr_n_dimensions

# --- are the top PCs technical? --------------------------------------------
# sc-best-practices asks for PCA coloured by QC metrics. A correlation table
# answers the same question quantitatively and in one screen instead of a wall
# of scatter panels: if PC1 tracks library size, the embedding is reporting
# sequencing depth rather than biology.
emb <- Seurat::Embeddings(seurat_object, reduction = "pca")
md  <- seurat_object@meta.data
qc_cols <- intersect(c("nCount_RNA", "nFeature_RNA", "percent_mito", "percent.mt"),
                     names(md))
npc_check <- min(15L, ncol(emb))
cor_dt <- NULL
if (length(qc_cols) && npc_check > 0) {
  cor_dt <- rbindlist(lapply(qc_cols, function(q) {
    v <- suppressWarnings(as.numeric(md[[q]]))
    data.table(metric = q, pc = seq_len(npc_check),
               r = vapply(seq_len(npc_check), function(i)
                     suppressWarnings(cor(emb[, i], v, method = "spearman",
                                          use = "complete.obs")), numeric(1)))
  }))
  cor_dt[is.na(r), r := 0]
}
```

```{r reduction-finding, results = 'asis'}
msg <- sprintf(paste("The first **%d** principal components capture **%.1f%%** of the",
                     "variance; **%d** are used downstream, capturing **%.1f%%**.",
                     "The scree curve's knee is at **PC%d**."),
               n_knee, 100 * cumvar[n_knee], n_used,
               100 * cumvar[min(n_used, length(cumvar))], n_knee)
if (n_used > 3 * n_knee) msg <- paste(msg,
  sprintf(paste("Using %d PCs is well past the knee. Extra components mostly add noise",
                "to the neighbour graph, which tends to fragment clusters rather than",
                "resolve them."), n_used))
if (n_used < n_knee) msg <- paste(msg,
  sprintf(paste("**Using %d PCs cuts below the knee at PC%d**, so structure the data",
                "supports is being discarded before clustering."), n_used, n_knee))
scratch_finding(msg, type = if (n_used < n_knee) "caution" else "note",
                title = "Dimensionality")

if (!is.null(cor_dt) && nrow(cor_dt)) {
  worst <- cor_dt[which.max(abs(r))]
  m2 <- sprintf(paste("The strongest association between a top-%d principal component",
                      "and a technical covariate is **%s vs PC%d (Spearman r = %.2f)**."),
                npc_check, worst$metric, worst$pc, worst$r)
  if (abs(worst$r) >= 0.5) m2 <- paste(m2,
    "**A correlation this strong means the embedding is partly reporting technical",
    "variation, not biology.** Clusters split along this axis should be treated as",
    "suspect until regression or integration removes it.")
  else if (abs(worst$r) >= 0.3) m2 <- paste(m2,
    "Moderate technical loading -- worth knowing, but not on its own a reason to act.")
  else m2 <- paste(m2, "No top component is strongly driven by depth or",
    "mitochondrial content, which is what you want to see here.")
  scratch_finding(m2, type = if (abs(worst$r) >= 0.5) "caution" else "note",
                  title = "Technical loading on the top PCs")
}
```

```{r scree-figure, message = FALSE, warning = FALSE}
#| fig-width: 8.5
#| fig-height: 4
sc <- data.table(pc = seq_along(varexp), var = 100 * varexp, cum = 100 * cumvar)
sc <- sc[pc <= min(60, .N)]
sc[, tip := sprintf("PC%d\n%.2f%% of variance\n%.1f%% cumulative", pc, var, cum)]
ggplot(sc, aes(pc, var, text = tip)) +
  geom_col(fill = "#3f6fa8", width = .8) +
  geom_vline(xintercept = n_knee, linetype = "dotted", colour = "#0d366b") +
  geom_vline(xintercept = n_used, linetype = "dashed", colour = "#8a5a00") +
  labs(title = "Variance explained per principal component",
       subtitle = sprintf("dotted = knee (PC%d); dashed = PCs used downstream (%d)",
                          n_knee, n_used),
       x = "principal component", y = "% of total variance") +
  theme_scratch()
```

```{r pc-qc-figure, message = FALSE, warning = FALSE}
#| fig-width: 9
#| fig-height: 3.2
if (!is.null(cor_dt) && nrow(cor_dt)) {
  cd <- copy(cor_dt)
  cd[, tip := sprintf("PC%d vs %s\nSpearman r = %.2f", pc, metric, r)]
  ggplot(cd, aes(factor(pc), metric, fill = r, text = tip)) +
    geom_tile(colour = "white", linewidth = .4) +
    scale_fill_gradient2(low = "#8c1d18", mid = "#f4f4f4", high = "#0d366b",
                         midpoint = 0, limits = c(-1, 1), name = "Spearman r") +
    labs(title = "Correlation of each principal component with QC metrics",
         subtitle = "strong colour = that component is tracking a technical covariate",
         x = "principal component", y = NULL) +
    theme_scratch()
}
```

```{r norm-figure, message = FALSE, warning = FALSE}
#| fig-width: 9
#| fig-height: 3.6
# Best practice asks for the raw and normalised distributions side by side --
# the point of normalisation is to remove the depth-driven spread, and the only
# way to see whether it did is to look at both.
cnt <- suppressWarnings(as.numeric(seurat_object$nCount_RNA))
lay <- tryCatch(SeuratObject::LayerData(seurat_object, layer = "data"),
                error = function(e) NULL)
cell_sum <- if (!is.null(lay)) Matrix::colSums(lay) else NULL
if (!is.null(cell_sum)) {
  nd <- rbind(
    data.table(stage = "raw UMIs per cell (log10)",        value = log10(cnt + 1)),
    data.table(stage = "log-normalised total per cell",    value = as.numeric(cell_sum)))
  ggplot(nd, aes(value, fill = stage)) +
    geom_histogram(bins = 60, show.legend = FALSE) +
    facet_wrap(~ stage, scales = "free") +
    scale_fill_manual(values = c("#8fb4dd", "#0d366b")) +
    labs(title = "Count distribution before and after normalisation",
         subtitle = "normalisation should compress the depth-driven spread on the left",
         x = NULL, y = "cells") +
    theme_scratch()
}
```

## Saving Seurat object

```{r object_dump}

if(auto_save) {
  
  saveRDS(seurat_object, file = paste0('data/', project_name, '_reduction_object.RDS'))
  
}

```

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

params

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```
