---
title: "SCRATCH-E2E — Tumour microenvironment characterisation"
execute:
  freeze: false
  cache: false
  warning: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    embed-resources: true
    code-fold: show
    fig-format: png
    fig-dpi: 300
    fig-align: center
params:
  seurat_object: null
  project_name: "project"
  work_directory: "."
  celltype_col: "azimuth_labels"
  sample_col: "sample_id"
  condition_col: "condition"
  patient_col: "patient_id"
  min_cells_per_group: 30
  top_n_celltypes: 12
  seed: 1234
  run_scoda: false            # opt-in; needs the scCODA image (large)
  propeller_transform: "logit"
  min_celltype_frac: 0.005    # rare-type guard: >= 0.5% of cells ...
  min_celltype_prevalence: 0.5  # ... and present in >= 50% of samples
  scoda_reference: "automatic"
  scoda_fdr: 0.05
  scoda_python: ""   # empty -> $SCRATCH_SCODA_PYTHON, else python3
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE}
suppressPackageStartupMessages({
  library(Seurat)
  library(data.table)
  library(ggplot2)
  library(scales)
})

set.seed(params$seed)
if (!is.null(params$work_directory) && dir.exists(params$work_directory)) {
  setwd(params$work_directory)
}
dir.create("data",    showWarnings = FALSE, recursive = TRUE)
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

# Categorical hues assigned in fixed slot order and never cycled, so a cell type
# keeps its colour when others are filtered out. Sequential magnitude uses one
# hue, light to dark.
CAT <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
         "#e87ba4", "#008300", "#4a3aa7", "#e34948")
SEQ_LOW <- "#cde2fb"; SEQ_HIGH <- "#0d366b"
DIV_LOW <- "#2a78d6"; DIV_MID <- "#f0efec"; DIV_HIGH <- "#e34948"
INK <- "#0f1417"; INK2 <- "#4e5a60"; MUTED <- "#898781"
GRID <- "#e1e0d9"; SURFACE <- "#fcfcfb"

theme_scratch <- function(base = 12) {
  theme_minimal(base_size = base) +
    theme(plot.background  = element_rect(fill = SURFACE, colour = NA),
          panel.background = element_rect(fill = SURFACE, colour = NA),
          panel.grid.major = element_line(colour = GRID, linewidth = .3),
          panel.grid.minor = element_blank(),
          axis.text  = element_text(colour = MUTED),
          axis.title = element_text(colour = INK2),
          plot.title = element_text(colour = INK, face = "bold"),
          plot.subtitle = element_text(colour = INK2, size = base * .9),
          legend.text = element_text(colour = INK2),
          legend.title = element_text(colour = INK2),
          strip.text = element_text(colour = INK2, face = "bold"))
}

fold_other <- function(x, k) {
  x <- as.character(x)
  keep <- names(sort(table(x), decreasing = TRUE))[seq_len(min(k, length(unique(x))))]
  factor(ifelse(x %in% keep, x, "Other"), levels = c(keep, "Other"))
}
note <- function(...) cat("\n_", ..., "_\n\n", sep = "")
save_fig <- function(p, n, w = 9, h = 5.5) {
  ggsave(file.path("figures", paste0(n, ".png")), p, width = w, height = h, dpi = 400); p
}
```

## Scope

This runs on the **batch-corrected non-malignant compartment** produced by the
stratification step. Malignant cells are analysed separately by the metaprogram
stage; excluding them here is what makes composition comparable between
patients, since tumour content otherwise dominates every proportion.

```{r load}
stopifnot(!is.null(params$seurat_object))
obj <- readRDS(params$seurat_object)
cat("Non-malignant compartment:", ncol(obj), "cells x", nrow(obj), "features\n")

meta <- as.data.table(obj@meta.data, keep.rownames = "cell_id")

pick <- function(pref, alts) {
  cand <- c(pref, alts); hit <- cand[cand %in% names(meta)]
  if (length(hit)) hit[1] else NA_character_
}
col_ct   <- pick(params$celltype_col,   c("azimuth_labels","celltype","cell_type","annotation","sctype","sctype_agg","sctype_labels"))
col_smp  <- pick(params$sample_col,     c("sample_id","orig.ident","sample"))
col_cond <- pick(params$condition_col,  c("condition","timepoint","group","response"))
col_pat  <- pick(params$patient_col,    c("patient_id","patient","donor"))

cat("Resolved -> celltype:", col_ct, "| sample:", col_smp,
    "| condition:", col_cond, "| patient:", col_pat, "\n")

if (is.na(col_ct)) stop("No cell-type annotation column found; TME characterisation needs one.")
meta[, ct := fold_other(get(col_ct), as.integer(params$top_n_celltypes))]
```

## Composition

```{r composition-sample, fig.width = 10, fig.height = 5.5}
if (!is.na(col_smp)) {
  comp <- meta[, .N, by = .(smp = get(col_smp), ct)][, frac := N / sum(N), by = smp]
  fwrite(comp, "data/tme_composition_by_sample.csv")

  p <- ggplot(comp, aes(x = smp, y = frac, fill = ct)) +
    geom_col(width = .74, colour = SURFACE, linewidth = 1) +
    scale_fill_manual(values = CAT, name = "cell type") +
    scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, .02))) +
    labs(title = "Microenvironment composition by sample",
         subtitle = "Non-malignant cells only", x = NULL, y = "share of cells") +
    theme_scratch() + theme(axis.text.x = element_text(angle = 40, hjust = 1))
  print(save_fig(p, "tme_composition_by_sample", 10, 5.5))
} else note("No sample column; per-sample composition skipped.")
```

```{r composition-condition, fig.width = 9, fig.height = 5.5}
if (!is.na(col_cond) && uniqueN(meta[[col_cond]]) > 1) {
  comp <- meta[, .N, by = .(cond = get(col_cond), ct)][, frac := N / sum(N), by = cond]
  fwrite(comp, "data/tme_composition_by_condition.csv")

  p <- ggplot(comp, aes(x = cond, y = frac, fill = ct)) +
    geom_col(width = .6, colour = SURFACE, linewidth = 1) +
    scale_fill_manual(values = CAT, name = "cell type") +
    scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, .02))) +
    labs(title = "Microenvironment composition by condition", x = NULL, y = "share of cells") +
    theme_scratch()
  print(save_fig(p, "tme_composition_by_condition"))
} else note("Fewer than two conditions; comparison skipped.")
```

## Differential abundance

```{r abundance, fig.width = 9, fig.height = 5}
# Per-sample proportions are the unit of analysis: testing per cell would treat
# thousands of cells from one patient as independent observations.
if (!is.na(col_smp) && !is.na(col_cond) && uniqueN(meta[[col_cond]]) > 1) {

  props <- meta[, .N, by = .(smp = get(col_smp), cond = get(col_cond), ct)]
  props[, total := sum(N), by = smp]
  props <- props[total >= as.integer(params$min_cells_per_group)]
  props[, frac := N / total]

  # Fill absent cell type / sample combinations with zero.
  grid <- CJ(smp = unique(props$smp), ct = unique(props$ct), unique = TRUE)
  key  <- unique(props[, .(smp, cond)])
  full <- merge(merge(grid, key, by = "smp"), props[, .(smp, ct, frac)],
                by = c("smp", "ct"), all.x = TRUE)
  full[is.na(frac), frac := 0]

  conds <- sort(unique(full$cond))

  # Omnibus test per cell type. Two groups -> Wilcoxon; more -> Kruskal-Wallis.
  # Either way the unit of observation is the SAMPLE proportion, not the cell:
  # testing per cell would treat thousands of cells from one patient as
  # independent and manufacture significance.
  omni <- full[, {
    grp <- split(frac, cond)
    grp <- grp[vapply(grp, length, integer(1)) >= 2]
    if (length(grp) < 2) {
      .(test = NA_character_, statistic = NA_real_, p = NA_real_, n_groups = length(grp))
    } else if (length(grp) == 2) {
      w <- suppressWarnings(stats::wilcox.test(grp[[1]], grp[[2]]))
      .(test = "Wilcoxon", statistic = unname(w$statistic), p = w$p.value, n_groups = 2L)
    } else {
      k <- suppressWarnings(stats::kruskal.test(frac ~ factor(cond)))
      .(test = "Kruskal-Wallis", statistic = unname(k$statistic), p = k$p.value,
        n_groups = length(grp))
    }
  }, by = ct]
  omni[, padj := p.adjust(p, method = "BH")]

  # Mean proportion per condition, joined on so the table is readable alone.
  means <- dcast(full[, .(m = mean(frac)), by = .(ct, cond)], ct ~ cond, value.var = "m")
  setnames(means, setdiff(names(means), "ct"), paste0("mean_", setdiff(names(means), "ct")))
  res <- merge(omni, means, by = "ct")
  setorder(res, padj, na.last = TRUE)
  fwrite(res, "data/tme_differential_abundance.csv")
  knitr::kable(res, digits = 4)

  # Every pairwise contrast, so an omnibus hit can be attributed to a pair.
  if (length(conds) > 2) {
    pw <- rbindlist(lapply(utils::combn(conds, 2, simplify = FALSE), function(cc) {
      full[, {
        a <- frac[cond == cc[1]]; b <- frac[cond == cc[2]]
        if (length(a) >= 2 && length(b) >= 2) {
          w <- suppressWarnings(stats::wilcox.test(a, b))
          .(group1 = cc[1], group2 = cc[2], mean1 = mean(a), mean2 = mean(b),
            log2fc = log2((mean(b) + 1e-4) / (mean(a) + 1e-4)), p = w$p.value)
        } else .(group1 = cc[1], group2 = cc[2], mean1 = mean(a), mean2 = mean(b),
                 log2fc = NA_real_, p = NA_real_)
      }, by = ct]
    }))
    pw[, padj := p.adjust(p, method = "BH")]      # corrected across ALL contrasts
    setorder(pw, padj, na.last = TRUE)
    fwrite(pw, "data/tme_differential_abundance_pairwise.csv")
  }

  # Plot: two groups -> signed fold change; more -> proportion heatmap with the
  # omnibus result marked, since a single log2FC has no meaning across >2 levels.
  if (length(conds) == 2 && nrow(res[is.finite(p)])) {
    fc <- merge(full[, .(m = mean(frac)), by = .(ct, cond)][, dcast(.SD, ct ~ cond, value.var = "m")],
                res[, .(ct, padj)], by = "ct")
    setnames(fc, conds, c("g1", "g2"))
    fc[, log2fc := log2((g2 + 1e-4) / (g1 + 1e-4))]
    p <- ggplot(fc, aes(reorder(ct, log2fc), log2fc, fill = log2fc)) +
      geom_col(width = .68) + geom_hline(yintercept = 0, colour = MUTED, linewidth = .4) +
      scale_fill_gradient2(low = DIV_LOW, mid = DIV_MID, high = DIV_HIGH, midpoint = 0, guide = "none") +
      geom_text(aes(label = ifelse(!is.na(padj) & padj < .05, "*", "")), hjust = -.4, size = 5) +
      coord_flip() +
      labs(title = sprintf("Differential abundance: %s vs %s", conds[2], conds[1]),
           subtitle = "Per-sample proportions, Wilcoxon, BH-adjusted; * = padj < 0.05",
           x = NULL, y = sprintf("log2 fold change (%s / %s)", conds[2], conds[1])) +
      theme_scratch()
    print(save_fig(p, "tme_differential_abundance"))
  } else if (length(conds) > 2) {
    hm <- full[, .(m = mean(frac)), by = .(ct, cond)]
    hm <- merge(hm, res[, .(ct, padj)], by = "ct")
    hm[, lab := ifelse(!is.na(padj) & padj < .05, paste0(ct, " *"), as.character(ct))]
    p <- ggplot(hm, aes(cond, reorder(lab, m), fill = m)) +
      geom_tile(colour = SURFACE, linewidth = 2) +
      geom_text(aes(label = percent(m, accuracy = 1)),
                colour = ifelse(hm$m > max(hm$m) * .55, "#ffffff", INK), size = 3) +
      scale_fill_gradient(low = SEQ_LOW, high = SEQ_HIGH, labels = percent_format(accuracy = 1),
                          name = "mean share") +
      labs(title = sprintf("Composition across %d conditions", length(conds)),
           subtitle = "Mean per-sample proportion; * = Kruskal-Wallis padj < 0.05 across all groups",
           x = NULL, y = NULL) +
      theme_scratch() + theme(panel.grid = element_blank())
    print(save_fig(p, "tme_differential_abundance", 9, 5.5))
  }

  scratch_finding(paste0(
    "Proportions are **compositional** (they sum to 1), so one population expanding ",
    "mechanically shrinks the others. These rank-based tests do not model that — ",
    "treat this table as a **screen**. The scCODA section below is the confirmatory ",
    "test that does model the constraint."),
    type = "caution", title = "Interpreting these p-values")

} else note("Needs a sample column and two conditions; differential abundance skipped.")
```

### Compositional test

::: {.callout-note collapse="true"}
## Why the table above needs a second test

The screen above asks "did this cell type's proportion move?". Because proportions
are locked to a simplex, a proportion can move purely because something *else*
moved -- a tumour-rich sample deflates every immune fraction at once without a
single immune cell being lost.

Two methods are available here. **propeller** (`speckle`) is the default: it
variance-stabilises the proportions with a logit transform and fits them with
limma's empirical Bayes, giving one p-value per cell type and an ANOVA F-test
across more than two groups. **scCODA** is opt-in (`run_scoda: true`) and models
the per-sample counts as Dirichlet-multinomial with spike-and-slab selection.
scCODA is the more principled treatment of the constraint, but it reports effects
only *relative to a reference cell type* and costs a much larger container image,
so it is a confirmatory tool rather than the default.
:::

```{r propeller, fig.width = 9, fig.height = 5}
# Rare cell types are the failure mode here. A logit transform on a population
# that is a fraction of a percent of the sample turns sampling noise into a very
# small p-value: on the HGSOC cohort an HSC population of 112 cells (0.05% of
# cells, absent from 15 of 39 samples) came out at FDR 1e-06. Types that thin are
# tested separately and reported as unreliable rather than silently ranked
# alongside real results.
prop_res <- NULL; prop_msg <- NULL; prop_rare <- NULL

if (is.na(col_smp) || is.na(col_cond) || uniqueN(meta[[col_cond]]) < 2) {
  prop_msg <- "Needs a sample column and at least two conditions."
} else if (!requireNamespace("speckle", quietly = TRUE)) {
  prop_msg <- paste0("The `speckle` package is not installed in this image, so the ",
                     "compositional test did not run. `BiocManager::install('speckle')`.")
} else {
  cells <- meta[, .(smp = as.character(get(col_smp)),
                    cond = as.character(get(col_cond)), ct = as.character(ct))]

  # Prevalence guard: a type must clear a minimum share AND be present in enough
  # samples for a per-sample proportion to mean anything.
  share <- cells[, .N, by = ct][, frac := N / sum(N)]
  present <- cells[, .(n_smp = uniqueN(smp)), by = ct]
  guard <- merge(share, present, by = "ct")
  min_frac <- as.numeric(params$min_celltype_frac)
  min_smp  <- max(3L, ceiling(uniqueN(cells$smp) * as.numeric(params$min_celltype_prevalence)))
  guard[, reliable := frac >= min_frac & n_smp >= min_smp]
  prop_rare <- guard[reliable == FALSE]

  keep_ct <- guard[reliable == TRUE, ct]
  if (length(keep_ct) < 2) {
    prop_msg <- sprintf(paste0("Only %d cell type(s) clear the prevalence guard ",
                               "(>= %.1f%% of cells and present in >= %d samples); ",
                               "too few to test."),
                        length(keep_ct), 100 * min_frac, min_smp)
  } else {
    cc <- cells[ct %in% keep_ct]
    pr <- try(speckle::propeller(clusters = cc$ct, sample = cc$smp, group = cc$cond,
                                 transform = params$propeller_transform), silent = TRUE)
    if (inherits(pr, "try-error")) {
      prop_msg <- paste("propeller failed:", sub("\n.*", "", as.character(pr)))
    } else {
      prop_res <- as.data.table(pr, keep.rownames = "cell_type")
      fwrite(prop_res, "data/tme_differential_abundance_propeller.csv")
      if (nrow(prop_rare)) fwrite(prop_rare, "data/tme_rare_celltypes_excluded.csv")
    }
  }
}

if (is.null(prop_res)) {
  scratch_finding(paste0(
    "**The compositional test did not run.** ", prop_msg,
    " The differential-abundance table above is therefore a non-compositional ",
    "screen with nothing confirming it -- do not report its p-values as evidence ",
    "of a real abundance change."),
    type = "warning", title = "Compositional test not run")
} else {
  pcol <- intersect(c("FDR", "P.Value"), names(prop_res))[1]
  hits <- prop_res[get(pcol) < 0.05, cell_type]

  scratch_table(prop_res, caption = sprintf(
    "propeller (%s transform, %s). Rare types excluded by the prevalence guard.",
    params$propeller_transform,
    if ("Fstatistic" %in% names(prop_res)) "ANOVA across groups" else "two-group test"))

  stat_col <- intersect(c("Fstatistic", "Tstatistic"), names(prop_res))[1]
  if (!is.na(stat_col) && nrow(prop_res)) {
    pl <- copy(prop_res)[, `:=`(sig = get(pcol) < 0.05,
                                nlp = -log10(pmax(get(pcol), .Machine$double.xmin)))]
    p <- ggplot(pl, aes(reorder(cell_type, nlp), nlp, fill = sig)) +
      geom_col(width = .68) +
      geom_hline(yintercept = -log10(0.05), linetype = "22", colour = MUTED, linewidth = .4) +
      scale_fill_manual(values = c(`TRUE` = CAT[1], `FALSE` = MUTED), guide = "none") +
      coord_flip() +
      labs(title = "Compositional differential abundance (propeller)",
           subtitle = sprintf("%s transform; dashed line = FDR 0.05",
                              params$propeller_transform),
           x = NULL, y = expression(-log[10]~FDR)) +
      theme_scratch()
    print(save_fig(p, "tme_propeller"))
  }

  if (nrow(prop_rare)) {
    scratch_finding(paste0(
      sprintf("**%d cell type(s) were excluded as too rare to test**: %s. ",
              nrow(prop_rare), paste(prop_rare$ct, collapse = ", ")),
      "A logit transform on a population that is a fraction of a percent turns ",
      "sampling noise into a very small p-value, so these are reported here ",
      "rather than ranked with the results above. Counts are in ",
      "`data/tme_rare_celltypes_excluded.csv`."),
      type = "caution", title = "Rare cell types excluded")
  }

  screen_hits <- if (exists("res")) res[!is.na(padj) & padj < .05, as.character(ct)] else character()
  only_screen <- setdiff(intersect(screen_hits, prop_res$cell_type), hits)
  scratch_finding(paste0(
    sprintf("propeller finds **%d cell type%s** with a credible compositional change ",
            length(hits), if (length(hits) == 1) "" else "s"),
    sprintf("at FDR 0.05%s. ",
            if (length(hits)) paste0(": ", paste(hits, collapse = ", ")) else ""),
    if (length(only_screen))
      sprintf(paste0("**%s** reached padj < 0.05 in the rank-based screen but not ",
                     "here -- consistent with those proportions moving because ",
                     "other populations moved. "),
              paste(only_screen, collapse = ", ")) else ""),
    type = if (length(only_screen)) "caution" else "info",
    title = "Compositional verdict")
}
```

### Confirmatory: scCODA {#scoda}

```{r scoda, fig.width = 9, fig.height = 5}
# Opt-in. Needs the scCODA image (containers/tme/Dockerfile.scoda); everything is
# guarded so a missing environment degrades visibly instead of failing.
scoda_res <- NULL
scoda_msg <- NULL

if (!isTRUE(as.logical(params$run_scoda))) {
  scoda_msg <- "Not enabled (`run_scoda: false`). propeller above is the default test."
} else if (is.na(col_smp) || is.na(col_cond) || uniqueN(meta[[col_cond]]) < 2) {
  scoda_msg <- "Needs a sample column and at least two conditions."
} else {
  cnt <- dcast(meta[, .N, by = .(smp = get(col_smp), ct)], smp ~ ct,
               value.var = "N", fill = 0)
  cov <- unique(meta[, .(smp = get(col_smp), condition = as.character(get(col_cond)))])
  cov <- cov[smp %in% cnt$smp][match(cnt$smp, smp)]

  d <- file.path(tempdir(), "scoda"); dir.create(d, showWarnings = FALSE)
  f_cnt <- file.path(d, "counts.csv"); f_cov <- file.path(d, "covariates.csv")
  f_out <- file.path(d, "effects.csv")
  fwrite(cnt, f_cnt); fwrite(cov, f_cov)

  # Nextflow puts the pipeline's bin/ on PATH, so the script resolves by name.
  py  <- if (nzchar(params$scoda_python)) params$scoda_python else
           Sys.getenv("SCRATCH_SCODA_PYTHON", "python3")
  scr <- Sys.which("scratch_scoda.py")
  if (!nzchar(scr)) for (cand in c("bin/scratch_scoda.py", "../bin/scratch_scoda.py",
                                   file.path(Sys.getenv("SCRATCH_BIN_DIR", "."), "scratch_scoda.py")))
    if (file.exists(cand)) { scr <- normalizePath(cand); break }

  if (!nzchar(scr)) {
    scoda_msg <- "`scratch_scoda.py` not found on PATH."
  } else {
    err <- file.path(d, "stderr.txt")
    st <- suppressWarnings(system2(py, c(shQuote(scr),
            "--counts", shQuote(f_cnt), "--covariates", shQuote(f_cov),
            "--formula", "condition",
            "--reference", shQuote(params$scoda_reference),
            "--fdr", params$scoda_fdr, "--seed", params$seed,
            "--out", shQuote(f_out)), stdout = TRUE, stderr = err))
    detail <- if (file.exists(err)) paste(readLines(err, warn = FALSE), collapse = " ") else ""
    code <- attr(st, "status"); code <- if (is.null(code)) 0L else code

    if (code == 2L) {
      scoda_msg <- paste0("scCODA is not installed in the notebook's Python (`", py,
                          "`). Use the scCODA image or set `scoda_python`. Detail: ", detail)
    } else if (code == 3L) {
      scoda_msg <- paste0("The design cannot support scCODA: ", detail)
    } else if (code != 0L || !file.exists(f_out)) {
      scoda_msg <- paste0("scCODA failed (exit ", code, "): ", detail)
    } else {
      scoda_res <- fread(f_out)
      fwrite(scoda_res, "data/tme_differential_abundance_scoda.csv")
    }
  }
}

if (is.null(scoda_res)) {
  note(scoda_msg)
} else {
  ref <- unique(scoda_res$reference_cell_type)[1]
  fdr <- unique(scoda_res$est_fdr)[1]

  # The reference cell type is the model's baseline: its effects are fixed at
  # zero by construction. Reporting it as "not credible" alongside genuinely
  # tested types would read as "no change found", which is not what it means.
  scoda_res[, status := fifelse(cell_type == ref, "reference (not estimable)",
                        fifelse(credible, "credible", "not credible"))]
  cred <- scoda_res[credible == TRUE & cell_type != ref]

  scratch_table(scoda_res[, .(covariate, cell_type, effect, log2fc,
                              inclusion_prob, status)],
                caption = sprintf("scCODA effects (reference: %s; FDR %.2f)", ref, fdr))

  if (nrow(cred)) {
    p <- ggplot(cred, aes(reorder(cell_type, log2fc), log2fc, fill = log2fc)) +
      geom_col(width = .68) +
      geom_hline(yintercept = 0, colour = MUTED, linewidth = .4) +
      scale_fill_gradient2(low = DIV_LOW, mid = DIV_MID, high = DIV_HIGH,
                           midpoint = 0, guide = "none") +
      coord_flip() +
      labs(title = "Credible compositional changes (scCODA)",
           subtitle = sprintf("Dirichlet-multinomial, FDR %.2f; effects are relative to %s",
                              fdr, ref),
           x = NULL, y = "log2 fold change") +
      theme_scratch()
    print(save_fig(p, "tme_scoda_effects"))
  }

  scratch_finding(paste0(
    sprintf("scCODA finds **%d credible change%s** at FDR %.2f, ",
            nrow(cred), if (nrow(cred) == 1) "" else "s", fdr),
    sprintf("all measured **relative to %s**, which is the model's baseline and ", ref),
    "therefore cannot itself be tested -- its absence from the credible list is a ",
    "structural property, not evidence that it is unchanged. ",
    if (nrow(cred)) paste0("Credible: ", paste(cred$cell_type, collapse = ", "), ".") else ""),
    type = "info", title = "scCODA verdict")
}
```

## Functional programmes

```{r signatures, fig.width = 10, fig.height = 6}
# Curated marker sets scored per cell. Genes absent from the object are dropped
# and the set is skipped if too little of it survives.
SIGNATURES <- list(
  Cytotoxicity  = c("GZMB","GZMA","GZMK","PRF1","NKG7","GNLY","KLRD1","IFNG"),
  Exhaustion    = c("PDCD1","CTLA4","LAG3","HAVCR2","TIGIT","TOX","ENTPD1"),
  Interferon    = c("ISG15","IFI6","IFIT1","IFIT3","MX1","OAS1","STAT1","IRF7"),
  Hypoxia       = c("VEGFA","SLC2A1","CA9","LDHA","PGK1","ADM","NDRG1"),
  Angiogenesis  = c("PECAM1","VWF","CDH5","KDR","ANGPT2","ESM1"),
  ECM_CAF       = c("COL1A1","COL1A2","COL3A1","FN1","POSTN","FAP","ACTA2","TAGLN"),
  Complement    = c("C1QA","C1QB","C1QC","C3","CFB"),
  Proliferation = c("MKI67","TOP2A","PCNA","CCNB1","BIRC5")
)

present <- rownames(obj)
sets <- lapply(SIGNATURES, function(g) intersect(g, present))
kept <- sets[vapply(sets, length, integer(1)) >= 3]

if (length(kept)) {
  # Not Seurat::AddModuleScore — it calls GetAssayData(slot=), defunct in
  # SeuratObject 5.x, and fails on ANY object under Seurat 5.3.0. Same method,
  # version-independent.
  obj <- scratch_module_score(obj, kept, name = "TMEsig", seed = params$seed)
  score_cols <- grep("^TMEsig\\d+$", colnames(obj@meta.data), value = TRUE)
  names(score_cols) <- names(kept)[seq_along(score_cols)]

  sc <- as.data.table(obj@meta.data[, score_cols, drop = FALSE], keep.rownames = "cell_id")
  setnames(sc, score_cols, names(score_cols))
  sc <- merge(sc, meta[, .(cell_id, ct)], by = "cell_id")

  long <- melt(sc, id.vars = c("cell_id", "ct"),
               variable.name = "signature", value.name = "score")
  summ <- long[, .(score = mean(score, na.rm = TRUE)), by = .(ct, signature)]
  fwrite(summ, "data/tme_signature_scores.csv")

  # Sequential encoding within each signature: scale per row so cell types are
  # comparable within a programme, not across programmes with different ranges.
  summ[, z := as.numeric(scale(score)), by = signature]

  p <- ggplot(summ, aes(x = ct, y = signature, fill = z)) +
    geom_tile(colour = SURFACE, linewidth = 2) +
    scale_fill_gradient2(low = DIV_LOW, mid = DIV_MID, high = DIV_HIGH,
                         midpoint = 0, name = "z (within\nsignature)") +
    labs(title = "Functional programme activity across the microenvironment",
         subtitle = sprintf("%d signatures scored; genes absent from the object were dropped",
                            length(kept)),
         x = NULL, y = NULL) +
    theme_scratch() +
    theme(panel.grid = element_blank(), axis.text.x = element_text(angle = 40, hjust = 1))
  print(save_fig(p, "tme_signature_heatmap", 10, 6))

  cat("\nSignatures scored:\n")
  knitr::kable(data.table(signature = names(kept),
                          genes_found = vapply(kept, length, integer(1)),
                          genes_in_set = vapply(SIGNATURES[names(kept)], length, integer(1))))
} else note("No signature retained enough genes to score.")
```

```{r signature-condition, fig.width = 9, fig.height = 5}
if (exists("sc") && !is.na(col_cond) && uniqueN(meta[[col_cond]]) > 1) {
  sc2 <- merge(sc, meta[, .(cell_id, cond = get(col_cond))], by = "cell_id")
  long2 <- melt(sc2, id.vars = c("cell_id", "ct", "cond"),
                variable.name = "signature", value.name = "score")
  summ2 <- long2[, .(score = mean(score, na.rm = TRUE)), by = .(cond, signature)]
  fwrite(summ2, "data/tme_signature_by_condition.csv")

  p <- ggplot(summ2, aes(x = signature, y = score, fill = cond)) +
    geom_col(position = position_dodge(width = .72), width = .64,
             colour = SURFACE, linewidth = .8) +
    scale_fill_manual(values = CAT, name = "condition") +
    geom_hline(yintercept = 0, colour = MUTED, linewidth = .4) +
    labs(title = "Programme activity by condition",
         x = NULL, y = "mean module score") +
    theme_scratch() + theme(axis.text.x = element_text(angle = 30, hjust = 1))
  print(save_fig(p, "tme_signature_by_condition"))
}
```

## Diversity

```{r diversity, fig.width = 9, fig.height = 4.5}
# Shannon entropy over cell-type proportions: how evenly mixed each sample's
# microenvironment is. A low value means one population dominates.
if (!is.na(col_smp)) {
  div <- meta[, .N, by = .(smp = get(col_smp), ct)][, p := N / sum(N), by = smp][
             , .(shannon = -sum(p * log(p)), n_types = .N, cells = sum(N)), by = smp]
  if (!is.na(col_cond)) {
    div <- merge(div, unique(meta[, .(smp = get(col_smp), cond = get(col_cond))]), by = "smp")
  }
  fwrite(div, "data/tme_diversity.csv")

  p <- ggplot(div, aes(x = reorder(smp, shannon), y = shannon)) +
    geom_col(fill = CAT[1], width = .68) +
    geom_text(aes(label = sprintf("%.2f", shannon)), hjust = -.2,
              colour = INK2, size = 3.2) +
    coord_flip() +
    scale_y_continuous(expand = expansion(mult = c(0, .14))) +
    labs(title = "Microenvironment diversity by sample",
         subtitle = "Shannon entropy over non-malignant cell-type proportions",
         x = NULL, y = "Shannon entropy") +
    theme_scratch()
  print(save_fig(p, "tme_diversity", 9, 4.5))
  knitr::kable(div, digits = 3)
}
```

## Summary

```{r summary}
produced <- list.files("data", pattern = "\\.csv$")
knitr::kable(data.table(output = produced))
```

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
