---
title: "Module - Sctype Aggregate annotation"
description: | 
  Description
execute:
  freeze: auto
  cache: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: 'Project'
  seurat_object: 'data/Project_major_annotation_object.RDS'
  cell_annotation_files: 'data/Project_T_Cells_annotation.csv;data/Project_Myeloid_annotation.csv;data/Project_Endothelial_Cells_annotation.csv;data/Project_B_Plasma_Cells_annotation.csv'
  n_threads: 8
  n_memory: 16
  work_directory: !expr here::here()
  auto_save: TRUE
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup}
#| include: false

# Project parameters 
project_name <- params$project_name
seurat_object <- params$seurat_object

# Inputs and thresholds
cell_annotation_files <- strsplit(
  params$cell_annotation_files, split = ';')[[1]]

# Optional parameters

# Dataflow/Computational parameters
n_threads <- params$n_threads
n_memory <- params$n_memory

# Output parameters
work_directory <- params$work_directory
timestamp <- params$timestamp
auto_save <- params$auto_save

# Rmarkdown options
knitr::opts_knit$set(
  root.dir = work_directory
  )

knitr::opts_chunk$set(
  fig.path = "figures/",   
  dev      = "png",
  dpi      = 300
)

```

:::{.callout-important collapse="true"}
Here we will print the parameters
:::

## Loading library

```{r}
#| output: false

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(HGNChelper)

```

## Folder structure

```{r project_directory}
#| output: false

if(!dir.exists(work_directory)) {
  dir.create(work_directory, recursive = T)
}

for(sub_directory in c('data', 'figures')) {
  dir.create(paste0(work_directory, '/', sub_directory))
}

```

## Loading Seurat object

```{r}

# Reading seurat object
seurat_object <- readRDS(file = seurat_object)
# seurat_object = project_major_annotation_object
# Setting random seed
random_seed <- 2203
set.seed(random_seed)

```

## Adding metadata

```{r}
#| output: false

# Read and combine all data files into a single data frame
# cell_annotation_files_aggregate = "../major/../../../../data-Final/SCRATCH_ANNOTATION:SCYTPE_MAJOR_ANNOTATION/data/project_major_annotation.csv"
cell_annotation_files_aggregate <- purrr::map_df(
    cell_annotation_files, read_csv
  )

```

```{r}

cell_annotation_files_aggregate <- cell_annotation_files_aggregate[
  !is.na(cell_annotation_files_aggregate$sctype),]

cell_annotation_files_aggregate <- cell_annotation_files_aggregate %>%
  tibble::column_to_rownames(var = "barcode") %>%
  dplyr::rename(
    scytpe_subset = sctype
  )

```


```{r}

seurat_object <- AddMetaData(
  seurat_object,
  cell_annotation_files_aggregate
)

seurat_object$scytpe_subset <- ifelse(
  is.na(seurat_object$scytpe_subset), seurat_object$sctype, seurat_object$scytpe_subset
)

seurat_object$predicted_labels <- seurat_object$scytpe_subset

```

## Data Visualization

```{r annotation_umap}
#| label: fig-sctype-final
#| fig-cap: 'UMAP'
#| fig-width: 15
#| fig-height: 10
#| fig-align: center
#| layout-nrow: 2

# Legend on the RIGHT, in ONE column. On top it took the upper third of the
# figure; in two columns beside it, it took roughly a third of the WIDTH, because
# these labels run to ~45 characters and two of them side by side is a very wide
# block. One column halves that and hands the space back to the embedding. 29
# entries at 8pt is about 4.5 inches tall, so it still fits the figure height.
SCP::CellDimPlot(
  srt = seurat_object,
  group.by = "scytpe_subset",
  reduction = "umap",
  label = TRUE,
  label_repel = TRUE,
  show_stat = TRUE,
  theme_use = "theme_blank"
) +
  theme(
    legend.position  = "right",
    legend.direction = "vertical",
    legend.justification = "top",
    legend.text      = element_text(size = 8),
    legend.key.size  = unit(8, "pt")
  ) +
  guides(colour = guide_legend(ncol = 1, override.aes = list(size = 2.5)),
         fill   = guide_legend(ncol = 1, override.aes = list(size = 2.5)))

```

:::{.callout-tip}
The @fig-sctype-final displays the final annotation after the hierarchical approach. Major cell types are shown when not further informantion is available on the cell annotation database.
:::

## scType dendrogram

```{r annotation_tree}
#| warning: false

# # Aggregate data by annotations
# cell_average <- AverageExpression(
#   seurat_object,
#   assay = 'RNA',
#   slot = 'scale.data',
#   return.seurat = FALSE,
#   group.by = "scytpe_subset")
# Aggregate data by annotations (Seurat 5 syntax)
cell_average <- AverageExpression(
  seurat_object,
  assay        = 'RNA',
  layer        = 'scale.data',    # use `layer`
  return.seurat = FALSE,
  group.by     = "scytpe_subset")


cell_average <- cell_average$RNA

# Compute distance matrix on aggregated data
cell_dist_matrix <- as.matrix(cell_average)
cell_dist_matrix <- stats::cor(cell_dist_matrix, method = "pearson")

# Convert correlation matrix to a distance matrix
cell_dist_matrix <- stats::as.dist(1 - cell_dist_matrix)

# Perform hierarchical clustering
cell_hierarchical <- stats::hclust(cell_dist_matrix)

```

```{r}
#| label: fig-dendrogram-final
#| fig-cap: 'Dendrogram'
#| fig-width: 10
#| fig-height: 12
#| fig-align: center

# Printing dendrogram

plot(cell_hierarchical, main = "Dendrogram", xlab = "Index", ylab = "Height")

```

:::{.callout-tip}
[Beta] The @fig-dendrogram-final displays the hierarchical clustering across the cell types based on average normalized expression from `scale.data`. The average expression per cell population is correlated based on Pearson. Therefore, correlated cells should display closet distances, i.e., clustering into near branches.
:::

## Cell composition

```{r barplot_composition}
#| label: fig-bar-composition
#| fig-cap: 'Cluster composition of each annotated population'
# Flipped bars need HEIGHT (one row per label), not width. 29 labels at ~0.45in.
#| fig-width: 12
#| fig-height: 15
#| fig-align: center

# Composition of each annotated population by cluster.
#
# `scytpe_subset` labels run to ~45 characters ("CD4_and_CD8_T_cells@Interferon_
# Response") and there are 29 of them. Rotated 45 degrees on the x axis they
# collided into a solid wedge of text taller than the plot itself. Flipping the
# bars puts every label on its own horizontal line, where label length costs
# width instead of stacking into an unreadable wedge.
SCP::CellStatPlot(
  seurat_object,
  stat.by = "seurat_clusters",
  group.by = "scytpe_subset",
  label = TRUE
) +
  coord_flip() +
  theme(
    axis.text.y = element_text(size = 8, hjust = 1),
    axis.text.x = element_text(size = 9, angle = 0, hjust = .5),
    legend.position = "right"
  )

```

:::{.callout-tip}
The @fig-bar-composition provides an overall view of cell composition across distinct clusters on the dataset.
:::

## Saving Seurat object

```{r object_dump}
#| cache: false

if(auto_save) {

  saveRDS(seurat_object, file = paste0("data/", project_name, "_sctype_agg_annotation_object.RDS"))

}

```

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

params

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```
