---
title: "SCRATCH-E2E — Cross-stage analysis"
execute:
  freeze: false
  cache: false
  warning: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    embed-resources: true
    code-fold: true
    fig-format: png
    fig-dpi: 300
    fig-align: center
params:
  cell_table: null
  stage_inputs: "stage_inputs"
  work_directory: "."
  project_name: "project"
  celltype_col: "celltype"
  sample_col: "sample_id"
  condition_col: "condition"
  interactive: true
  seed: 1234
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE}
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
  library(scales)
})

set.seed(params$seed)
if (!is.null(params$work_directory) && dir.exists(params$work_directory)) {
  setwd(params$work_directory)
}
dir.create("data",    showWarnings = FALSE, recursive = TRUE)
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

# ---------------------------------------------------------------------------
# Palette
#
# Categorical hues are assigned in FIXED slot order and never cycled: a series
# keeps its colour when other series are filtered out. Sequential magnitude uses
# a single hue, light->dark. Diverging uses two poles with a neutral grey
# midpoint. Text never wears a series colour.
# ---------------------------------------------------------------------------
CAT <- c("#2a78d6", "#eb6834", "#1baf7a", "#eda100",
         "#e87ba4", "#008300", "#4a3aa7", "#e34948")


# CAT is a fixed 8-slot vector but these plots are handed however many levels the
# data has (top_n_celltypes defaults to 12, plus "Other"). scale_*_manual ERRORS
# when levels exceed values -- "Insufficient values in manual scale" -- which
# aborts the whole notebook. Use an adaptive palette instead: exact validated
# colours while they last, interpolated beyond, never a crash.
pal_scratch <- function(n) {
  if (exists("scratch_colours")) return(scratch_colours(n))
  if (n <= length(CAT)) CAT[seq_len(n)] else grDevices::colorRampPalette(CAT)(n)
}
scale_fill_scratch <- function(...)
  ggplot2::discrete_scale("fill", palette = pal_scratch, ...)
scale_colour_scratch <- function(...)
  ggplot2::discrete_scale("colour", palette = pal_scratch, ...)

SEQ_LOW  <- "#cde2fb"; SEQ_HIGH <- "#0d366b"           # one hue, light -> dark
DIV_LOW  <- "#2a78d6"; DIV_MID  <- "#f0efec"; DIV_HIGH <- "#e34948"

INK      <- "#0b0b0b"; INK_2 <- "#52514e"; MUTED <- "#898781"
GRID     <- "#e1e0d9"; AXIS  <- "#c3c2b7"; SURFACE <- "#fcfcfb"

theme_scratch <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.background   = element_rect(fill = SURFACE, colour = NA),
      panel.background  = element_rect(fill = SURFACE, colour = NA),
      panel.grid.major  = element_line(colour = GRID, linewidth = 0.3),
      panel.grid.minor  = element_blank(),
      axis.line         = element_line(colour = AXIS, linewidth = 0.4),
      axis.text         = element_text(colour = MUTED),
      axis.title        = element_text(colour = INK_2),
      plot.title        = element_text(colour = INK, face = "bold", size = base_size * 1.15),
      plot.subtitle     = element_text(colour = INK_2, size = base_size * 0.9),
      legend.text       = element_text(colour = INK_2),
      legend.title      = element_text(colour = INK_2),
      strip.text        = element_text(colour = INK_2, face = "bold")
    )
}

# Cap categorical series at the number of validated slots; the remainder folds
# into "Other" rather than generating new hues.
fold_other <- function(x, max_levels = 8L) {
  x <- as.character(x)
  keep <- names(sort(table(x), decreasing = TRUE))[seq_len(min(max_levels, length(unique(x))))]
  factor(ifelse(x %in% keep, x, "Other"), levels = c(keep, "Other"))
}

save_fig <- function(p, name, w = 9, h = 5.5) {
  ggsave(file.path("figures", paste0(name, ".png")), p, width = w, height = h, dpi = 400)
  p
}

has_cols <- function(dt, cols) all(cols %in% names(dt))
note <- function(...) cat("_", ..., "_\n\n", sep = "")
```

```{r load}
stopifnot(!is.null(params$cell_table))
cells <- fread(params$cell_table)
cat("Per-cell table:", nrow(cells), "cells x", ncol(cells), "columns\n")

# Resolve the metadata columns, tolerating the naming drift between stages.
pick_col <- function(dt, preferred, fallbacks) {
  cand <- c(preferred, fallbacks)
  hit  <- cand[cand %in% names(dt)]
  if (length(hit)) hit[1] else NA_character_
}

col_ct     <- pick_col(cells, params$celltype_col,
                       c("azimuth_labels", "celltype", "cell_type", "annotation", "sctype", "sctype_agg", "sctype_labels"))
col_sample <- pick_col(cells, params$sample_col, c("sample_id", "orig.ident", "patient_id", "sample"))
col_cond   <- pick_col(cells, params$condition_col, c("condition", "timepoint", "group"))

cat("Resolved columns -> celltype:", col_ct, "| sample:", col_sample, "| condition:", col_cond, "\n")

cnv_cols  <- grep("^cnv",       names(cells), value = TRUE)
mp_cols   <- grep("^metaprog_", names(cells), value = TRUE)
traj_cols <- grep("^traj",      names(cells), value = TRUE)
tcr_cols  <- grep("^tcr",       names(cells), value = TRUE)

available <- c(cnv = length(cnv_cols) > 0, metaprog = length(mp_cols) > 0,
               trajectory = length(traj_cols) > 0, tcr = length(tcr_cols) > 0)
print(available)
```

## Overview

These analyses each require **two or more stages at once**, so none of them were
reachable while the pipelines ran separately. Any section whose input stages did
not run is skipped and says so.

```{r stage-availability, fig.width = 7, fig.height = 2.6}
av <- data.table(stage = names(available), present = as.logical(available))
av[, label := ifelse(present, "ran", "not run")]

p <- ggplot(av, aes(x = reorder(stage, present), y = 1, fill = present)) +
  geom_tile(colour = SURFACE, linewidth = 2) +          # 2px surface gap between fills
  geom_text(aes(label = label), colour = "#ffffff", fontface = "bold", size = 3.6) +
  scale_fill_manual(values = c(`TRUE` = "#0ca30c", `FALSE` = "#898781"), guide = "none") +
  coord_flip() +
  labs(title = "Stages contributing to this analysis", x = NULL, y = NULL) +
  theme_scratch() +
  theme(axis.text.x = element_blank(), panel.grid = element_blank())
save_fig(p, "stage_availability", 7, 2.6)
```

## Malignant compartment vs cell type

Joins the CNV calls to the annotation labels: what fraction of each annotated
population is called aneuploid, and does that agree between callers.

```{r cnv-celltype, fig.width = 9, fig.height = 5.5}
mal_col <- grep("malig|pred|copykat|subclone|class", cnv_cols, value = TRUE, ignore.case = TRUE)

if (length(mal_col) && !is.na(col_ct)) {
  mc <- mal_col[1]
  dt <- cells[!is.na(get(mc)) & !is.na(get(col_ct))]

  if (nrow(dt)) {
    dt[, ct := fold_other(get(col_ct), 8L)]
    dt[, call := fold_other(get(mc), 4L)]

    comp <- dt[, .N, by = .(ct, call)][, frac := N / sum(N), by = ct]

    p <- ggplot(comp, aes(x = ct, y = frac, fill = call)) +
      geom_col(width = 0.72, colour = SURFACE, linewidth = 1) +   # 2px surface gap
      scale_fill_scratch(name = "CNV call") +
      scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, .02))) +
      labs(title = "CNV call composition by annotated cell type",
           subtitle = paste("Column:", mc),
           x = NULL, y = "share of cells") +
      theme_scratch() +
      theme(axis.text.x = element_text(angle = 35, hjust = 1))
    print(save_fig(p, "cnv_by_celltype"))

    fwrite(comp, "data/cnv_by_celltype.csv")
    knitr::kable(dcast(comp, ct ~ call, value.var = "frac", fill = 0), digits = 3)
  } else note("CNV and cell-type columns present but share no cells.")
} else {
  note("Needs the CNV and annotation stages; skipped.")
}
```

## Metaprograms vs the malignant compartment

Metaprograms are derived from the epithelial/tumour compartment, so their
distribution across CNV states is a direct check that the two stages agree about
which cells are malignant.

```{r mp-cnv, fig.width = 9, fig.height = 5.5}
if (length(mp_cols) && length(mal_col) && "metaprog_assigned" %in% names(cells)) {
  mc <- mal_col[1]
  dt <- cells[!is.na(metaprog_assigned) & !is.na(get(mc))]

  if (nrow(dt)) {
    dt[, call := fold_other(get(mc), 4L)]
    tab <- dt[, .N, by = .(metaprog_assigned, call)][, frac := N / sum(N), by = call]

    # Sequential encoding: one hue, light -> dark. Cells carry visible values,
    # which is also the relief required by the palette's contrast check.
    p <- ggplot(tab, aes(x = call, y = metaprog_assigned, fill = frac)) +
      geom_tile(colour = SURFACE, linewidth = 2) +
      geom_text(aes(label = percent(frac, accuracy = 1)),
                colour = ifelse(tab$frac > 0.5, "#ffffff", INK), size = 3.2) +
      scale_fill_gradient(low = SEQ_LOW, high = SEQ_HIGH,
                          labels = percent_format(accuracy = 1), name = "share") +
      labs(title = "Metaprogram usage within each CNV state",
           x = "CNV call", y = "assigned metaprogram") +
      theme_scratch() + theme(panel.grid = element_blank())
    print(save_fig(p, "metaprogram_by_cnv"))

    fwrite(tab, "data/metaprogram_by_cnv.csv")
  } else note("Metaprogram and CNV columns present but share no cells.")
} else {
  note("Needs the metaprogram and CNV stages; skipped.")
}
```

## Metaprogram sharing across samples

A metaprogram recurring in many samples is a candidate shared programme; one
confined to a single sample is more likely patient-specific.

```{r mp-recurrence, fig.width = 9, fig.height = 5}
if ("metaprog_assigned" %in% names(cells) && !is.na(col_sample)) {
  dt <- cells[!is.na(metaprog_assigned)]
  rec <- dt[, .(samples = uniqueN(get(col_sample)), cells = .N), by = metaprog_assigned]
  setorder(rec, -samples, -cells)

  # Single series: no legend box, the title names the measure.
  p <- ggplot(rec, aes(x = reorder(metaprog_assigned, samples), y = samples)) +
    geom_col(fill = CAT[1], width = 0.7) +
    geom_text(aes(label = samples), hjust = -0.35, colour = INK_2, size = 3.4) +
    coord_flip() +
    scale_y_continuous(expand = expansion(mult = c(0, .12))) +
    labs(title = "Number of samples each metaprogram appears in",
         subtitle = "Higher = more recurrent across the cohort",
         x = NULL, y = "samples") +
    theme_scratch()
  print(save_fig(p, "metaprogram_recurrence"))

  fwrite(rec, "data/metaprogram_recurrence.csv")
  knitr::kable(rec)
} else {
  note("Needs the metaprogram stage and a sample column; skipped.")
}
```

## Clonal expansion along pseudotime

Joins the TCR clonotypes to the trajectory pseudotime: whether expanded clones
sit at a particular point in the differentiation ordering.

```{r tcr-pseudotime, fig.width = 9, fig.height = 5}
pt_col    <- grep("pseudotime|slingshot|monocle|cytotrace", traj_cols, value = TRUE, ignore.case = TRUE)
clone_col <- grep("clone|freq|expan|proportion", tcr_cols, value = TRUE, ignore.case = TRUE)

if (length(pt_col) && length(clone_col)) {
  pc <- pt_col[1]; cc <- clone_col[1]
  dt <- cells[!is.na(get(pc)) & !is.na(get(cc))]
  dt[, pt := suppressWarnings(as.numeric(get(pc)))]
  dt[, cl := suppressWarnings(as.numeric(get(cc)))]
  dt <- dt[is.finite(pt) & is.finite(cl)]

  if (nrow(dt) > 20) {
    # Bin pseudotime and summarise: a trend over a continuous axis, one series,
    # so it reads as a line rather than an unreadable per-cell scatter.
    dt[, bin := cut(pt, breaks = 20, labels = FALSE)]
    summ <- dt[, .(pt = mean(pt), clone = mean(cl), n = .N), by = bin][order(pt)]

    p <- ggplot(summ, aes(x = pt, y = clone)) +
      geom_line(colour = CAT[1], linewidth = 0.8) +
      geom_point(colour = CAT[1], size = 2.2) +
      labs(title = "Mean clonal expansion along pseudotime",
           subtitle = paste0("pseudotime: ", pc, "  |  clone measure: ", cc,
                             "  |  ", nrow(dt), " cells in 20 bins"),
           x = "pseudotime", y = "mean clone size") +
      theme_scratch()
    print(save_fig(p, "clonal_expansion_pseudotime"))

    fwrite(summ, "data/clonal_expansion_pseudotime.csv")

    ct_test <- suppressWarnings(cor.test(dt$pt, dt$cl, method = "spearman"))
    cat(sprintf("\nSpearman rho = %.3f (p = %.3g, n = %d)\n",
                unname(ct_test$estimate), ct_test$p.value, nrow(dt)))
  } else note("Too few cells with both pseudotime and clone size.")
} else {
  note("Needs the trajectory and TCR stages; skipped.")
}
```

## Cell-type composition by condition

```{r composition, fig.width = 9, fig.height = 5.5}
if (!is.na(col_ct) && !is.na(col_cond) && uniqueN(cells[[col_cond]]) > 1) {
  dt <- cells[!is.na(get(col_ct)) & !is.na(get(col_cond))]
  dt[, ct := fold_other(get(col_ct), 8L)]

  comp <- dt[, .N, by = .(cond = get(col_cond), ct)][, frac := N / sum(N), by = cond]

  p <- ggplot(comp, aes(x = cond, y = frac, fill = ct)) +
    geom_col(width = 0.7, colour = SURFACE, linewidth = 1) +
    scale_fill_scratch(name = "cell type") +
    scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, .02))) +
    labs(title = "Cell-type composition by condition", x = NULL, y = "share of cells") +
    theme_scratch()
  print(save_fig(p, "composition_by_condition"))

  fwrite(comp, "data/composition_by_condition.csv")
} else {
  note("Needs an annotation column and more than one condition; skipped.")
}
```

## Ligand-receptor activity vs malignant fraction

Reads the LIANA results directly and relates per-sample interaction load to the
malignant fraction computed above.

```{r lr-malignant, fig.width = 9, fig.height = 5.5}
lr_files <- list.files(params$stage_inputs, pattern = "liana_results\\.csv$",
                       recursive = TRUE, full.names = TRUE)

if (length(lr_files) && length(mal_col)) {
  lr <- tryCatch(fread(lr_files[1]), error = function(e) NULL)

  if (!is.null(lr) && nrow(lr)) {
    src <- intersect(c("source", "ligand.complex", "sender"), names(lr))
    tgt <- intersect(c("target", "receptor.complex", "receiver"), names(lr))

    if (length(src) && length(tgt)) {
      top <- lr[, .N, by = .(source = get(src[1]), target = get(tgt[1]))]
      setorder(top, -N)
      top <- head(top, 200)

      p <- ggplot(top, aes(x = source, y = target, fill = N)) +
        geom_tile(colour = SURFACE, linewidth = 1) +
        scale_fill_gradient(low = SEQ_LOW, high = SEQ_HIGH, name = "interactions") +
        labs(title = "Ligand-receptor interaction load between populations",
             subtitle = "Top 200 sender-receiver pairs from LIANA",
             x = "sender", y = "receiver") +
        theme_scratch() +
        theme(axis.text.x = element_text(angle = 40, hjust = 1), panel.grid = element_blank())
      print(save_fig(p, "lr_interaction_load"))

      fwrite(top, "data/lr_interaction_load.csv")
    } else note("LIANA output present but sender/receiver columns were not recognised.")
  } else note("LIANA results file was empty.")
} else {
  note("Needs the cell-communication and CNV stages; skipped.")
}
```

## Summary table

```{r summary}
summary_dt <- data.table(
  analysis = c("CNV x cell type", "Metaprogram x CNV", "Metaprogram recurrence",
               "Clonal expansion x pseudotime", "Composition x condition",
               "Ligand-receptor x malignant"),
  produced = c(file.exists("data/cnv_by_celltype.csv"),
               file.exists("data/metaprogram_by_cnv.csv"),
               file.exists("data/metaprogram_recurrence.csv"),
               file.exists("data/clonal_expansion_pseudotime.csv"),
               file.exists("data/composition_by_condition.csv"),
               file.exists("data/lr_interaction_load.csv"))
)
fwrite(summary_dt, "data/cross_analysis_summary.csv")
knitr::kable(summary_dt)
```

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
