---
title: "SCRATCH-E2E — Master object integration"
execute:
  freeze: false
  cache: false
  warning: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    embed-resources: true
    code-fold: true
    fig-format: png
    fig-dpi: 300
    fig-align: center
params:
  seurat_object: null
  project_name: "project"
  stage_inputs: "stage_inputs"
  work_directory: "."
  celltype_col: "celltype"
  sample_col: "sample_id"
  condition_col: "condition"
  seed: 1234
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE}
suppressPackageStartupMessages({
  library(Seurat)
  library(Matrix)
  library(data.table)
  library(ggplot2)
})

set.seed(params$seed)
if (!is.null(params$work_directory) && dir.exists(params$work_directory)) {
  setwd(params$work_directory)
}

dir.create("data",    showWarnings = FALSE, recursive = TRUE)
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

stage_root <- params$stage_inputs
```

## What this step does

Each analysis stage of SCRATCH-E2E writes its own per-cell result: copy-number
calls, metaprogram scores, pseudotime, TCR clonotypes. Individually they are
tables keyed by cell barcode. This step joins all of them back onto the object
the downstream stages were run from, producing:

* `<project>_master_object.RDS` — one Seurat object carrying every stage's
  per-cell annotation in its metadata
* `<project>_cell_table.csv` — the same metadata as a tidy table, which is what
  the cross-stage analysis and the master report read

Joining once here means the cross-stage analyses never have to re-open the
large object, and any stage that did not run simply contributes no columns.

```{r load-object}
stopifnot(!is.null(params$seurat_object))
obj <- readRDS(params$seurat_object)

cat("Master object:", ncol(obj), "cells x", nrow(obj), "features\n")

meta <- as.data.table(obj@meta.data, keep.rownames = "cell_id")
setkey(meta, cell_id)
cat("Starting metadata columns:", ncol(meta) - 1L, "\n")
```

```{r helpers}
# Discover a stage's contributed files without assuming they exist. A stage that
# was skipped leaves an empty (or absent) directory and is simply not joined.
stage_files <- function(stage, pattern = "\\.(csv|tsv|txt)$") {
  d <- file.path(stage_root, stage)
  if (!dir.exists(d)) return(character(0))
  list.files(d, pattern = pattern, recursive = TRUE, full.names = TRUE)
}

# Read a per-cell table and normalise its barcode column to `cell_id`.
read_cell_table <- function(path) {
  dt <- tryCatch(fread(path), error = function(e) NULL)
  if (is.null(dt) || !nrow(dt)) return(NULL)

  cand <- intersect(c("cell_id", "cell", "barcode", "cellname", "Cell", "Barcode", "V1"),
                    names(dt))
  if (!length(cand)) {
    # Fall back to the first character column that looks like a barcode set.
    chr <- names(dt)[vapply(dt, is.character, logical(1))]
    if (!length(chr)) return(NULL)
    cand <- chr[1]
  }
  setnames(dt, cand[1], "cell_id")
  dt[, cell_id := as.character(cell_id)]
  unique(dt, by = "cell_id")
}

# Left-join a stage table onto the master metadata, prefixing its columns so two
# stages can never silently overwrite each other's annotation.
join_stage <- function(meta, dt, prefix) {
  if (is.null(dt)) return(meta)
  keep <- setdiff(names(dt), "cell_id")
  if (!length(keep)) return(meta)
  setnames(dt, keep, paste0(prefix, "_", keep))
  merge(meta, dt, by = "cell_id", all.x = TRUE, sort = FALSE)
}

joined <- character(0)
```

## Copy number

```{r cnv}
for (stage in c("cnv_infercnv", "cnv_scevan")) {
  fs <- stage_files(stage)
  if (!length(fs)) { cat("no output from", stage, "\n"); next }

  # Prefer a file that plainly holds per-cell calls.
  pick <- fs[grepl("cell|prediction|subclone|malig", basename(fs), ignore.case = TRUE)]
  if (!length(pick)) pick <- fs
  dt <- read_cell_table(pick[1])

  if (!is.null(dt)) {
    meta <- join_stage(meta, dt, sub("cnv_", "cnv", stage))
    joined <- c(joined, stage)
    cat("joined", stage, "from", basename(pick[1]), "\n")
  }
}
```

## Metaprograms

```{r metaprog}
fs <- stage_files("metaprog")
mp_files <- fs[grepl("MP_list_final|metaprogram|mp_score", basename(fs), ignore.case = TRUE)]

if (length(mp_files)) {
  # MP_list_final.csv is programme x gene, not per-cell: score cells against
  # each metaprogram so the result is joinable and directly comparable.
  mp <- tryCatch(fread(mp_files[1]), error = function(e) NULL)

  if (!is.null(mp) && ncol(mp) > 1L) {
    mp_sets <- lapply(names(mp)[-1], function(cn) {
      g <- toupper(as.character(mp[[cn]]))
      intersect(unique(g[nzchar(g) & !is.na(g)]), toupper(rownames(obj)))
    })
    names(mp_sets) <- names(mp)[-1]
    mp_sets <- mp_sets[vapply(mp_sets, length, integer(1)) >= 5L]

    if (length(mp_sets)) {
      # Map back to the object's actual gene casing before scoring.
      lookup <- setNames(rownames(obj), toupper(rownames(obj)))
      mp_sets <- lapply(mp_sets, function(g) unname(lookup[g]))

      # See tme_characterisation.qmd: Seurat::AddModuleScore is defunct-dependent
      # under SeuratObject 5.x. scratch_module_score implements the same method.
      obj <- scratch_module_score(obj, mp_sets, name = "MPscore", seed = params$seed)
      score_cols <- grep("^MPscore\\d+$", colnames(obj@meta.data), value = TRUE)
      names(score_cols) <- names(mp_sets)[seq_along(score_cols)]

      sc <- as.data.table(obj@meta.data[, score_cols, drop = FALSE], keep.rownames = "cell_id")
      setnames(sc, score_cols, paste0("metaprog_", names(score_cols)))

      # Assign each cell to its highest-scoring metaprogram.
      mat <- as.matrix(sc[, -1, with = FALSE])
      sc[, metaprog_assigned := colnames(mat)[max.col(mat, ties.method = "first")]]
      sc[, metaprog_assigned := sub("^metaprog_", "", metaprog_assigned)]

      meta <- merge(meta, sc, by = "cell_id", all.x = TRUE, sort = FALSE)
      joined <- c(joined, "metaprog")
      cat("scored", length(mp_sets), "metaprograms\n")
    }
  }
} else {
  cat("no metaprogram output\n")
}
```

## Trajectory

```{r trajectory}
fs <- stage_files("trajectory")
pt <- fs[grepl("pseudotime|concordance|slingshot|monocle|cytotrace",
               basename(fs), ignore.case = TRUE)]
if (length(pt)) {
  meta <- join_stage(meta, read_cell_table(pt[1]), "traj")
  joined <- c(joined, "trajectory")
  cat("joined trajectory from", basename(pt[1]), "\n")
} else {
  cat("no trajectory output\n")
}
```

## TCR

```{r tcr}
fs <- stage_files("tcr")
tc <- fs[grepl("cell|clone|clonotype", basename(fs), ignore.case = TRUE)]
if (length(tc)) {
  meta <- join_stage(meta, read_cell_table(tc[1]), "tcr")
  joined <- c(joined, "tcr")
  cat("joined TCR from", basename(tc[1]), "\n")
} else {
  cat("no TCR output\n")
}
```

## Assemble and save

```{r assemble}
setDF(meta)
rownames(meta) <- meta$cell_id

# Only add columns the object does not already carry.
new_cols <- setdiff(colnames(meta), c("cell_id", colnames(obj@meta.data)))
if (length(new_cols)) {
  obj <- AddMetaData(obj, meta[colnames(obj), new_cols, drop = FALSE])
}

cat("Stages joined:", if (length(joined)) paste(joined, collapse = ", ") else "none", "\n")
cat("New per-cell columns added:", length(new_cols), "\n")

saveRDS(obj, file.path("data", paste0(params$project_name, "_master_object.RDS")))
fwrite(meta, file.path("data", paste0(params$project_name, "_cell_table.csv")))

# A compact record of which stages contributed, for the master report.
fwrite(data.table(stage = c("cnv_infercnv", "cnv_scevan", "metaprog", "trajectory", "tcr"),
                  joined = c("cnv_infercnv", "cnv_scevan", "metaprog", "trajectory", "tcr") %in% joined),
       file.path("data", "stage_join_status.csv"))
```

```{r overview-plot, fig.width = 9, fig.height = 4}
avail <- intersect(c(params$celltype_col, "metaprog_assigned"), colnames(obj@meta.data))
if (length(avail) && "umap" %in% Reductions(obj)) {
  for (col in avail) {
    print(DimPlot(obj, group.by = col, reduction = "umap", label = TRUE, repel = TRUE) +
            ggtitle(col) + theme(legend.position = "right"))
    ggsave(file.path("figures", paste0("umap_", col, ".png")), width = 9, height = 6, dpi = 400)
  }
} else {
  cat("No UMAP reduction on the object; skipping overview plots.\n")
}
```

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
