---
title: "SCRATCH-E2E — Master object integration"
execute:
  freeze: false
  cache: false
  warning: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    embed-resources: true
    code-fold: true
    fig-format: png
    fig-dpi: 300
    fig-align: center
params:
  seurat_object: null
  project_name: "project"
  stage_inputs: "stage_inputs"
  work_directory: "."
  celltype_col: "celltype"
  sample_col: "sample_id"
  condition_col: "condition"
  seed: 1234
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE}
suppressPackageStartupMessages({
  library(Seurat)
  library(Matrix)
  library(data.table)
  library(ggplot2)
})

set.seed(params$seed)
if (!is.null(params$work_directory) && dir.exists(params$work_directory)) {
  setwd(params$work_directory)
}

dir.create("data",    showWarnings = FALSE, recursive = TRUE)
dir.create("figures", showWarnings = FALSE, recursive = TRUE)

stage_root <- params$stage_inputs
```

## What this step does

Each analysis stage of SCRATCH-E2E writes its own per-cell result: copy-number
calls, metaprogram scores, pseudotime, TCR clonotypes. Individually they are
tables keyed by cell barcode. This step joins all of them back onto the object
the downstream stages were run from, producing:

* `<project>_master_object.RDS` — one Seurat object carrying every stage's
  per-cell annotation in its metadata
* `<project>_cell_table.csv` — the same metadata as a tidy table, which is what
  the cross-stage analysis and the master report read

Joining once here means the cross-stage analyses never have to re-open the
large object, and any stage that did not run simply contributes no columns.

```{r load-object}
stopifnot(!is.null(params$seurat_object))
obj <- readRDS(params$seurat_object)

cat("Master object:", ncol(obj), "cells x", nrow(obj), "features\n")

meta <- as.data.table(obj@meta.data, keep.rownames = "cell_id")
setkey(meta, cell_id)
cat("Starting metadata columns:", ncol(meta) - 1L, "\n")
```

```{r helpers}
# Discover a stage's contributed files without assuming they exist. A stage that
# was skipped leaves an empty (or absent) directory and is simply not joined.
stage_files <- function(stage, pattern = "\\.(csv|tsv|txt)$") {
  d <- file.path(stage_root, stage)
  if (!dir.exists(d)) return(character(0))
  list.files(d, pattern = pattern, recursive = TRUE, full.names = TRUE)
}

# Read a per-cell table and normalise its barcode column to `cell_id`.
read_cell_table <- function(path) {
  dt <- tryCatch(fread(path), error = function(e) NULL)
  if (is.null(dt) || !nrow(dt)) return(NULL)

  cand <- intersect(c("cell_id", "cell", "barcode", "cellname", "Cell", "Barcode", "V1"),
                    names(dt))
  if (!length(cand)) {
    # Fall back to the first character column that looks like a barcode set.
    chr <- names(dt)[vapply(dt, is.character, logical(1))]
    if (!length(chr)) return(NULL)
    cand <- chr[1]
  }
  setnames(dt, cand[1], "cell_id")
  dt[, cell_id := as.character(cell_id)]
  unique(dt, by = "cell_id")
}

# Every join is recorded here with the number of cells it actually matched.
# `joined` used to be appended the moment a file parsed, so a stage counted as
# joined when the merge matched ZERO cells -- which is exactly what happened to
# cnv_infercnv: it "joined" a heatmap-threshold file and the status table
# reported TRUE while the master object gained no CNV column at all.
join_log <- data.table(stage = character(), file = character(),
                       matched = integer(), total = integer(), columns = integer())

record_join <- function(stage, file, matched, total, columns) {
  join_log <<- rbind(join_log, data.table(stage = stage, file = file,
                                          matched = as.integer(matched),
                                          total = as.integer(total),
                                          columns = as.integer(columns)))
}

# Left-join a stage table onto the master metadata, prefixing its columns so two
# stages can never silently overwrite each other's annotation. Returns the merged
# table and reports how many of the master's cells the stage table actually hit.
join_stage <- function(meta, dt, prefix, stage, file = NA_character_) {
  if (is.null(dt)) { record_join(stage, file, 0L, nrow(meta), 0L); return(meta) }
  keep <- setdiff(names(dt), "cell_id")
  if (!length(keep)) { record_join(stage, file, 0L, nrow(meta), 0L); return(meta) }

  matched <- sum(meta$cell_id %in% dt$cell_id)
  # A zero-match join is not a join. Leave the columns off entirely rather than
  # adding a block of NAs that later chunks would treat as real data.
  if (matched == 0L) { record_join(stage, file, 0L, nrow(meta), 0L); return(meta) }

  setnames(dt, keep, paste0(prefix, "_", keep))
  record_join(stage, file, matched, nrow(meta), length(keep))
  merge(meta, dt, by = "cell_id", all.x = TRUE, sort = FALSE)
}

# A stage counts as joined only if it matched at least one cell.
joined_stages <- function() unique(join_log[matched > 0L, stage])
```

## Copy number

```{r cnv}
for (stage in c("cnv_infercnv")) {
  fs <- stage_files(stage)
  if (!length(fs)) { cat("no output from", stage, "\n"); next }

  # inferCNV's run directory holds ~30 .txt files and only ONE of them is a
  # per-cell table: map_metadata_from_infercnv.txt (barcode x has_cnv_chr*,
  # proportion_*). The rest are heatmap thresholds, gene orderings and group
  # summaries. The previous selector matched "cell|prediction|subclone|malig"
  # against the basename, matched nothing, fell back to "take fs[1]", and picked
  # infercnv.17_HMM_...heatmap_thresholds.txt purely because it sorts first.
  # Named files first, in preference order; only then a pattern.
  preferred <- c("map_metadata_from_infercnv", "cell_groupings", "predictions")
  pick <- character(0)
  for (nm in preferred) {
    hit <- fs[grepl(nm, basename(fs), ignore.case = TRUE)]
    if (length(hit)) { pick <- hit; break }
  }
  if (!length(pick))
    pick <- fs[grepl("cell|prediction|subclone|malig", basename(fs), ignore.case = TRUE)]
  if (!length(pick)) {
    cat("no per-cell table among", length(fs), "files from", stage, "\n")
    record_join(stage, NA_character_, 0L, nrow(meta), 0L)
    next
  }

  dt <- read_cell_table(pick[1])
  meta <- join_stage(meta, dt, sub("cnv_", "cnv", stage), stage, basename(pick[1]))
  cat(stage, "->", basename(pick[1]), "\n")
}
```

## Cell-type annotation

```{r annotation}
# The base object is the AZIMUTH output, so it carries azimuth_labels and nothing
# from scType -- this pipeline's own annotator, six notebooks' worth. The
# concordance table is per-cell and holds every annotator side by side, so the
# master object stops being Azimuth-only.
fs <- stage_files("annotation")
ann <- fs[grepl("annotation_concordance", basename(fs), ignore.case = TRUE)]
if (!length(ann)) ann <- fs[grepl("annotation\\.csv$", basename(fs), ignore.case = TRUE)]

if (length(ann)) {
  meta <- join_stage(meta, read_cell_table(ann[1]), "ann", "annotation", basename(ann[1]))
  cat("annotation ->", basename(ann[1]), "\n")
} else {
  cat("no annotation table\n")
  record_join("annotation", NA_character_, 0L, nrow(meta), 0L)
}
```

## Compartment (malignant / non-malignant)

```{r compartment}
# STRATIFY's assignment was staged into stage_inputs/compartment on every run and
# this notebook had no code path that read it, so `malignant_status` -- the
# central result of the pipeline -- never reached the master object or the cell
# table, and `stratify` was not even a row in the status file.
fs <- stage_files("compartment")
asg <- fs[grepl("malignancy_assignment", basename(fs), ignore.case = TRUE)]

if (length(asg)) {
  dt <- read_cell_table(asg[1])
  # Kept UNPREFIXED: downstream notebooks, the cross-analysis and the stage
  # reports all look for `malignant_status` by that exact name.
  if (!is.null(dt)) {
    matched <- sum(meta$cell_id %in% dt$cell_id)
    if (matched > 0L) {
      meta <- merge(meta, dt, by = "cell_id", all.x = TRUE, sort = FALSE)
      record_join("stratify", basename(asg[1]), matched, nrow(meta),
                  length(setdiff(names(dt), "cell_id")))
    } else {
      record_join("stratify", basename(asg[1]), 0L, nrow(meta), 0L)
    }
  }
  cat("stratify ->", basename(asg[1]), "\n")
} else {
  cat("no compartment assignment\n")
  record_join("stratify", NA_character_, 0L, nrow(meta), 0L)
}
```

## Metaprograms

```{r metaprog}
fs <- stage_files("metaprog")
mp_files <- fs[grepl("MP_list_final|metaprogram|mp_score", basename(fs), ignore.case = TRUE)]

if (length(mp_files)) {
  # MP_list_final.csv is programme x gene, not per-cell: score cells against
  # each metaprogram so the result is joinable and directly comparable.
  mp <- tryCatch(fread(mp_files[1]), error = function(e) NULL)

  if (!is.null(mp) && ncol(mp) > 1L) {
    mp_sets <- lapply(names(mp)[-1], function(cn) {
      g <- toupper(as.character(mp[[cn]]))
      intersect(unique(g[nzchar(g) & !is.na(g)]), toupper(rownames(obj)))
    })
    names(mp_sets) <- names(mp)[-1]
    mp_sets <- mp_sets[vapply(mp_sets, length, integer(1)) >= 5L]

    if (length(mp_sets)) {
      # Map back to the object's actual gene casing before scoring.
      lookup <- setNames(rownames(obj), toupper(rownames(obj)))
      mp_sets <- lapply(mp_sets, function(g) unname(lookup[g]))

      # See tme_characterisation.qmd: Seurat::AddModuleScore is defunct-dependent
      # under SeuratObject 5.x. scratch_module_score implements the same method.
      obj <- scratch_module_score(obj, mp_sets, name = "MPscore", seed = params$seed)
      score_cols <- grep("^MPscore\\d+$", colnames(obj@meta.data), value = TRUE)
      names(score_cols) <- names(mp_sets)[seq_along(score_cols)]

      sc <- as.data.table(obj@meta.data[, score_cols, drop = FALSE], keep.rownames = "cell_id")
      setnames(sc, score_cols, paste0("metaprog_", names(score_cols)))

      # Assign each cell to its highest-scoring metaprogram.
      mat <- as.matrix(sc[, -1, with = FALSE])
      sc[, metaprog_assigned := colnames(mat)[max.col(mat, ties.method = "first")]]
      sc[, metaprog_assigned := sub("^metaprog_", "", metaprog_assigned)]

      # Metaprograms are derived from the MALIGNANT compartment, so assigning
      # every cell its best-scoring programme made the recurrence table read
      # "V2, 2 samples, 24,415 cells" -- i.e. the whole dataset, stroma and
      # immune included. Scores stay on all cells (they are just signature
      # scores); the ASSIGNMENT is restricted to the compartment the programmes
      # came from, when the split is available.
      if ("malignant_status" %in% names(meta)) {
        mal <- meta[match(sc$cell_id, meta$cell_id), malignant_status]
        sc[!(mal %in% "malignant"), metaprog_assigned := NA_character_]
        cat("metaprog assignment restricted to",
            sum(!is.na(sc$metaprog_assigned)), "malignant cells\n")
      }

      matched <- sum(meta$cell_id %in% sc$cell_id)
      meta <- merge(meta, sc, by = "cell_id", all.x = TRUE, sort = FALSE)
      record_join("metaprog", basename(mp_files[1]), matched, nrow(meta),
                  length(setdiff(names(sc), "cell_id")))
      cat("scored", length(mp_sets), "metaprograms\n")
    }
  }
} else {
  cat("no metaprogram output\n")
}
```

## Trajectory

```{r trajectory}
fs <- stage_files("trajectory")
pt <- fs[grepl("pseudotime|concordance|slingshot|monocle|cytotrace",
               basename(fs), ignore.case = TRUE)]
if (length(pt)) {
  meta <- join_stage(meta, read_cell_table(pt[1]), "traj", "trajectory", basename(pt[1]))
  cat("trajectory ->", basename(pt[1]), "\n")
} else {
  cat("no trajectory output\n")
  record_join("trajectory", NA_character_, 0L, nrow(meta), 0L)
}
```

## TCR

```{r tcr}
fs <- stage_files("tcr")
tc <- fs[grepl("cell|clone|clonotype", basename(fs), ignore.case = TRUE)]
if (length(tc)) {
  meta <- join_stage(meta, read_cell_table(tc[1]), "tcr", "tcr", basename(tc[1]))
  cat("tcr ->", basename(tc[1]), "\n")
} else {
  cat("no TCR output\n")
  record_join("tcr", NA_character_, 0L, nrow(meta), 0L)
}
```

## Assemble and save

```{r assemble}
setDF(meta)
rownames(meta) <- meta$cell_id

# Only add columns the object does not already carry.
new_cols <- setdiff(colnames(meta), c("cell_id", colnames(obj@meta.data)))
if (length(new_cols)) {
  obj <- AddMetaData(obj, meta[colnames(obj), new_cols, drop = FALSE])
}

joined <- joined_stages()
cat("Stages joined:", if (length(joined)) paste(joined, collapse = ", ") else "none", "\n")
cat("New per-cell columns added:", length(new_cols), "\n")

saveRDS(obj, file.path("data", paste0(params$project_name, "_master_object.RDS")))
fwrite(meta, file.path("data", paste0(params$project_name, "_cell_table.csv")))

# The status table now carries the EVIDENCE for each verdict, not just the
# verdict: which file was read, how many of the master's cells it matched, and
# how many columns it contributed. A TRUE with matched = 0 is impossible by
# construction now, and a stage that was staged but matched nothing is visible
# rather than indistinguishable from a stage that never ran.
all_stages <- c("annotation", "cnv_infercnv", "stratify",
                "metaprog", "trajectory", "tcr")
status <- data.table(stage = all_stages)
status <- merge(status, join_log, by = "stage", all.x = TRUE, sort = FALSE)
status[is.na(matched), `:=`(matched = 0L, columns = 0L)]
status[, total := nrow(meta)]
status[, joined := matched > 0L]
status[, match_pct := round(100 * matched / pmax(1L, total), 1)]
setcolorder(status, c("stage", "joined", "matched", "total", "match_pct", "columns", "file"))
fwrite(status, file.path("data", "stage_join_status.csv"))
```

```{r join-findings, results = 'asis'}
# A stage that contributed a file but matched no cells is the failure this
# notebook used to hide. Say so where it will be read.
staged_but_empty <- status[!is.na(file) & matched == 0L]
if (nrow(staged_but_empty)) {
  scratch_finding(sprintf(paste(
    "**%s** contributed a file but matched **0 of %s** cells: %s.",
    "Its columns were left off the master object rather than added as NAs.",
    "The usual cause is a barcode format that does not match the master object's",
    "`cell_id` (a sample prefix present on one side only)."),
    paste(staged_but_empty$stage, collapse = ", "),
    format(nrow(meta), big.mark = ","),
    paste(sprintf("%s -> %s", staged_but_empty$stage, staged_but_empty$file),
          collapse = "; ")),
    type = "caution", title = "Stage joined nothing")
}

partial <- status[matched > 0L & match_pct < 95]
if (nrow(partial)) {
  scratch_finding(sprintf(
    "Partial joins: %s. Cells outside the matched set carry NA for those columns.",
    paste(sprintf("**%s** %s%% (%s of %s)", partial$stage, partial$match_pct,
                  format(partial$matched, big.mark = ","),
                  format(partial$total, big.mark = ",")), collapse = "; ")),
    type = "note", title = "Partial joins")
}

ok <- status[joined == TRUE]
if (nrow(ok)) {
  scratch_finding(sprintf(
    "Joined %s stage(s) onto %s cells: %s.",
    nrow(ok), format(nrow(meta), big.mark = ","),
    paste(sprintf("**%s** (+%s cols)", ok$stage, ok$columns), collapse = ", ")),
    type = "note", title = "Cross-stage join")
}

scratch_table(status, caption = "Per-stage join result, with the file each stage contributed")
```

```{r overview-plot, fig.width = 9, fig.height = 4}
avail <- intersect(c(params$celltype_col, "metaprog_assigned"), colnames(obj@meta.data))
if (length(avail) && "umap" %in% Reductions(obj)) {
  for (col in avail) {
    print(DimPlot(obj, group.by = col, reduction = "umap", label = TRUE, repel = TRUE) +
            ggtitle(col) + theme(legend.position = "right"))
    ggsave(file.path("figures", paste0("umap_", col, ".png")), width = 9, height = 6, dpi = 400)
  }
} else {
  cat("No UMAP reduction on the object; skipping overview plots.\n")
}
```

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
