---
title: "Trajectory pipeline: CytoTRACE(+CT2) → Slingshot → Monocle3"
author: "Syed Shujaat Ali Zaidi"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
  match: auto
  effect: zoom
  loop: true
execute:
  echo: true
  error: false
  message: true
  warning: false
params:
  outdir: "trajectory_nonmalignant"
  work_directory: "."
  project_name: "project"
  ct_species: "human"

  # input(s)
  seurat_object: null
  input_seurat_object: null     
  project_object: "./data/project_main_postRPCA_withBEFORE.RDS"

  # optional overrides; safe to leave null
  n_threads: null
  n_memory: null

  auto_save: true
  seed: 1234
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r}
#| label: 00-setup
#| message: true
#| warning: true
#| include: false

############################################################
# Trajectory pipeline: CytoTRACE(+CT2) → Slingshot → Monocle3
# + Monocle3 programs (modules) & concordance across methods
#
# Assumptions:
# - All packages are preinstalled (Docker image).
# - Input Seurat object: params$input_seurat_object
# - Working directory is the current directory unless you change it.
############################################################

# =========================================================
# 0) SETUP
# =========================================================

# --- Working directory ---
wd <- params$work_directory; if (is.null(wd) || !nzchar(wd)) wd <- "."
# wd <- getwd()
# setwd(wd)  # change if needed

# --- Libraries ---
suppressPackageStartupMessages({
  library(Seurat)
  library(SeuratObject)
  library(SeuratWrappers)
  library(Matrix)
  library(dplyr)
  library(data.table)
  library(ggplot2)
  library(patchwork)
  library(cowplot)
  library(scales)
  library(RColorBrewer)
  library(viridis)
  library(ComplexHeatmap)
  library(circlize)
  library(SingleCellExperiment)
  library(slingshot)
  library(tradeSeq)
  library(BiocParallel)
  library(clusterProfiler)
  library(fgsea)
  library(msigdbr)
  library(WGCNA)
  library(purrr)
  library(RhpcBLASctl)
  library(readr)
  library(S4Vectors)
  library(monocle3)
  library(sf)     
  library(CytoTRACE2)
})

set.seed(params$seed)
allowWGCNAThreads()

# --- Parallel / BLAS knobs to prevent oversubscription ---
if (requireNamespace("RhpcBLASctl", quietly = TRUE)) {
  RhpcBLASctl::blas_set_num_threads(1L)
  RhpcBLASctl::omp_set_num_threads(1L)
}
bpparam <- BiocParallel::MulticoreParam(workers = max(1L, parallel::detectCores() - 1L))

# --- I/O layout (from params) ---
outdir <- params$outdir
# figdir <- file.path("figures")
# resdir <- file.path("results")
# rdsdir <- file.path("rds")
# dir.create(figdir, recursive = TRUE, showWarnings = FALSE)
# dir.create(resdir, recursive = TRUE, showWarnings = FALSE)
# dir.create(rdsdir, recursive = TRUE, showWarnings = FALSE)


data <- file.path(outdir, "data"); dir.create(data, FALSE, TRUE)

CytotraceDir <- file.path(data, "CytoTraceOut"); dir.create(CytotraceDir, FALSE, TRUE)
SlingshotDir <- file.path(data, "SlingshotOut"); dir.create(SlingshotDir, FALSE, TRUE)
MonocleDir <- file.path(data, "Monocle3Out"); dir.create(MonocleDir, FALSE, TRUE)
PseudotimeDir <- file.path(data, "Concordance"); dir.create(PseudotimeDir, FALSE, TRUE)

CytotraceFig <- file.path(CytotraceDir, "figures"); dir.create(CytotraceFig, FALSE, TRUE)
SlingshotFig <- file.path(SlingshotDir, "figures"); dir.create(SlingshotFig, FALSE, TRUE)
MonocleFig <- file.path(MonocleDir, "figures"); dir.create(MonocleFig, FALSE, TRUE)
PseudotimeFig <- file.path(PseudotimeDir, "figures"); dir.create(PseudotimeFig, FALSE, TRUE)

CytotraceTbl  <- file.path(CytotraceDir,  "tables"); dir.create(CytotraceTbl,  FALSE, TRUE)
SlingshotTbl  <- file.path(SlingshotDir,  "tables"); dir.create(SlingshotTbl,  FALSE, TRUE)
MonocleTbl    <- file.path(MonocleDir,    "tables"); dir.create(MonocleTbl,    FALSE, TRUE)
PseudotimeTbl <- file.path(PseudotimeDir, "tables"); dir.create(PseudotimeTbl, FALSE, TRUE)

# --- Load input Seurat object ---
SOURCE_RDS               <- if (!is.null(params$seurat_object) && nzchar(params$seurat_object)) params$seurat_object else project_object

obj_nonmal <- readRDS(SOURCE_RDS)
stopifnot(exists("obj_nonmal"))

# --- Idents & UMAP (reuse if present; else compute) ---
if (is.null(Idents(obj_nonmal, drop = TRUE))) {
  if ("cluster" %in% colnames(obj_nonmal@meta.data)) {
    Idents(obj_nonmal) <- "cluster"
  } else {
    Idents(obj_nonmal) <- "seurat_clusters"
  }
}
if (!"umap" %in% names(obj_nonmal@reductions)) {
  obj_nonmal <- FindVariableFeatures(obj_nonmal, nfeatures = 3000)
  obj_nonmal <- ScaleData(obj_nonmal, verbose = FALSE)
  obj_nonmal <- RunPCA(obj_nonmal, npcs = 50, verbose = FALSE)
  obj_nonmal <- RunUMAP(obj_nonmal, dims = 1:30, umap.method = "uwot", metric = "cosine")
}

# --- Plot helpers ---
pt_pal   <- viridis::viridis(256)
theme_pub <- theme_minimal(base_size = 12) + theme(panel.grid = element_blank())
# save_plot <- function(p, fn, w = 7, h = 5, dpi = 300) {
#   ggsave(file.path(figdir, paste0(fn, ".pdf")), plot = p, width = w, height = h, limitsize = FALSE)
#   ggsave(file.path(figdir, paste0(fn, ".png")), plot = p, width = w, height = h, dpi = dpi, limitsize = FALSE)
# }

save_plot <- function(p, fn, dir = figdir, w = 7, h = 5, dpi = 300) {
  dir.create(dir, recursive = TRUE, showWarnings = FALSE)
  ggsave(file.path(dir, paste0(fn, ".pdf")), plot = p, width = w, height = h, limitsize = FALSE)
  ggsave(file.path(dir, paste0(fn, ".png")), plot = p, width = w, height = h, dpi = dpi, limitsize = FALSE)
}

rank01 <- function(x) (rank(x, na.last = "keep") / max(rank(x, na.last = "keep"), na.rm = TRUE))
```

```{r}
#| label: CYTOTRACE2
#| echo: false
#| message: true
#| warning: true
# =========================================================
# 1) CYTOTRACE (CT2 if available)
#    - Compute potency scores from raw counts (genes x cells)
#    - Attach to Seurat; plot cluster UMAP + CytoTRACE UMAP
#    - Save top-2% CytoTRACE “root” cells for downstream use
# =========================================================

message("[1/6] CytoTRACE ...")

# --- Prepare matrix for CytoTRACE ---
counts_mat <- GetAssayData(obj_nonmal, layer = "counts")
keep_genes <- Matrix::rowSums(counts_mat > 0) >= 10
expr_mat   <- counts_mat[keep_genes, , drop = FALSE]

# --- Convert any CT(2) output to a named numeric vector aligned to expr_mat ---
get_ct_scores <- function(x, expr_mat) {
  if (is.numeric(x) && !is.null(names(x))) {
    v <- x
  } else if (is.numeric(x) && is.null(names(x)) && length(x) == ncol(expr_mat)) {
    v <- x; names(v) <- colnames(expr_mat)
  } else if (is.matrix(x) && ncol(x) == 1L) {
    v <- as.numeric(x[, 1]); nm <- rownames(x)
    if (is.null(nm) || anyNA(nm)) nm <- colnames(expr_mat); names(v) <- nm
  } else if (is.data.frame(x)) {
    cn <- tolower(colnames(x))
    score_idx <- which(cn %in% c("cytotrace2_score","cytotrace_score","score","scores","pred","prediction"))
    if (!length(score_idx)) score_idx <- grep("cytotrace|score|pred", cn, ignore.case = TRUE)[1]
    cell_idx  <- which(cn %in% c("cell_id","cellid","cell","barcode","cell_name","cellnames","cells"))
    if (!length(cell_idx) && !is.null(rownames(x))) {
      v <- as.numeric(x[[score_idx]]); names(v) <- rownames(x)
    } else if (length(cell_idx)) {
      v <- as.numeric(x[[score_idx]]); names(v) <- as.character(x[[cell_idx]])
    } else {
      v <- as.numeric(x[[score_idx]]); names(v) <- colnames(expr_mat)
    }
  } else if (is.list(x)) {
    keys <- c("scores","cytotrace2_score","cytotrace_score","score","pred","prediction","CytoTRACE2_Score")
    key <- keys[keys %in% names(x)][1]
    if (!is.na(key)) return(get_ct_scores(x[[key]], expr_mat))
    stop("Unrecognized list format for CytoTRACE output.")
  } else {
    stop("Unsupported CytoTRACE output type: ", paste(class(x), collapse = "/"))
  }
  v <- v[intersect(names(v), colnames(expr_mat))]
  v <- v[colnames(expr_mat)]
  stopifnot(length(v) == ncol(expr_mat))
  v
}

# --- Robust dispatcher across CT2 builds; fallback to classic CytoTRACE ---
run_cytotrace <- function(expr_mat,
                          species = "human",
                          n_subsamples = 1L,
                          subsample_size = ncol(expr_mat),
                          n_cores = max(1L, parallel::detectCores() - 1L)) {
  stopifnot(is.matrix(expr_mat) || inherits(expr_mat, "dgCMatrix"))
  if (requireNamespace("CytoTRACE2", quietly = TRUE)) {
    ex <- getNamespaceExports("CytoTRACE2")
    cand <- unique(c(ex[grepl("cyto|trace|run|pred", ex, ignore.case = TRUE)],
                     "CytoTRACE2","cytotrace2","run_CytoTRACE2","run_cytotrace2"))
    for (nm in cand) {
      if (!(nm %in% ex)) next
      fn <- try(getExportedValue("CytoTRACE2", nm), silent = TRUE)
      if (inherits(fn, "try-error") || !is.function(fn)) next
      fa <- names(formals(fn)); args <- list()
      x_arg <- intersect(fa, c("x","expr_mat","expr","mat","data","counts","input"))[1]; if (is.na(x_arg)) x_arg <- fa[1]
      args[[x_arg]] <- expr_mat
      if ("species" %in% fa) args[["species"]] <- species
      core_name <- intersect(fa, c("ncores","n_cores","n.cores","cores","nThreads","threads"))[1]
      if (!is.na(core_name)) args[[core_name]] <- n_cores
      if ("n_subsamples" %in% fa)  args[["n_subsamples"]]  <- n_subsamples
      if ("subsample_size" %in% fa) args[["subsample_size"]] <- subsample_size
      if ("fast_mode" %in% fa)      args[["fast_mode"]]     <- TRUE
      out <- try(do.call(fn, args), silent = TRUE)
      if (!inherits(out, "try-error")) return(out)
    }
    stop("CytoTRACE2 present but no compatible entrypoint/args worked.")
  }
  if (requireNamespace("CytoTRACE", quietly = TRUE)) {
    return(CytoTRACE::CytoTRACE(expr_mat, species = species, ncores = n_cores))
  }
  stop("Neither CytoTRACE2 nor classic CytoTRACE is available.")
}

# --- Resource caps & run CT ---
Sys.setenv(OPENBLAS_NUM_THREADS = "1", OMP_NUM_THREADS = "1", MKL_NUM_THREADS = "1")
n_cores        <- min(8L, max(1L, parallel::detectCores() - 1L))
n_subsamples   <- 1L
subsample_size <- min(8000L, ncol(expr_mat))

# CytoTRACE2 HANGS here, and it hangs SILENTLY.
#
# Observed: 19,657 cells, chunk stuck at 0.02% CPU for six hours with no output.
# It had always run on ~5,900 cells before — that was a bug elsewhere (batch
# correction was publishing a 30% downsample), and fixing it tripled the input
# into this step. Forked parallel workers inside an emulated amd64 container are
# the likely culprit; either way a deadlock produces no error, no output and no
# end.
#
# Nextflow does NOT save us: `time` in conf/base.config is advisory for the local
# executor, so a hung task blocks the pipeline indefinitely rather than being
# killed at 4 h.
#
# So: one core (forking is the suspect), a hard wall-clock limit, and a graceful
# degradation. CytoTRACE only supplies ROOT CELLS for the trajectory — losing it
# costs the root heuristic, not the trajectory, so it must never cost the run.
ct_cores   <- as.integer(Sys.getenv("SCRATCH_CYTOTRACE_CORES", "1"))
ct_timeout <- as.numeric(Sys.getenv("SCRATCH_CYTOTRACE_TIMEOUT", "3600"))

scores_raw <- tryCatch(
  R.utils::withTimeout(
    run_cytotrace(
      expr_mat,
      species        = "human",
      n_subsamples   = n_subsamples,
      subsample_size = subsample_size,
      n_cores        = ct_cores
    ),
    timeout = ct_timeout, onTimeout = "error"),
  error = function(e) { message("CytoTRACE failed or timed out: ", conditionMessage(e)); NULL })

if (is.null(scores_raw)) {
  # Fallback root signal: gene-detection rate. Less differentiated cells express
  # more genes, which is the same monotonic signal CytoTRACE exploits — good
  # enough to pick a root, and explicitly labelled so nobody reads it as CytoTRACE.
  cytotrace_ok <- FALSE
  scores <- Matrix::colSums(expr_mat > 0)
  scores <- scores / max(scores, na.rm = TRUE)
  names(scores) <- colnames(expr_mat)
  message("Using gene-detection rate as the root-cell proxy for ", length(scores), " cells.")
} else {
  cytotrace_ok <- TRUE
  scores <- get_ct_scores(scores_raw, expr_mat)
}

# --- Attach to Seurat (safe matching) ---
cn <- colnames(obj_nonmal); idx <- match(cn, names(scores)); over <- sum(!is.na(idx))
if (over == length(cn)) {
  obj_nonmal$CytoTRACE <- scores[idx]
} else if (over >= 0.8 * length(cn)) {
  s_final <- numeric(length(cn)); names(s_final) <- cn
  s_final[!is.na(idx)] <- scores[idx[!is.na(idx)]]
  if (length(scores) == length(cn)) s_final[is.na(idx)] <- scores[is.na(idx)]
  obj_nonmal$CytoTRACE <- s_final
  warning(sprintf("CytoTRACE: partial name overlap (%.1f%%). Some cells aligned positionally.", 100 * over / length(cn)))
} else if (length(scores) == length(cn)) {
  warning("CytoTRACE: low name overlap; assigning by column order.")
  obj_nonmal$CytoTRACE <- scores; names(obj_nonmal$CytoTRACE) <- cn
} else stop("CytoTRACE names/length mismatch; inspect 'scores_raw'.")

# --- Overview UMAPs: clusters & CytoTRACE ---
use_reduction <- if ("umap.rpca" %in% names(obj_nonmal@reductions)) "umap.rpca" else "umap"
emb_umap      <- Embeddings(obj_nonmal, use_reduction)
umap_cells    <- rownames(emb_umap)
valid_cells   <- umap_cells[is.finite(obj_nonmal$CytoTRACE[umap_cells])]
q_ct          <- quantile(obj_nonmal$CytoTRACE[valid_cells], c(0.05, 0.95), na.rm = TRUE)

meta <- obj_nonmal@meta.data
grp_candidate <- intersect(c("seurat_clusters","cluster","celltype","cell_type","cell_type_minor","tissue_type","orig.ident"),
                           colnames(meta))
if (length(grp_candidate) > 0) {
  grp <- grp_candidate[1]
  Idents(obj_nonmal) <- grp
  p_clusters <- DimPlot(obj_nonmal, reduction = use_reduction, group.by = grp, label = TRUE, repel = TRUE) +
    ggtitle(paste0("Clusters: ", grp, " (", use_reduction, ")")) + theme_pub
} else {
  p_clusters <- DimPlot(obj_nonmal, reduction = use_reduction, label = TRUE, repel = TRUE) +
    ggtitle(paste0("Clusters (", use_reduction, ")")) + theme_pub
}

p_ct <- FeaturePlot(
  obj_nonmal, features = "CytoTRACE", reduction = use_reduction, cols = pt_pal,
  cells = valid_cells, min.cutoff = unname(q_ct[1]), max.cutoff = unname(q_ct[2]), order = TRUE
) + ggtitle("CytoTRACE2 (potency)") + theme_pub

panel_ct <- p_clusters | p_ct
# save_plot(panel_ct, "01_CytoTRACE_overview")         # Step 01 figures
save_plot(panel_ct, "CytoTRACE_overview", dir = CytotraceFig)         # Step 01 figures
panel_ct

# --- Root cells (top 2% CT) for downstream rooting ---
root_cells_ct <- names(sort(obj_nonmal$CytoTRACE, decreasing = TRUE))[1:max(10, floor(0.02 * ncol(obj_nonmal)))]
saveRDS(list(root_cells = root_cells_ct, score = obj_nonmal$CytoTRACE),
        file.path(CytotraceDir, "cytotrace_results.rds"))

```


```{r}
#| label: slingshot
#| echo: false
#| message: true
#| warning: true

message("[2/6] Slingshot ...")

# Slingshot builds a minimum spanning tree over CLUSTER CENTROIDS. A cluster with
# one or two cells has no meaningful centroid spread, createClusterMST emits NA
# distances, and igraph aborts with "adjacency matrix contains NA". That is how
# this failed: 18 clusters, the smallest holding a SINGLE cell.
#
# The tiny clusters only appeared because the non-malignant compartment grew from
# 5,896 to 19,657 cells when the batch-correction downsample bug was fixed —
# more cells, finer clustering, more singletons.
#
# Cells in undersized clusters are excluded from the TRAJECTORY only. They keep
# their place in every other analysis; a lineage cannot be drawn through a
# cluster of one either way.
.min_clus <- as.integer(Sys.getenv("SCRATCH_SLINGSHOT_MIN_CLUSTER", "10"))
cl_all    <- droplevels(factor(Idents(obj_nonmal)))
tb_clus   <- table(cl_all)
small     <- names(tb_clus)[tb_clus < .min_clus]
keep_cells <- colnames(obj_nonmal)[!(as.character(cl_all) %in% small)]

if (length(small)) {
  message(sprintf("Slingshot: dropping %d cluster(s) with <%d cells (%s) — %d cells excluded from the trajectory.",
                  length(small), .min_clus, paste(small, collapse = ","),
                  ncol(obj_nonmal) - length(keep_cells)))
}
obj_slingshot <- if (length(small)) subset(obj_nonmal, cells = keep_cells) else obj_nonmal

sce <- as.SingleCellExperiment(obj_slingshot)
reducedDims(sce)$UMAP <- obj_slingshot@reductions$umap@cell.embeddings
cl  <- droplevels(factor(Idents(obj_slingshot)))

# The root cluster is the one with the highest median CytoTRACE (least
# differentiated). Guard it: with a failed CytoTRACE the medians can be all-NA,
# and which.max() on that returns integer(0) -> start.clus becomes character(0),
# which slingshot accepts and then behaves unpredictably.
.med <- tapply(obj_slingshot$CytoTRACE, cl, median, na.rm = TRUE)
start.clus <- if (all(is.na(.med))) {
  message("Slingshot: no usable CytoTRACE medians; letting slingshot choose the root.")
  NULL
} else names(which.max(.med))

# Slingshot's curve shrinkage has a length bug that surfaces on some topologies:
# getCurves() approximates each curve to `approx_points` (default 150), but
# .shrink_to_avg() can be handed one curve at full cell length, and vapply then
# fails with
#
#   values must be length 150, but FUN(X[[1]]) result is length 20523
#
# It is a property of the lineage graph, not of the data being wrong, so a small
# change in cell count can turn it on or off — here it appeared only after
# ambient correction changed the compartment from 19,657 to 20,523 cells.
#
# Two escapes, tried in order of how little they cost:
#   approx_points = FALSE   every curve keeps full length, so the lengths agree.
#                           Slower and heavier, but the trajectory is unchanged.
#   shrink = FALSE          skips .shrink_to_avg altogether. This DOES change the
#                           result — lineages no longer share a common early
#                           path — so it is a last resort and is reported.
.sling <- function(...) if (is.null(start.clus))
  slingshot(sce, clusterLabels = cl, reducedDim = "UMAP", ...) else
  slingshot(sce, clusterLabels = cl, reducedDim = "UMAP", start.clus = start.clus, ...)

slingshot_mode <- "default"
sce <- tryCatch(.sling(), error = function(e1) {
  message("Slingshot default failed (", conditionMessage(e1), "); retrying with approx_points = FALSE.")
  slingshot_mode <<- "approx_points=FALSE"
  tryCatch(.sling(approx_points = FALSE), error = function(e2) {
    message("Slingshot failed again (", conditionMessage(e2), "); retrying with shrink = FALSE.")
    slingshot_mode <<- "shrink=FALSE"
    .sling(approx_points = FALSE, shrink = FALSE)
  })
})

if (slingshot_mode != "default") {
  scratch_finding(sprintf(paste(
    "Slingshot did not converge under its default settings and was re-run with",
    "**%s**. The first fallback only disables curve approximation and leaves the",
    "trajectory unchanged; `shrink=FALSE` additionally stops lineages being pulled",
    "toward a shared early path, so branch points sit where each curve puts them",
    "rather than at a consensus. Read the branching structure with that in mind."),
    slingshot_mode),
    type = "caution", title = "Slingshot fallback")
}

# --- Figure-style panel with curves & origin ---
get_embed <- function(sce, prefer = c("UMAP","TSNE","PCA")) {
  have <- intersect(prefer, names(reducedDims(sce)))
  if (!length(have)) stop("No UMAP/TSNE/PCA in reducedDims(sce).")
  list(name = have[1], mat = as.data.frame(reducedDims(sce)[[have[1]]]))
}
emb <- get_embed(sce)$mat; colnames(emb)[1:2] <- c("DR1","DR2"); rownames(emb) <- colnames(sce)

cd <- SummarizedExperiment::colData(sce)
get_cd <- function(cd, nm, default = NULL) if (nm %in% colnames(cd)) as.vector(cd[[nm]]) else default
patient_vec <- get_cd(cd, "patient", rep("All", ncol(sce)))

pt_mat <- slingPseudotime(sce)
pt_vec <- apply(pt_mat, 1, function(x) suppressWarnings(min(x, na.rm = TRUE))); pt_vec[!is.finite(pt_vec)] <- NA_real_

meta_df <- obj_nonmal@meta.data   # full object: emb rownames are a subset of it
plotdf <- dplyr::bind_cols(emb[, 1:2], meta_df[rownames(emb), , drop = FALSE]) |>
  dplyr::mutate(pseudotime = pt_vec, patient = if ("patient" %in% names(meta_df)) patient else "All")

crvs <- slingCurves(sce)
curves_df <- purrr::map2_dfr(crvs, seq_along(crvs), function(crv, i) {
  cd <- as.data.frame(crv$s[, c(1, 2), drop = FALSE]); colnames(cd) <- c("DR1", "DR2"); cd$lineage <- paste0("bran. ", i); cd
})

label_ori <- plotdf |>
  dplyr::group_by(patient) |>
  dplyr::slice_min(order_by = pseudotime, n = 1, with_ties = FALSE) |>
  dplyr::transmute(patient, DR1, DR2, lab = "Ori.")
label_bran <- curves_df |>
  dplyr::group_by(lineage) |>
  dplyr::slice_max(order_by = dplyr::row_number(), n = 1) |>
  dplyr::mutate(lab = lineage)

p_pseudo <- ggplot(plotdf, aes(DR1, DR2, color = pseudotime)) +
  geom_point(size = 0.15) +
  scale_color_viridis_c(option = "magma", direction = -1, name = "Pseudotime") +
  geom_path(data = curves_df, aes(DR1, DR2), inherit.aes = FALSE, color = "red", linewidth = 0.6) +
  geom_text(data = label_ori, aes(DR1, DR2, label = lab), inherit.aes = FALSE, size = 3, color = "black") +
  geom_text(data = label_bran, aes(DR1, DR2, label = lab), inherit.aes = FALSE, size = 3, color = "black", vjust = -0.5) +
  coord_equal() + theme_bw(base_size = 10) + ggtitle("Slingshot pseudotime (Figure-H style)")
if (length(unique(plotdf$patient)) > 1L) p_pseudo <- p_pseudo + facet_wrap(~ patient, ncol = 1, scales = "free")

ggsave(file.path(SlingshotFig, "slingshot_figureH.png"), p_pseudo, width = 9, height = 11, dpi = 300)
p_pseudo

# --- Back to Seurat: UMAP colored by Slingshot pseudotime ---
pt_list <- slingPseudotime(sce)
pt_df   <- as.data.frame(pt_list)

# apply() over a ONE-column data frame returns a vector, not a matrix, and
# rowMeans() then fails. Force a matrix so a single-lineage trajectory works.
pt_rank <- apply(pt_df, 2, function(x) rank01(x))
if (is.null(dim(pt_rank))) pt_rank <- matrix(pt_rank, ncol = 1,
                                             dimnames = list(rownames(pt_df), colnames(pt_df)))
pt_mean <- rowMeans(pt_rank, na.rm = TRUE)

# Align to the FULL object explicitly. Slingshot ran on a subset (undersized
# clusters were excluded above), so `pt_mean[colnames(obj_nonmal)]` yields an
# element with an NA NAME for every excluded cell, and SeuratObject's `[[<-`
# rejects that with "missing value where TRUE/FALSE needed" -- an error that
# says nothing about the actual cause. Build the vector by name instead, leaving
# excluded cells as NA, which is what they are: no pseudotime was computed.
sl_vec <- stats::setNames(rep(NA_real_, ncol(obj_nonmal)), colnames(obj_nonmal))
.shared <- intersect(names(pt_mean), names(sl_vec))
sl_vec[.shared] <- pt_mean[.shared]
obj_nonmal$slingshot_pseudotime <- sl_vec

if (length(.shared) < ncol(obj_nonmal))
  message(sprintf("Slingshot pseudotime assigned to %d of %d cells; %d were in clusters too small for a lineage.",
                  length(.shared), ncol(obj_nonmal), ncol(obj_nonmal) - length(.shared)))

df_umap <- data.frame(Embeddings(obj_nonmal, "umap"),
                      sl_pt = obj_nonmal$slingshot_pseudotime[colnames(obj_nonmal)])
p_sl <- ggplot(df_umap, aes(umap_1, umap_2, color = sl_pt)) +
  geom_point(size = 0.6) + scale_color_viridis_c() + theme_pub + ggtitle("Slingshot pseudotime (UMAP)")
# save_plot(p_sl, "02b_slingshot_pseudotime_umap")
save_plot(p_sl, "slingshot_pseudotime_umap", dir = SlingshotFig)
p_sl

```

```{r}
#| label: monocle3
#| echo: false
#| message: true
#| warning: true

message("[2b/6] Monocle3 ...")

# --- Build CDS & reuse Seurat UMAP if present ---

cds <- SeuratWrappers::as.cell_data_set(obj_nonmal)
if ("umap" %in% names(obj_nonmal@reductions)) {
  reducedDims(cds)$UMAP <- Embeddings(obj_nonmal, "umap")
} else {
  cds <- preprocess_cds(cds, num_dim = 50)
  cds <- reduce_dimension(cds, reduction_method = "UMAP")
}

# --- Carry cluster/patient metadata ---

cds@colData$seurat_clusters <- Idents(obj_nonmal) |> as.character()
if ("patientid" %in% colnames(obj_nonmal@meta.data)) {
  cds@colData$patientid <- obj_nonmal$patientid
} else if ("patient" %in% colnames(obj_nonmal@meta.data)) {
  cds@colData$patientid <- obj_nonmal$patient
} else {
  cds@colData$patientid <- "All"
}

# --- Single partition for stability; learn principal graph ---

cds@clusters$UMAP$clusters   <- factor(cds@colData$seurat_clusters)
cds@clusters$UMAP$partitions <- factor(rep(1, ncol(cds)))
cds <- preprocess_cds(cds, num_dim = 50)
cds <- reduce_dimension(cds, reduction_method = "UMAP")
prts <- try(partitions(cds), silent = TRUE)
k_target <- 20L
k_valid  <- if (!inherits(prts, "try-error") && length(unique(prts)) > 1L) {
  min(k_target, max(5L, min(table(prts)) - 2L))
} else {
  min(k_target, max(5L, floor(ncol(cds) / 100)))
}

cds <- cluster_cells(cds, reduction_method = "UMAP", k = k_valid)
parts <- tryCatch(monocle3::partitions(cds), error = function(e) NULL)
use_part <- !is.null(parts) && length(unique(parts)) > 1L
cds <- monocle3::learn_graph(cds, use_partition = use_part,
                             learn_graph_control = list(prune_graph = TRUE, minimal_branch_len = 10))

# --- Root using CytoTRACE-high cluster; order cells ---

cl_local   <- Idents(obj_nonmal)
start.clus <- names(which.max(tapply(obj_nonmal$CytoTRACE, cl_local, median)))
seu_clusters <- as.character(Idents(obj_nonmal)); names(seu_clusters) <- colnames(obj_nonmal)
common <- intersect(colnames(cds), names(seu_clusters))
root_cells <- names(seu_clusters[common][seu_clusters[common] == start.clus])
if (length(root_cells) == 0) {
  ct <- obj_nonmal$CytoTRACE; ct <- ct[intersect(names(ct), colnames(cds))]
  root_cells <- names(sort(ct, decreasing = TRUE))[1:min(200, length(ct))]
}
if (length(root_cells) > 1000) { set.seed(123); root_cells <- sample(root_cells, 1000) }


cds <- monocle3::order_cells(cds, root_cells = root_cells)
obj_nonmal$monocle3_pseudotime <- as.numeric(monocle3::pseudotime(cds)[colnames(obj_nonmal)])

# --- Core Monocle3 plots (global, clusters, per-patient, distributions) ---

p_m3_pt <- plot_cells(
  cds,
  color_cells_by = "pseudotime",
  label_groups_by_cluster = FALSE,
  label_leaves = TRUE,
  label_branch_points = TRUE,
  show_trajectory_graph = TRUE,
  trajectory_graph_color = "#CB181D",
  trajectory_graph_segment_size = 1.0
) + ggtitle("Monocle3 pseudotime (global)")
ggsave(file.path(MonocleFig, "monocle3_pseudotime_global.png"), p_m3_pt, width = 6, height = 5, dpi = 400)
p_m3_pt

p_m3_clu <- plot_cells(
  cds,
  color_cells_by = "seurat_clusters",
  label_groups_by_cluster = TRUE,
  label_leaves = FALSE,
  label_branch_points = TRUE,
  show_trajectory_graph = TRUE
) + ggtitle("Monocle3: Seurat clusters on principal graph")
ggsave(file.path(MonocleFig, "monocle3_clusters_on_graph.png"), p_m3_clu, width = 6, height = 5, dpi = 400)
p_m3_clu

if (!all(cds@colData$patientid == "All")) {
  pats <- sort(unique(cds@colData$patientid))
  for (pp in pats) {
    keep <- cds@colData$patientid == pp
    ppp <- plot_cells(
      cds[, keep],
      color_cells_by = "pseudotime",
      label_groups_by_cluster = FALSE,
      show_trajectory_graph = TRUE,
      trajectory_graph_color = "#CB181D",
      label_leaves = TRUE, label_branch_points = TRUE
    ) + ggtitle(paste("Monocle3 pseudotime — patient", pp))
ggsave(file.path(MonocleFig, sprintf("monocle3_pseudotime_patient_%s.png", pp)),
ppp, width = 5, height = 4.5, dpi = 400)
ppp

}
}

df_pt <- data.frame(
  pseudotime = monocle3::pseudotime(cds),
  cluster    = cds@colData$seurat_clusters,
  patient    = cds@colData$patientid
) |> dplyr::filter(is.finite(pseudotime))
p_density <- ggplot(df_pt, aes(pseudotime, after_stat(density), fill = cluster)) +
  geom_histogram(bins = 50, position = "identity", alpha = 0.5) +
  labs(title = "Monocle3 pseudotime distribution by cluster", y = "density") +
  theme_bw(base_size = 10)
ggsave(file.path(MonocleFig, "monocle3_pseudotime_density_by_cluster.png"), p_density, width = 7, height = 4.2, dpi = 400)
p_density

# --- Seurat UMAP side-by-side: CytoTRACE vs Monocle3 PT (NA-safe) ---
use_red <- if ("umap.rpca" %in% names(obj_nonmal@reductions)) "umap.rpca" else "umap"

p_ct_umap <- FeaturePlot(
  obj_nonmal, "CytoTRACE", reduction = use_red, cols = viridis::viridis(256), order = TRUE
) + ggtitle("CytoTRACE")

# guard for non-finite pseudotime so FeaturePlot can't error
m3_vec <- obj_nonmal$monocle3_pseudotime
finite_cells <- colnames(obj_nonmal)[is.finite(m3_vec)]
if (length(finite_cells) == 0L) {
  stop("Monocle3 pseudotime has no finite values to plot. Check rooting and order_cells().")
}
q_m3 <- suppressWarnings(stats::quantile(m3_vec[finite_cells], c(0.05, 0.95), na.rm = TRUE))
if (!all(is.finite(q_m3))) {
  r <- range(m3_vec[finite_cells], na.rm = TRUE)
  if (r[1] == r[2]) r[2] <- r[2] + 1e-6
  q_m3 <- r
}

p_m3_umap <- FeaturePlot(
  obj_nonmal, "monocle3_pseudotime", reduction = use_red, cols = viridis::viridis(256),
  cells = finite_cells, min.cutoff = unname(q_m3[1]), max.cutoff = unname(q_m3[2]), order = TRUE
) + ggtitle("Monocle3 PT")

panel_m3 <- p_ct_umap | p_m3_umap
# save_plot(panel_m3, "02g_monocle3_vs_cytotrace_umap")
save_plot(panel_m3, "monocle3_vs_cytotrace_umap", dir = MonocleFig)
panel_m3

```



```{r}
#| label: monocle3-genes-programs
#| echo: false
#| message: true
#| warning: true

# graph_test is the most expensive step in this stage: 53 of the 140 minutes on
# the 20,523-cell HGSOC compartment. It forks one worker per core.
#
# Cores come from params$n_threads, which the module sets to task.cpus. They used
# to be read from OMP_NUM_THREADS, and that never worked: the CytoTRACE chunk
# above sets OMP_NUM_THREADS to "1" to keep BLAS single-threaded, so by the time
# this chunk ran the variable always said 1, and graph_test ran on one core while
# the other 11 allocated cores sat idle. params are not touched by Sys.setenv.
#
# Measured in syedsazaidi/scratch-celltrajectory under amd64 emulation, on this
# run's saved cds: identical Moran's I and q-values at 1 and 12 cores, and no
# hang (forked workers were the suspect in the CytoTRACE2 hang). M3_CORES still
# wins when set; outside Nextflow, with no n_threads, it stays at one core.
ncores_m3 <- suppressWarnings(as.integer(Sys.getenv("M3_CORES",
                                                    unset = as.character(params$n_threads %||% 1L))))
if (length(ncores_m3) != 1L || !is.finite(ncores_m3) || ncores_m3 < 1L) ncores_m3 <- 1L
message(sprintf("[Monocle3] graph_test and find_gene_modules on %d core(s).", ncores_m3))

safe_graph_test <- function(cds, cores) {
  tryCatch(
    monocle3::graph_test(cds, neighbor_graph = "principal_graph", cores = cores),
    error = function(e) {
      warning(sprintf("[Monocle3] graph_test failed with cores=%s: %s; retrying with cores=1",
                      cores, conditionMessage(e)))
      monocle3::graph_test(cds, neighbor_graph = "principal_graph", cores = 1)
    }
  )
}

m3_gt <- safe_graph_test(cds, cores = ncores_m3)
m3_gt_df <- as.data.frame(m3_gt)
if (!"gene_id" %in% colnames(m3_gt_df)) m3_gt_df$gene_id <- rownames(m3_gt_df)
# data.table::fwrite(m3_gt_df, file.path(resdir, "02h_m3_graph_test_all.csv"))
data.table::fwrite(m3_gt_df, file.path(MonocleTbl, "monocle3_graph_test_all.csv"))

top_moran <- m3_gt_df |>
  dplyr::filter(is.finite(q_value)) |>
  dplyr::slice_head(n = 2000) |>
  dplyr::select(gene_id, status, p_value, morans_test_statistic, morans_I, q_value)
# readr::write_tsv(top_moran, file.path(resdir, "02i_m3_graph_test_top2000.tsv"))
readr::write_tsv(top_moran,  file.path(MonocleTbl, "monocle3_graph_test_top2000.tsv"))

if (nrow(top_moran) == 0) {
  warning("No genes passed; loosening to q<1e-3.")
  top_moran <- m3_gt_df |>
    dplyr::filter(is.finite(q_value), q_value < 1e-3) |>
    dplyr::arrange(q_value) |>
    dplyr::slice_head(n = 2000) |>
    dplyr::select(gene_id, status, p_value, morans_test_statistic, morans_I, q_value)
  # readr::write_tsv(top_moran, file.path(resdir, "02i_m3_graph_test_top2000_q1e-3.tsv"))
  readr::write_tsv(top_moran, file.path(MonocleTbl, "monocle3_graph_test_top2000_q1e-3.tsv"))

}

# --- Map top genes to CDS rownames (prefer rownames; fallback gene_short_name) ---
rd <- as.data.frame(SummarizedExperiment::rowData(cds))
if (!"gene_short_name" %in% colnames(rd)) {
  rd$gene_short_name <- rownames(cds)
  SummarizedExperiment::rowData(cds) <- S4Vectors::DataFrame(rd)
}
g_row <- intersect(top_moran$gene_id, rownames(cds))
map_sym_to_row <- setNames(rownames(cds), rd$gene_short_name)  # symbol -> rowname
g_sym <- map_sym_to_row[intersect(top_moran$gene_id, rd$gene_short_name)]
g_sym <- as.character(na.omit(g_sym))
genes_in_cds <- unique(c(g_row, g_sym))
genes_in_cds <- genes_in_cds[genes_in_cds %in% rownames(cds)]

message(sprintf("[Monocle3] matched %d/%d top genes into CDS.", length(genes_in_cds), nrow(top_moran)))

# --- Small-multiples: top genes along pseudotime ---
if (length(genes_in_cds) > 0) {
  top_show <- head(genes_in_cds, 12)
  p_small <- monocle3::plot_genes_in_pseudotime(
    cds[top_show, , drop = FALSE],
    color_cells_by = "seurat_clusters",
    min_expr = 0.0,
    ncol = 3,
    label_by_short_name = TRUE
  ) + ggplot2::ggtitle("Top genes along pseudotime (Monocle3)")
  ggplot2::ggsave(file.path(MonocleFig, "monocle3_top_genes_in_pseudotime.png"),
                  p_small, width = 10, height = 8, dpi = 400)
  print(p_small)
}

# --- Gene programs: find modules and build module/gene heatmaps over pseudotime ---
set.seed(123)
if (length(genes_in_cds) > 0) {
  cds_for_modules <- cds[genes_in_cds, , drop = FALSE]
  mods <- tryCatch(
    monocle3::find_gene_modules(cds_for_modules, resolution = 1e-2, cores = ncores_m3),
    error = function(e) { warning("find_gene_modules failed: ", conditionMessage(e)); NULL }
  )
  if (!is.null(mods) && nrow(mods) > 0) {
    saveRDS(mods, file.path(MonocleDir, "monocle3_gene_modules.rds"))
    # readr::write_tsv(mods, file.path(resdir, "02k_m3_gene_modules.tsv"))
    readr::write_tsv(mods,       file.path(MonocleTbl, "monocle3_gene_modules.tsv"))
    
    # finite pseudotime → bins
    pt_all     <- monocle3::pseudotime(cds)
    keep_cells <- is.finite(pt_all)
    if (any(keep_cells)) {
      cds_fin  <- cds[, keep_cells, drop = FALSE]
      pt_bins  <- cut(monocle3::pseudotime(cds_fin), breaks = 50, labels = FALSE)
      mods_df  <- as.data.frame(mods)
      if (!"gene" %in% colnames(mods_df) && "id" %in% colnames(mods_df)) mods_df$gene <- mods_df$id
      mods_df  <- mods_df[, c("gene","module")]
      mods_df$gene   <- as.character(mods_df$gene)
      mods_df$module <- as.character(mods_df$module)
      
      g_keep <- intersect(mods_df$gene, rownames(cds_fin))
      if (length(g_keep) >= 2) {
        lc <- SummarizedExperiment::assay(cds_fin[g_keep, ], "logcounts"); lc <- as.matrix(lc)
        
        # MODULE × BIN mean heatmap
        df_cells <- data.frame(cell = colnames(cds_fin), bin = as.integer(pt_bins))
        gene_bin_means <- lapply(split(seq_len(ncol(lc)), df_cells$bin), function(cols_idx) {
          rowMeans(lc[, cols_idx, drop = FALSE], na.rm = TRUE)
        })
        gene_bin_mat <- do.call(cbind, gene_bin_means)  # genes x bins
        colnames(gene_bin_mat) <- as.character(sort(unique(df_cells$bin)))
        
        gb_df <- as.data.frame(gene_bin_mat); gb_df$gene <- rownames(gb_df)
        gb_long <- tidyr::pivot_longer(gb_df, cols = -gene, names_to = "bin", values_to = "expr")
        gb_long <- dplyr::left_join(gb_long, mods_df, by = "gene")
        gb_long$bin <- as.integer(gb_long$bin)
        
        mod_bin <- gb_long |>
          dplyr::group_by(module, bin) |>
          dplyr::summarise(expr = mean(expr, na.rm = TRUE), .groups = "drop")
        
        p_heat_mod <- ggplot2::ggplot(mod_bin, ggplot2::aes(x = bin, y = module, fill = expr)) +
          ggplot2::geom_tile() +
          ggplot2::scale_fill_gradientn(colours = c("#313695", "#ffffbf", "#a50026"), na.value = "grey90") +
          ggplot2::labs(x = "Pseudotime bin", y = "Module", fill = "mean expr",
                        title = "Monocle3 gene programs (modules) along pseudotime") +
          ggplot2::theme_bw(base_size = 10)
        ggplot2::ggsave(file.path(MonocleFig, "monocle3_programs_pseudotime_heatmap_modules.png"),
                        p_heat_mod, width = 8, height = 10, dpi = 400)
        print(p_heat_mod)
        
        # GENE × BIN z-scored, faceted by module
        gb_long_z <- gb_long |>
          dplyr::group_by(gene) |>
          dplyr::mutate(z = {
            mu <- mean(expr, na.rm = TRUE); sdv <- sd(expr, na.rm = TRUE)
            ifelse(is.finite(sdv) & sdv > 0, (expr - mu)/sdv, 0)
          }) |>
          dplyr::ungroup()
        
        p_heat_gene <- ggplot2::ggplot(gb_long_z, ggplot2::aes(x = bin, y = gene, fill = z)) +
          ggplot2::geom_tile() +
          ggplot2::facet_wrap(~ module, scales = "free_y", ncol = 3) +
          ggplot2::scale_fill_gradientn(colours = c("#313695", "#ffffbf", "#a50026"), na.value = "grey90") +
          ggplot2::labs(x = "Pseudotime bin", y = "Genes", fill = "z",
                        title = "Monocle3 gene programs (genes) along pseudotime") +
          ggplot2::theme_bw(base_size = 9) +
          ggplot2::theme(axis.text.y = ggplot2::element_blank(),
                         axis.ticks.y = ggplot2::element_blank())
        ggplot2::ggsave(file.path(MonocleFig, "monocle3_programs_pseudotime_heatmap_genes.png"),
                        p_heat_gene, width = 10, height = 8, dpi = 400)
        print(p_heat_gene)
      }
    }
  } else {
    warning("No gene modules were produced (mods is NULL or empty).")
  }
}

# --- Save cell-level pseudotime table from Monocle3 ---
readr::write_tsv(
  data.frame(
    cell = colnames(cds),
    pseudotime = monocle3::pseudotime(cds),
    cluster = cds@colData$seurat_clusters,
    patient = cds@colData$patientid
  ),
  # file.path(resdir, "02n_m3_celllevel_pseudotime.tsv")
  file.path(MonocleTbl, "monocle3_celllevel_pseudotime.tsv")
)

# --- Seurat UMAP with Monocle3 pseudotime (NA-safe) ---
use_red <- if ("umap.rpca" %in% names(obj_nonmal@reductions)) "umap.rpca" else "umap"

m3 <- obj_nonmal$monocle3_pseudotime
finite_cells <- colnames(obj_nonmal)[is.finite(m3)]

cat(sprintf("[Monocle3 PT] finite=%d / total=%d\n", length(finite_cells), ncol(obj_nonmal)))
if (length(finite_cells) == 0L) {
  stop("Monocle3 pseudotime has no finite values. Re-check order_cells() and root_cells.")
}

# robust cutoffs
q <- suppressWarnings(quantile(m3[finite_cells], c(0.05, 0.95), na.rm = TRUE))
if (!all(is.finite(q))) {
  r <- range(m3[finite_cells], na.rm = TRUE)
  if (!all(is.finite(r))) stop("Monocle3 pseudotime range is not finite.")
  if (r[1] == r[2]) r[2] <- r[2] + 1e-6
  q <- r
}

p_m3_umap_only <- FeaturePlot(
  obj_nonmal,
  features   = "monocle3_pseudotime",
  reduction  = use_red,
  cols       = viridis::viridis(256),
  cells      = finite_cells,           # <— only plot finite PT cells
  min.cutoff = unname(q[1]),
  max.cutoff = unname(q[2]),
  order      = TRUE
) + ggtitle("Monocle3 pseudotime") + theme_pub

# save_plot(p_m3_umap_only, "02o_monocle3_pseudotime_umap")
save_plot(p_m3_umap_only, "monocle3_pseudotime_umap", dir = MonocleFig)
p_m3_umap_only

```


```{r}
#| label: Concordance
#| echo: false
#| message: true
#| warning: true

comp_df <- data.frame(
  CytoTRACE = rank01(obj_nonmal$CytoTRACE),
  slingshot = rank01(obj_nonmal$slingshot_pseudotime),
  monocle3  = if ("monocle3_pseudotime" %in% colnames(obj_nonmal@meta.data))
    rank01(obj_nonmal$monocle3_pseudotime) else NA_real_
) |> na.omit()


cors <- cor(comp_df, method = "spearman")
data.table::fwrite(as.data.frame(cors), file.path(PseudotimeTbl, "pseudotime_spearman_correlations.csv"))

ht <- Heatmap(cors, name = "ρ", cluster_rows = FALSE, cluster_columns = FALSE,
              col = colorRamp2(c(0, 0.5, 1), c("#f0f0f0","#74add1","#08306b")))
pdf(file.path(PseudotimeFig, "pseudotime_correlation_heatmap.pdf"), width = 2.8, height = 2.4); draw(ht); dev.off()
ht

saveRDS(
  list(
    comp_df = comp_df,
    cors = cors
  ),
  file.path(PseudotimeDir, "concordance_inputs.rds")
)

#####
# ================================
# Repair: realign PT tracks by cell name & replot panel
# ================================

message("=== Diagnose finite counts before repair ===")
features <- c("CytoTRACE","slingshot_pseudotime","monocle3_pseudotime")
# ================================
# NA-safe UMAP panel WITHOUT FeaturePlot (robust alignment)
# ================================

use_red  <- if ("umap.rpca" %in% names(obj_nonmal@reductions)) "umap.rpca" else "umap"
umap_mat <- Embeddings(obj_nonmal, use_red)
stopifnot(!is.null(umap_mat))
cell_ids <- rownames(umap_mat)

# If not already defined above in your script, include the same normalizer used elsewhere
normalize_barcode <- function(x) {
  x <- as.character(x)
  x <- sub("\\s+$", "", x)
  x <- sub("^\\s+", "", x)
  x <- sub("-[0-9]+$", "", x)
  x <- sub("\\s.*$", "", x)
  x <- sub(":.*$", "", x)
  x
}

# Helper: pull one meta column aligned to UMAP rows (order-first, then barcode map)
extract_meta_aligned <- function(obj, feature) {
  if (!(feature %in% colnames(obj@meta.data))) return(rep(NA_real_, nrow(obj@meta.data)))
  
  # Fast path: if UMAP rows == meta rownames, just index by row order
  if (all(cell_ids %in% rownames(obj@meta.data))) {
    v <- obj@meta.data[cell_ids, feature, drop = TRUE]
  } else {
    # Fallback: barcode-normalized matching (handles -1 suffixes etc.)
    meta_ids_norm  <- normalize_barcode(rownames(obj@meta.data))
    umap_ids_norm  <- normalize_barcode(cell_ids)
    map_idx        <- match(umap_ids_norm, meta_ids_norm)
    v              <- obj@meta.data[[feature]]
    v              <- v[map_idx]
  }
  
  # Coerce to clean numeric (your helper)
  coerce_one_meta <- function(v) {
    if (is.list(v)) {
      v <- vapply(v, function(x) suppressWarnings(as.numeric(if (length(x)) x[[1]] else NA)), numeric(1))
    } else if (is.factor(v)) {
      v <- suppressWarnings(as.numeric(as.character(v)))
    } else {
      v <- suppressWarnings(as.numeric(v))
    }
    v[!is.finite(v)] <- NA_real_
    v
  }
  
  coerce_one_meta(v)
}

# Build one UMAP colored by a numeric meta vector (already aligned)
build_umap_panel <- function(emb, vec, title) {
  df <- data.frame(x = emb[, 1], y = emb[, 2], val = vec)
  df <- df[is.finite(df$val), , drop = FALSE]
  if (nrow(df) == 0L) return(NULL)
  
  qs <- stats::quantile(df$val, c(0.05, 0.95), na.rm = TRUE)
  if (!all(is.finite(qs)) || qs[1] == qs[2]) {
    rng <- range(df$val, na.rm = TRUE)
    if (rng[1] == rng[2]) rng[2] <- rng[2] + 1e-6
    qs <- rng
  }
  
  ggplot(df, aes(x, y, color = val)) +
    geom_point(size = 0.4) +
    scale_color_viridis_c(limits = c(unname(qs[1]), unname(qs[2])), oob = scales::squish) +
    coord_equal() +
    labs(x = paste0(toupper(use_red), "_1"),
         y = paste0(toupper(use_red), "_2"),
         color = "",
         title = title) +
    theme_minimal(base_size = 12) +
    theme(panel.grid = element_blank())
}

# Extract aligned vectors
v_ct <- if ("CytoTRACE" %in% colnames(obj_nonmal@meta.data))
  extract_meta_aligned(obj_nonmal, "CytoTRACE") else NULL
v_sl <- if ("slingshot_pseudotime" %in% colnames(obj_nonmal@meta.data))
  extract_meta_aligned(obj_nonmal, "slingshot_pseudotime") else NULL
v_m3 <- if ("monocle3_pseudotime" %in% colnames(obj_nonmal@meta.data))
  extract_meta_aligned(obj_nonmal, "monocle3_pseudotime") else NULL

# Build plots
p_ct <- if (!is.null(v_ct)) build_umap_panel(umap_mat, v_ct, "CytoTRACE2") else NULL
p_sl <- if (!is.null(v_sl)) build_umap_panel(umap_mat, v_sl, "Slingshot")  else NULL
p_m3 <- if (!is.null(v_m3)) build_umap_panel(umap_mat, v_m3, "Monocle3")   else NULL

plots <- Filter(Negate(is.null), list(p_ct, p_sl, p_m3))

# Extra diagnostics in case something is still off
diag_counts <- c(
  CytoTRACE = if (is.null(v_ct)) 0 else sum(is.finite(v_ct)),
  Slingshot = if (is.null(v_sl)) 0 else sum(is.finite(v_sl)),
  Monocle3  = if (is.null(v_m3)) 0 else sum(is.finite(v_m3))
)
message(sprintf("[UMAP panel] finite counts — CT:%s  SL:%s  M3:%s",
                diag_counts[1], diag_counts[2], diag_counts[3]))

if (length(plots)) {
  panel_pt <- patchwork::wrap_plots(plots, nrow = 1)
  save_plot(panel_pt, "umap_pseudotime_panel", dir = PseudotimeFig, w = 4 * length(plots), h = 4)
  print(panel_pt)
} else {
  message("No finite panels to plot after robust alignment. Check barcode compatibility across objects.")
  message("Try: head(colnames(obj_nonmal)); head(colnames(cds)); colnames(slingPseudotime(sce)).")
}


```

```{r}
#| label: 04-saving-output-files
#| echo: false
#| message: true
#| warning: true

# saveRDS(obj_nonmal, file.path(MonocleDir, "04a_obj_nonmal.rds"))
saveRDS(sce,        file.path(SlingshotDir, "sce_slingshot.rds"))
saveRDS(cds,        file.path(MonocleDir, "cds_monocle3.rds"))
saveRDS(obj_nonmal, file.path(PseudotimeDir, "final_obj_nonmal_with_pseudotime.rds"))

message("Pipeline completed. | Results in ", normalizePath(data))

```


# Do the three methods agree?

Three pseudotime estimates were computed — CytoTRACE (differentiation potency),
Slingshot (lineage curves through cluster centroids) and Monocle3 (principal
graph). They rest on different assumptions, so **their agreement is the evidence
that any of them is measuring biology rather than an artefact of its own model.**
That correlation matrix was written to disk and never reported.

```{r traj-report-prepare}
#| code-fold: true
suppressPackageStartupMessages({library(data.table); library(ggplot2)})

# The correlation matrix written by the concordance block.
cor_f <- list.files("data", pattern = "pseudotime_spearman_correlations\\.csv$",
                    recursive = TRUE, full.names = TRUE)
cmat <- NULL
if (length(cor_f)) {
  cm <- tryCatch(as.matrix(read.csv(cor_f[1], check.names = FALSE)), error = function(e) NULL)
  if (!is.null(cm) && nrow(cm) == ncol(cm)) {
    rownames(cm) <- colnames(cm)
    cmat <- cm
  }
}

# Per-cell Monocle3 pseudotime, to measure how much of the object it reached.
pt_f <- list.files("data", pattern = "monocle3_celllevel_pseudotime\\.tsv$",
                   recursive = TRUE, full.names = TRUE)
pt <- if (length(pt_f)) tryCatch(fread(pt_f[1]), error = function(e) NULL) else NULL

# Genes varying along the graph.
gt_f <- list.files("data", pattern = "monocle3_graph_test_all\\.csv$",
                   recursive = TRUE, full.names = TRUE)
gt <- if (length(gt_f)) tryCatch(fread(gt_f[1]), error = function(e) NULL) else NULL
cat("correlation matrix:", !is.null(cmat), "| pseudotime rows:",
    if (is.null(pt)) 0 else nrow(pt), "| graph-test genes:", if (is.null(gt)) 0 else nrow(gt), "\n")
```

```{r traj-concordance-finding, results = 'asis'}
if (!is.null(cmat) && nrow(cmat) >= 2) {
  # Pseudotime DIRECTION is arbitrary -- it depends which cell was chosen as the
  # root -- so a negative correlation can simply mean opposite roots. What matters
  # is the MAGNITUDE: |r| near zero means the methods disagree about the ordering
  # itself, which no choice of root can fix.
  pairs <- which(upper.tri(cmat), arr.ind = TRUE)
  pr <- data.table(a = rownames(cmat)[pairs[, 1]],
                   b = colnames(cmat)[pairs[, 2]],
                   r = cmat[upper.tri(cmat)])
  pr[, absr := abs(r)]
  setorder(pr, -absr)
  worst <- pr[.N]; best <- pr[1]

  verdict <- if (best$absr >= 0.7) "the methods broadly agree on the ordering"
             else if (best$absr >= 0.4) "the methods agree only moderately"
             else "**the methods do not agree on the ordering of cells at all**"

  scratch_finding(sprintf(paste(
    "Pairwise Spearman correlation between the pseudotime estimates: %s.",
    "Strongest agreement is %s vs %s at |r| = %.3f; weakest is %s vs %s at",
    "|r| = %.3f. %s.%s"),
    paste(sprintf("%s/%s %.3f", pr$a, pr$b, pr$r), collapse = "; "),
    best$a, best$b, best$absr, worst$a, worst$b, worst$absr, verdict,
    if (best$absr < 0.4)
      paste(" Direction is arbitrary (it follows the root cell), so the sign can be",
            "ignored — but a magnitude this low means the three models are ordering",
            "cells differently, not merely in opposite directions. A trajectory that",
            "only one method supports is a property of that method. Treat any",
            "pseudotime-dependent result below as unvalidated.")
    else ""),
    type = if (best$absr < 0.4) "caution" else "note",
    title = "Pseudotime method concordance")
}
```

```{r}
#| label: fig-traj-concordance
#| fig-cap: "Agreement between pseudotime methods"
#| fig-width: 6.5
#| fig-height: 5
if (!is.null(cmat) && nrow(cmat) >= 2) {
  d <- as.data.table(as.table(cmat)); setnames(d, c("method_a", "method_b", "r"))
  ggplot(d, aes(x = method_a, y = method_b, fill = r,
                text = sprintf("%s vs %s\nSpearman r = %.3f", method_a, method_b, r))) +
    geom_tile(colour = "white", linewidth = 1) +
    geom_text(aes(label = sprintf("%.2f", r)), colour = "grey15", size = 4) +
    scale_fill_gradient2(low = "#8c1d18", mid = "#f2f2f2", high = "#1f4e79",
                         midpoint = 0, limits = c(-1, 1), name = "Spearman r") +
    labs(title = "Do the pseudotime methods agree?",
         subtitle = "Near zero means the methods order cells differently; sign follows the arbitrary root choice",
         x = NULL, y = NULL) +
    theme_scratch() + theme(panel.grid = element_blank())
}
```

```{r traj-coverage-finding, results = 'asis'}
if (!is.null(pt) && "pseudotime" %in% names(pt)) {
  v <- suppressWarnings(as.numeric(pt$pseudotime))
  n_inf <- sum(!is.finite(v)); n_fin <- sum(is.finite(v)); n <- length(v)

  if (n_inf > 0) {
    scratch_finding(sprintf(paste(
      "Monocle3 assigned a finite pseudotime to **%s of %s cells (%.1f%%)**;",
      "the remaining **%s (%.1f%%) are infinite**. In Monocle3 an infinite",
      "pseudotime means the cell sits in a partition with no path to the root, so",
      "the trajectory never reached it.%s"),
      format(n_fin, big.mark = ","), format(n, big.mark = ","), 100 * n_fin / n,
      format(n_inf, big.mark = ","), 100 * n_inf / n,
      if (n_inf > 0.5 * n)
        paste(" **Most of the object is not on this trajectory.** The lineage",
              "describes a minority of cells; any statement about 'the trajectory of",
              "this compartment' applies to that minority only. A disconnected graph",
              "usually means the populations are genuinely distinct rather than",
              "continuous — worth asking whether a trajectory is the right model here.")
      else ""),
      type = if (n_inf > 0.5 * n) "caution" else "note",
      title = "Trajectory coverage")
  }
}
```

```{r}
#| label: fig-traj-coverage
#| fig-cap: "Pseudotime distribution among cells the trajectory reached"
#| fig-width: 9
#| fig-height: 4.5
if (!is.null(pt) && "pseudotime" %in% names(pt)) {
  d <- copy(pt)
  d[, pseudotime := suppressWarnings(as.numeric(pseudotime))]
  fin <- d[is.finite(pseudotime)]
  if (nrow(fin)) {
    grp <- intersect(c("cluster", "patient"), names(fin))[1]
    if (is.na(grp)) { fin[, .grp := "all cells"]; grp <- ".grp" }
    fin[, g := as.character(get(grp))]
    ggplot(fin, aes(x = pseudotime, fill = g,
                    text = sprintf("%s\npseudotime %.2f", g, pseudotime))) +
      geom_histogram(bins = 50, colour = NA) +
      scale_fill_scratch(name = grp) +
      labs(title = "Where the reachable cells sit along the trajectory",
           subtitle = sprintf("%s of %s cells; the rest have infinite pseudotime and are not shown",
                              format(nrow(fin), big.mark = ","), format(nrow(d), big.mark = ",")),
           x = "pseudotime", y = "cells") +
      theme_scratch()
  }
}
```

```{r traj-genes-finding, results = 'asis'}
if (!is.null(gt) && all(c("q_value", "morans_I") %in% names(gt))) {
  sig <- gt[!is.na(q_value) & q_value < 0.05]
  top <- sig[order(-morans_I)][1:min(8, nrow(sig))]

  # Moran's I near 0 means no spatial autocorrelation along the graph: the gene is
  # not actually structured by pseudotime even if its q-value is small, which is
  # easy with tens of thousands of cells.
  strong <- sig[morans_I >= 0.25]

  scratch_finding(sprintf(paste(
    "**%s of %s genes** vary significantly along the graph (q < 0.05), but only",
    "**%s** have Moran's I >= 0.25 — with this many cells a small q-value is easy",
    "to reach, and Moran's I is the measure of whether the gene is actually",
    "structured along the trajectory. Strongest: %s."),
    format(nrow(sig), big.mark = ","), format(nrow(gt), big.mark = ","),
    format(nrow(strong), big.mark = ","),
    paste(sprintf("**%s** (I=%.2f)", top$gene_id, top$morans_I), collapse = ", ")),
    type = "note", title = "Genes varying along the trajectory")
}
```

```{r traj-genes-table}
if (!is.null(gt) && all(c("q_value", "morans_I") %in% names(gt))) {
  sig <- gt[!is.na(q_value) & q_value < 0.05][order(-morans_I)]
  keep <- intersect(c("gene_id", "morans_I", "morans_test_statistic", "p_value", "q_value"),
                    names(sig))
  out <- sig[1:min(200, nrow(sig)), ..keep]
  num <- setdiff(keep, "gene_id")
  out[, (num) := lapply(.SD, function(x) signif(as.numeric(x), 4)), .SDcols = num]
  scratch_table(out, caption = "Genes varying along the Monocle3 graph (top 200 by Moran's I)")
}
```
