---
title: "Module - META PROGRAM & DIFFERENTIAL EXPRESSION"
author: "Syed Shujaat Ali Zaidi"
execute:
  freeze: auto
  cache: false
  warning: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
  match: auto
  effect: zoom
  loop: true
params:
  # paths & context
  seurat_object: null          # path to input .RDS (Seurat object)
  project_name: "project"      # used in filenames where relevant
  work_directory: "."          # Nextflow passes $PWD

  # epithelial subsetting
  subset_col: "azimuth_labels"
  subset_values: "Epithelial"  # comma/semicolon/pipe-separated ok
  min_cells_per_sample_epith: 150
  use_leiden: false

  # QC / Leiden / signatures
  complexity_filter: 1000
  mito_filter: 20
  hvg_keep: 7000
  dims_use_max: 50
  resolution: 1.0
  genes_per_program: 50
  sig_th: 0.005

  # MP clustering
  mp_num: 13
  mp_min_size: 4
  mp_max_size_frac: 0.5
  seed: 1234
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```


```{r setup}
suppressPackageStartupMessages({
  library(Seurat)
  library(Matrix)
  library(dplyr)
  library(ggplot2)
  library(data.table)
  library(stringr)
  library(reshape2)
  library(colorspace)
  library(scalop)     
  library(patchwork) 
  library(future)
  library(future.apply)
})

set.seed(params$seed)

# If Nextflow provided a working directory, use it
if (!is.null(params$work_directory) && dir.exists(params$work_directory)) {
  setwd(params$work_directory)
}

# Local output layout (picked up by Nextflow publishDir)
dir.create("figures/metaprog", recursive = TRUE, showWarnings = FALSE)
dir.create("data",             recursive = TRUE, showWarnings = FALSE)
dir.create("fig",              recursive = TRUE, showWarnings = FALSE)  
dir.create("orig",             recursive = TRUE, showWarnings = FALSE)

# ---- Parallelism / threads ----
guess_threads <- function() {
  v <- Sys.getenv("OMP_NUM_THREADS", unset = NA)
  if (is.na(v) || !nzchar(v)) v <- Sys.getenv("R_PARALLEL_NUM_THREADS", unset = NA)
  n <- suppressWarnings(as.integer(v))
  if (is.na(n) || n < 1) n <- parallel::detectCores()
  if (is.na(n) || n < 1) n <- 1L
  n
}
n_workers <- guess_threads()

ok <- TRUE
tryCatch({
  if (.Platform$OS.type == "unix") {
    future::plan(future::multicore, workers = n_workers)
  } else {
    future::plan(future::multisession, workers = n_workers)
  }
}, error = function(e) { ok <<- FALSE })
if (!ok) future::plan(future::multisession, workers = max(1L, n_workers))

options(future.globals.maxSize = 8 * 1024^3)

```

```{r Load object & epithelial subset}
#| echo: false
stopifnot(!is.null(params$seurat_object))
seu <- readRDS(params$seurat_object)

assay    <- "RNA"
split_by <- "orig.ident"

stopifnot(split_by %in% colnames(seu@meta.data))
stopifnot(assay %in% names(seu@assays))
DefaultAssay(seu) <- assay

# Drop precomputed graphs (avoids Seurat v5 subset()->as.Graph() dimname errors)
if (!is.null(seu@graphs) && length(seu@graphs) > 0) seu@graphs <- list()

# Subset to epithelial cells
stopifnot(params$subset_col %in% colnames(seu@meta.data))
vals <- seu@meta.data[[params$subset_col]]

# Allow comma/semicolon/pipe separated input
epi_vals <- unlist(strsplit(params$subset_values, "[,;|]")) |>
  stringr::str_trim()
epi_vals <- epi_vals[nzchar(epi_vals)]

keep <- rownames(seu@meta.data)[ vals %in% epi_vals ]
if (!length(keep))
  stop("No cells matched ", params$subset_col, " ∈ {", paste(epi_vals, collapse=", "), "}")

seu_sub <- subset(seu, cells = keep)
message("Retained ", ncol(seu_sub), " epithelial cells")

```

```{r Gate samples by epithelial size}
#| echo: false
stopifnot(split_by %in% colnames(seu_sub@meta.data))
epi_counts <- sort(table(seu_sub@meta.data[[split_by]]), decreasing = TRUE)

write.table(
  data.frame(sample = names(epi_counts), epithelial_cells = as.integer(epi_counts)),
  file = file.path("data", "epithelial_counts_per_sample.tsv"),
  sep = "\t", quote = FALSE, row.names = FALSE
)

keep_samples <- names(epi_counts)[epi_counts >= params$min_cells_per_sample_epith]
message(sprintf("[Gate] Keeping %d/%d samples (≥ %d epithelial cells)",
                length(keep_samples), length(epi_counts), params$min_cells_per_sample_epith))
if (!length(keep_samples))
  stop("No samples passed the epithelial cell gate; nothing to process.")

```

```{r Per-sample Leiden → markers → per-cluster top genes}
#| echo: false
genes_list   <- list()   # per-sample/cluster top genes
spots_assign <- list()   # per-sample cell→cluster
markers_all  <- list()   # per-sample FindAllMarkers

for (i in seq_along(keep_samples)) {
  s <- keep_samples[i]
  message(sprintf("[Leiden] %d/%d: %s", i, length(keep_samples), s))

  # subset *from epithelial object*
  sobj <- subset(seu_sub, subset = !!as.name(split_by) == s)
  DefaultAssay(sobj) <- assay

  # QC fields
  nCount_col <- paste0("nCount_", assay)
  if (!nCount_col %in% colnames(sobj@meta.data)) {
    sobj@meta.data[[nCount_col]] <- Matrix::colSums(
      GetAssayData(sobj, assay = assay, layer = "counts")
    )
  }
  if (!"percent.mt" %in% colnames(sobj@meta.data)) {
    sobj[["percent.mt"]] <- PercentageFeatureSet(sobj, pattern = "^MT-")
  }

  # QC filter -> explicit subset
  qc_keep <- rownames(sobj@meta.data)[
    sobj@meta.data[[nCount_col]] > params$complexity_filter &
      sobj@meta.data[["percent.mt"]] < params$mito_filter
  ]
  sobj <- subset(sobj, cells = qc_keep)

  # guard tiny samples
  if (ncol(sobj) < 50) {
    warning(sprintf("Sample %s has <50 post-QC epithelial cells; skipping.", s))
    next
  }

  # Normalize -> HVGs -> Scale
  sobj <- NormalizeData(sobj, assay = assay, normalization.method = "LogNormalize",
                        scale.factor = 10000, verbose = FALSE)
  sobj <- FindVariableFeatures(sobj, selection.method = "vst",
                               nfeatures = min(params$hvg_keep, nrow(sobj)),
                               verbose   = FALSE)
  if (length(VariableFeatures(sobj)) < 200) {
    warning(sprintf("Sample %s has <200 HVGs after filtering; skipping.", s))
    next
  }
  sobj <- ScaleData(sobj, verbose = FALSE)

  # PCA (dynamic dims)
  sobj <- RunPCA(sobj, features = VariableFeatures(sobj), verbose = FALSE)
  npcs_avail <- ncol(Embeddings(sobj, "pca"))
  if (is.null(npcs_avail) || npcs_avail < 2) {
    warning(sprintf("Sample %s: <2 PCs; skipping.", s)); next
  }
  dims_use <- seq_len(min(params$dims_use_max, npcs_avail))

  # Graph & clusters (Leiden if R 'leiden' available, else Louvain)
  sobj <- FindNeighbors(sobj, dims = dims_use, verbose = FALSE)
  LEIDEN_OK <- isTRUE(params$use_leiden) &&
               requireNamespace("reticulate", quietly = TRUE) &&
               tryCatch(reticulate::py_module_available("leidenalg"), error = function(e) FALSE)
  
  algo <- if (LEIDEN_OK) 4L else 3L
  if (!LEIDEN_OK)
    message("Using Louvain (algorithm=3). Set params$use_leiden: true AND have python 'leidenalg' to use Leiden.")

  sobj <- FindClusters(sobj, algorithm = algo, resolution = params$resolution, verbose = FALSE)

  # Markers per cluster (positive only)
  set.seed(params$seed + i)
  mk <- tryCatch(
    FindAllMarkers(sobj, only.pos = TRUE, return.thresh = params$sig_th, verbose = FALSE),
    error = function(e) {
      warning(sprintf("FindAllMarkers failed for sample %s: %s", s, e$message))
      NULL
    }
  )
  if (is.null(mk) || !nrow(mk)) {
    warning(sprintf("No markers returned for sample %s; skipping.", s))
    next
  }
  mk$sample <- s
  markers_all[[s]] <- mk

  # per cluster top N by avg_log2FC
  clids <- sort(unique(mk$cluster))
  for (k in clids) {
    mk_k <- mk[mk$cluster == k, , drop = FALSE]
    ord  <- order(mk_k$avg_log2FC, decreasing = TRUE)
    topg <- mk_k$gene[ord][seq_len(min(params$genes_per_program, nrow(mk_k)))]
    genes_list[[paste0(s, "_", k)]] <- topg
  }

  # cell→cluster assignment
  spots_assign[[s]] <- data.frame(
    barcodes = colnames(sobj),
    cluster  = sobj$seurat_clusters,
    sample   = s,
    stringsAsFactors = FALSE
  )
}

# Persist tables
if (!length(genes_list))
  stop("No programs collected from any sample (after QC/gating).")

spots_assign_df <- do.call(rbind, spots_assign)
markers_all_df  <- data.table::rbindlist(markers_all, fill = TRUE)

saveRDS(genes_list,      file.path("orig", "leiden_programs_list.rds"))
saveRDS(spots_assign_df, file.path("data", "spot_assignments.rds"))
saveRDS(markers_all_df,  file.path("data", "markers_all.rds"))

write.table(spots_assign_df, file.path("data", "spot_assignments.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)
fwrite(markers_all_df, file.path("data", "markers_all.tsv"))


```

# Cluster markers

`FindAllMarkers` returns MARKERS, which are positive by definition — genes
enriched in a cluster relative to the rest. That is a one-sided quantity, so a
volcano is the wrong instrument for it: the empty left half would imply nothing
was downregulated when in truth it was never computed. The ranked panel below is
the honest equivalent — effect size on x, evidence as colour, detection rate as
size.

```{r mk-report-prepare}
#| code-fold: true
suppressPackageStartupMessages({library(data.table); library(ggplot2)})
mk <- tryCatch(as.data.table(markers_all_df), error = function(e) NULL)
if (!is.null(mk) && !nrow(mk)) mk <- NULL
cat("marker rows:", if (is.null(mk)) 0 else nrow(mk),
    "| clusters:", if (is.null(mk)) 0 else uniqueN(mk$cluster), "\n")
```

```{r}
#| label: fig-markers
#| fig-cap: "Top markers per cluster"
#| fig-width: 12
#| fig-height: 8
if (!is.null(mk) && all(c("cluster", "gene", "avg_log2FC", "p_val_adj") %in% names(mk))) {
  scratch_marker_panel(mk, group = "cluster", gene = "gene",
                       lfc = "avg_log2FC", p = "p_val_adj",
                       pct = if ("pct.1" %in% names(mk)) "pct.1" else NULL,
                       n_top = 8, n_groups = 9) +
    labs(title = "Top markers per cluster",
         subtitle = "Markers are one-sided by construction; ranked by effect size within each cluster")
}
```

```{r mk-finding, results = 'asis'}
if (!is.null(mk) && all(c("cluster", "avg_log2FC", "p_val_adj") %in% names(mk))) {
  n_neg <- sum(mk$avg_log2FC < 0, na.rm = TRUE)
  sig   <- mk[!is.na(p_val_adj) & p_val_adj < 0.05]
  per   <- sig[, .N, by = cluster][order(-N)]

  scratch_finding(sprintf(paste(
    "**%s marker calls** across %d cluster(s); **%s** reach adjusted P < 0.05.",
    "Range %s (%s) to %s (%s) per cluster.%s"),
    format(nrow(mk), big.mark = ","), uniqueN(mk$cluster),
    format(nrow(sig), big.mark = ","),
    format(per$N[1], big.mark = ","), per$cluster[1],
    format(per$N[nrow(per)], big.mark = ","), per$cluster[nrow(per)],
    if (n_neg == 0)
      paste(" Every value is positive, confirming these are one-sided marker calls",
            "(`only.pos`), not a two-sided contrast — they say what distinguishes a",
            "cluster, not what it lost.")
    else ""),
    type = "note", title = "Cluster markers")
}
```

```{r mk-table}
if (!is.null(mk)) {
  keep <- intersect(c("cluster", "gene", "avg_log2FC", "pct.1", "pct.2", "p_val_adj", "sample"),
                    names(mk))
  out <- mk[order(cluster, -avg_log2FC)][, head(.SD, 20), by = cluster][, ..keep]
  num <- intersect(c("avg_log2FC", "pct.1", "pct.2", "p_val_adj"), names(out))
  out[, (num) := lapply(.SD, function(x) signif(as.numeric(x), 4)), .SDcols = num]
  scratch_table(out, caption = "Top 20 markers per cluster")
}
```

```{r}
#| label: Jaccard
#| echo: false
jac_mat <- scalop::Jaccard(genes_list)   # programs × programs
jac_mat <- as.matrix(jac_mat); attr(jac_mat, "class") <- "matrix"

# order programs
if (requireNamespace("seriation", quietly = TRUE)) {
  hc  <- hclust(as.dist(1 - jac_mat), method = "average")
  ord <- seriation::get_order(
    seriation::seriate(as.dist(1 - jac_mat), method = "OLO", control = list(hclust = hc))
  )
} else {
  message("Package 'seriation' not available; using hierarchical clustering order.")
  ord <- hclust(as.dist(1 - jac_mat), method = "average")$order
}
jac_matord <- jac_mat[ord, ord, drop = FALSE]

# cut into mp_num, then enforce group size caps
hc2  <- hclust(as.dist(1 - jac_matord), method = "average")

# cutree() errors outright if k exceeds the number of leaves, so a cohort that
# yields fewer programs than mp_num kills the stage instead of returning the
# programs it did find. Seen for real: 604 malignant cells produced 10 programs
# against the default mp_num = 13. Clamping changes nothing when there are
# enough programs; it only prevents the crash when there are not.
# Clamping k to the NUMBER OF PROGRAMS is not enough: with 11 programs that
# gives 11 groups of one member each, and every one is then discarded by the
# min.size filter below -- the clamp guarantees the failure it was meant to
# prevent. k must also leave each group able to REACH min.size, so cap it at
# floor(n_programs / mp_min_size).
n_prog  <- nrow(jac_matord)
min_sz  <- max(1L, as.integer(params$mp_min_size))
k_cap   <- max(1L, floor(n_prog / min_sz))
k_use   <- min(as.integer(params$mp_num), n_prog, k_cap)
if (k_use < as.integer(params$mp_num))
  message(sprintf(paste0("Only %d programs available and mp_min_size=%d, so cutting ",
                         "into %d metaprogram(s) instead of mp_num=%s. Treat this ",
                         "cohort's metaprograms as under-powered."),
                  n_prog, min_sz, k_use, params$mp_num))
cut  <- cutree(hc2, k = k_use)
groups0 <- split(names(cut), cut)

enforce_group_sizes <- function(groups, sim, min.size = 4L, max.frac = 0.5) {
  sim <- as.matrix(sim)
  if (!isSymmetric(sim)) sim <- (sim + t(sim))/2
  diag(sim) <- 1
  nprog <- ncol(sim)
  max.size <- max(1L, floor(max.frac * nprog))

  queue <- groups
  out <- list()
  while (length(queue)) {
    g <- queue[[1]]; queue <- queue[-1]
    g <- intersect(g, colnames(sim))
    if (length(g) < min.size) next
    if (length(g) <= max.size) { out[[length(out)+1]] <- g; next }
    sub <- sim[g, g, drop = FALSE]
    hc  <- hclust(as.dist(1 - sub), method = "average")
    lab <- cutree(hc, k = 2)
    sp  <- split(g, lab)
    queue <- c(queue, sp)
  }
  # paste0("G", integer(0)) returns "G", NOT character(0) -- so naming an empty
  # `out` fails with "'names' attribute [1] must be the same length as the vector
  # [0]", which says nothing about the actual cause. The real cause is that no
  # group reached min.size, i.e. the cohort produced too few programs to build
  # metaprograms from. Say that instead.
  if (!length(out))
    stop(sprintf(paste0("No metaprogram groups survived the size filter: %d program(s) ",
                        "were cut into %d group(s), none reaching mp_min_size = %d. ",
                        "This cohort has too few malignant cells / programs for ",
                        "metaprogram discovery -- lower mp_min_size, lower mp_num, ",
                        "or supply more malignant cells."),
                 ncol(sim), length(groups), as.integer(min.size)), call. = FALSE)
  names(out) <- paste0("G", seq_along(out))
  out
}
groups <- enforce_group_sizes(groups0, jac_matord,
                              min.size = params$mp_min_size,
                              max.frac = params$mp_max_size_frac)

# group label per program
clusterCut <- unlist(lapply(names(groups), function(i) {
  x <- rep(i, length(groups[[i]])); names(x) <- groups[[i]]; x
}))

# Build MP gene sets: ≥25% frequency in group (min 3), cap 50 genes
MPs <- lapply(groups, function(pr_names) {
  tbl <- sort(table(unlist(genes_list[pr_names])), decreasing = TRUE)
  cutoff <- max(3L, ceiling(0.25 * length(pr_names)))
  names(tbl)[tbl >= cutoff][1:min(50, length(tbl))]
})

# Save MP artifacts
saveRDS(jac_matord, file.path("orig", "leiden_jaccard_matrix_ordered.rds"))
saveRDS(groups,     file.path("orig", "leiden_groups.rds"))
saveRDS(MPs,        file.path("data", "leiden_metaprograms.rds"))
fwrite(data.frame(MP = names(MPs),
                  genes = sapply(MPs, function(v) paste(v, collapse=","))),
       file.path("data", "leiden_metaprograms.csv"))

```

```{r Plots (heatmap + annotated 3-panel)}
#| echo: false
# Plain heatmap (capped scale)
long <- reshape2::melt(jac_matord)
cap  <- 0.35
long$value[long$value >= cap] <- cap
p_heat_only <- ggplot(long, aes(Var1, Var2, fill = value)) +
  geom_raster() +
  scale_fill_gradientn(colours = colorspace::sequential_hcl(9, palette="YlOrRd", rev=TRUE),
                       limits = c(0, cap), name = "Jaccard\nIndex") +
  theme_void() + theme(legend.position = "bottom")
ggsave(file.path("figures", "metaprog", "leiden_jaccard_heatmap.pdf"),
       p_heat_only, width = 9, height = 7)

p_heat_only
# Annotated 3-panel (heatmap + MP bar + sample bar)
jac_plot <- reshape2::melt(jac_matord)
jac_plot$value[jac_plot$value >= 0.35] <- 0.35
p_heat <- ggplot(jac_plot, aes(Var1, Var2, fill = value)) +
  geom_raster() +
  scale_fill_gradientn(colours = colorspace::sequential_hcl(9, palette="YlOrRd", rev=TRUE),
                       limits = c(0, 0.35), name = "Jaccard\nIndex") +
  theme_void() + theme(legend.position = "bottom")

prog_ordered <- colnames(jac_matord)
metaprog_df <- data.frame(
  program = prog_ordered,
  MP      = factor(clusterCut[prog_ordered]),
  sample  = sub("_.*$", "", prog_ordered),
  stringsAsFactors = FALSE
)

ann_bar <- function(df, fill_col, ylab_text) {
  ggplot(df, aes(x = factor(program, levels = prog_ordered), y = 1)) +
    geom_raster(aes(fill = .data[[fill_col]])) +
    labs(x = NULL, y = ylab_text) +
    theme(
      legend.position = "none",
      axis.ticks = element_blank(),
      axis.text  = element_blank(),
      axis.title.x = element_blank(),
      panel.background = element_rect(fill = "white"),
      axis.line = element_blank()
    )
}
p_mp     <- ann_bar(metaprog_df, "MP",     "MP")
p_sample <- ann_bar(metaprog_df, "sample", "sample")

MP_plot <- p_heat / p_mp / p_sample + plot_layout(heights = c(40, 2, 1))
ggsave(file.path("figures", "metaprog", "leiden_jaccard_with_annotations.pdf"),
       MP_plot, width = 9, height = 8)
print(MP_plot)
# Optional: save program metadata
saveRDS(metaprog_df, file.path("data", "leiden_metaprog_df.rds"))
write.table(metaprog_df, file.path("data", "leiden_metaprog_df.tsv"),
            sep = "\t", quote = FALSE, row.names = FALSE)

# ---- Barrier artifact for Nextflow ----
out_rds <- paste0(params$project_name, "_leiden.rds")
barrier <- list(
  done = TRUE,
  n_epith_cells = ncol(seu_sub),
  samples = sort(unique(seu_sub@meta.data[[params$subset_col]])),
  timestamp = Sys.time()
)
saveRDS(barrier, file = out_rds) 

```

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

params

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```


