---
title: "Module - META PROGRAM & DIFFERENTIAL EXPRESSION — PostNMF"
author: "Syed Shujaat Ali Zaidi"
execute:
  freeze: auto
  cache: false
  warning: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
params:
  # Paths
  work_directory: "."    # Nextflow will pass $PWD; we look under data/per_sample_mat/nmf_fit here
  project_name: "project"

  # Clustering / MP construction thresholds (same meanings as your script)
  min_intersect_initial: 12
  min_intersect_cluster: 12
  min_group_size: 4
  intra_min: 25
  intra_max: 12
  inter_min: 8

  # Signature length. Must be declared here because Quarto rejects any -P key
  # that is not present in the params block. Previously the value 50 was
  # hardcoded in several places; it is now a single parameter.
  genes_per_program: 50

  # Which factorisation engine produced the inputs: r | python | both.
  # Recorded for provenance; the reading logic auto-detects either format.
  nmf_engine: "python"

  # Seed (only for reproducible tie-breaks or sampling if any)
  seed: 1234
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup}
suppressPackageStartupMessages({
  library(NMF)
  library(Matrix)
  library(ggplot2)
  library(reshape2)
  library(viridis)     # magma() palette
  library(pheatmap)
  library(grid)        # for legend title hack on pheatmap
})

set.seed(params$seed)

# Respect Nextflow-provided working directory
if (!is.null(params$work_directory) && dir.exists(params$work_directory)) {
  setwd(params$work_directory)
}

# Layout used by previous steps
out_dir  <- file.path("data", "per_sample_mat")
fit_dir  <- file.path(out_dir, "nmf_fit")
plot_dir <- file.path("figures", "metaprog")

dir.create(plot_dir, recursive = TRUE, showWarnings = FALSE)

message("Looking for NMF results in: ", normalizePath(fit_dir, mustWork = FALSE))

# ---------------------------------------------------------------------------
# Two engines can feed this step (see --metaprog_nmf_engine):
#
#   r       <sample>_nmf_fit.rds       an NMF/NMFfitX/NMF.rank object
#   python  <sample>_basis_matrix.csv  genes x programs, cols "<rank>_<comp>"
#
# Everything below this chunk consumes `prog_genes_ls` — a per-sample
# genes x programs matrix — so both engines converge on the same structure here
# and the metaprogram construction that follows is untouched.
# ---------------------------------------------------------------------------
fit_files   <- list.files(fit_dir, pattern = "_nmf_fit\\.rds$",      full.names = TRUE)
basis_files <- list.files(fit_dir, pattern = "_basis_matrix\\.csv$", full.names = TRUE)

if (!length(fit_files) && !length(basis_files)) {
  stop("No NMF results found at: ", fit_dir,
       "\nExpected <sample>_nmf_fit.rds (R engine) or <sample>_basis_matrix.csv ",
       "(python engine). Check --metaprog_nmf_engine and publishDir/staging.")
}

# ---- R engine ----
nmf_results <- list()
if (length(fit_files)) {
  nmf_results <- setNames(lapply(fit_files, readRDS),
                          sub("_nmf_fit\\.rds$", "", basename(fit_files)))
  is_bad <- vapply(nmf_results,
                   function(x) is.null(x) || inherits(x, "error") || inherits(x, "try-error"),
                   logical(1))
  nmf_results <- nmf_results[!is_bad]
  cat("Loaded R NMF fits for ", length(nmf_results), " sample(s): ",
      paste(names(nmf_results), collapse = ", "), "\n", sep = "")
}

# ---- Python engine ----
# Read as a plain gene x program matrix. Column names arrive as "<rank>_<comp>"
# and are re-emitted below as "<sample>.<rank>.<comp>", which is the naming the
# metaprogram clustering relies on to recover a program's sample of origin.
py_basis <- list()
if (length(basis_files)) {
  py_basis <- setNames(lapply(basis_files, function(f) {
    df <- utils::read.csv(f, row.names = 1, check.names = FALSE)
    m  <- as.matrix(df)
    storage.mode(m) <- "double"
    m
  }), sub("_basis_matrix\\.csv$", "", basename(basis_files)))

  py_basis <- py_basis[vapply(py_basis, function(m) !is.null(dim(m)) && ncol(m) > 0L, logical(1))]
  cat("Loaded python NMF basis matrices for ", length(py_basis), " sample(s): ",
      paste(names(py_basis), collapse = ", "), "\n", sep = "")
}

if (!length(nmf_results) && !length(py_basis)) {
  stop("All loaded NMF results were empty or unreadable. Cannot proceed.")
}

# A sample name containing "." would break the `strsplit(prog, '[.]')[[1]][[1]]`
# lookup used later to map a program back to its sample.
bad_ids <- grep("\\.", c(names(nmf_results), names(py_basis)), value = TRUE)
if (length(bad_ids)) {
  stop("Sample id(s) contain a '.', which collides with the ",
       "sample.rank.component program naming: ", paste(bad_ids, collapse = ", "))
}
```

``` {r Extract programs (basis columns) and build top-50 gene signatures}
# Return a named list of basis matrices by rank for a given fit
get_basis_by_rank <- function(fit) {
  if (inherits(fit, "NMFfit")) {
    B <- NMF::basis(fit)
    return(setNames(list(B), as.character(ncol(B))))
  }
  if (inherits(fit, "NMFfitX")) {
    out <- lapply(names(fit), function(rk) NMF::basis(fit[[rk]]))
    names(out) <- names(fit)
    return(out)
  }
  if (inherits(fit, "NMF.rank")) {
    obj <- NULL
    obj <- try(fit@fit,  silent=TRUE); if (inherits(obj, "try-error")) obj <- NULL
    if (is.null(obj)) {
      obj <- try(fit@fits, silent=TRUE); if (inherits(obj, "try-error")) obj <- NULL
    }
    if (is.null(obj) && is.list(fit) && !is.null(fit$fit)) obj <- fit$fit
    if (is.null(obj)) stop("NMF.rank object doesn’t expose per-rank fits via @fit/@fits/$fit")
    out <- lapply(names(obj), function(rk) NMF::basis(obj[[rk]]))
    names(out) <- names(obj)
    return(out)
  }
  stop("Unexpected fit class: ", paste(class(fit), collapse="/"))
}

# Build per-sample basis matrices (genes × components)
prog_genes_ls <- list()
for (s in names(nmf_results)) {
  fit <- nmf_results[[s]]
  br  <- get_basis_by_rank(fit)
  if (!length(br)) next

  mats <- list()
  for (rk in names(br)) {
    B <- br[[rk]]
    if (is.null(dim(B)) || ncol(B) == 0L) next
    if (is.null(rownames(B))) rownames(B) <- paste0("g", seq_len(nrow(B)))
    rownames(B) <- toupper(rownames(B))  # unify gene case
    colnames(B) <- paste0(s, ".", rk, ".", seq_len(ncol(B)))  # sample.rank.component
    mats[[rk]] <- B
  }
  if (length(mats)) prog_genes_ls[[s]] <- do.call(cbind, mats)
}

# ---- Python engine: same structure, read straight from the basis CSV --------
# Columns arrive as "<rank>_<component>"; they are renamed to
# "<sample>.<rank>.<component>" so the downstream clustering, which splits on
# "." to recover the sample, sees an identical layout to the R path.
for (s in names(py_basis)) {
  B <- py_basis[[s]]
  if (is.null(dim(B)) || ncol(B) == 0L) next
  if (is.null(rownames(B))) rownames(B) <- paste0("g", seq_len(nrow(B)))
  rownames(B) <- toupper(rownames(B))

  parts <- strsplit(colnames(B), "_", fixed = TRUE)
  colnames(B) <- vapply(seq_along(parts), function(i) {
    pr <- parts[[i]]
    if (length(pr) >= 2L) {
      # "<rank>_<component>"; component is 0-based in python, 1-based in R
      paste0(s, ".", pr[[1]], ".", as.integer(pr[[2]]) + 1L)
    } else {
      paste0(s, ".", ncol(B), ".", i)
    }
  }, character(1))

  # If both engines ran (--metaprog_nmf_engine both) keep them side by side
  # rather than letting one overwrite the other.
  key <- if (!is.null(prog_genes_ls[[s]])) paste0(s, "_py") else s
  prog_genes_ls[[key]] <- B
}

stopifnot(length(prog_genes_ls) > 0L)

cat("Per-sample program counts:\n")
print(vapply(prog_genes_ls, ncol, integer(1)))

# Top-N genes per program
genes_per_program <- if (!is.null(params$genes_per_program)) as.integer(params$genes_per_program) else 50L

nmf_programs_by_sample <- lapply(
  prog_genes_ls,
  function(M) {
    nN <- min(genes_per_program, nrow(M))
    apply(M, 2, function(col) names(sort(col, decreasing = TRUE))[seq_len(nN)])
  }
) # N × K (char) per sample

nmf_programs_sig <- do.call(cbind, nmf_programs_by_sample)  # N × P
cat("Total programs:", ncol(nmf_programs_sig), "\n")

```

``` {r Robust filtering of programs (your original logic)}
robust_nmf_programs <- function(nmf_programs, intra_min = params$intra_min, intra_max = params$intra_max,
                                inter_filter = TRUE, inter_min = params$inter_min) {

  intra_intersect <- lapply(
    nmf_programs,
    function(z) apply(z, 2, function(x) apply(z, 2, function(y) length(intersect(x, y))))
  )
  intra_intersect_max <- lapply(
    intra_intersect,
    function(x) apply(x, 2, function(y) sort(y, decreasing = TRUE)[2])
  )

  nmf_sel <- lapply(
    names(nmf_programs),
    function(x) nmf_programs[[x]][, intra_intersect_max[[x]] >= intra_min, drop = FALSE]
  )
  names(nmf_sel) <- names(nmf_programs)

  if (all(vapply(nmf_sel, function(a) ncol(a) == 0L, logical(1)))) return(character(0))

  nmf_sel_unlist  <- do.call(cbind, nmf_sel)
  inter_intersect <- apply(nmf_sel_unlist, 2, function(x)
    apply(nmf_sel_unlist, 2, function(y) length(intersect(x, y)))
  )

  final_filter <- character(0)
  for (i in names(nmf_sel)) {
    cols_i     <- grep(paste0("^", i, "\\."), colnames(inter_intersect))
    cols_other <- setdiff(seq_len(ncol(inter_intersect)), cols_i)
    if (!length(cols_i)) next
    a <- inter_intersect[cols_other, cols_i, drop = FALSE]
    if (!nrow(a) || !ncol(a)) next

    b <- sort(apply(a, 2, max), decreasing = TRUE)
    if (inter_filter) b <- b[b >= params$inter_min]
    if (!length(b)) next

    keep <- names(b)[1]
    if (length(b) > 1) {
      for (nm in names(b)[-1]) {
        if (max(inter_intersect[keep, nm]) <= intra_max) keep <- c(keep, nm)
      }
    }
    final_filter <- c(final_filter, keep)
  }
  final_filter
}

nmf_filter_all <- robust_nmf_programs(
  nmf_programs_by_sample,
  intra_min    = params$intra_min,
  intra_max    = params$intra_max,
  inter_filter = TRUE,
  inter_min    = params$inter_min
)

if (length(nmf_filter_all)) {
  nmf_programs_sig <- nmf_programs_sig[, colnames(nmf_programs_sig) %in% nmf_filter_all, drop = FALSE]
  message("Filtered programs kept: ", ncol(nmf_programs_sig))
} else {
  warning("No programs survived robust filtering — proceeding with unfiltered set")
}

# Persist filtered signatures
path <- file.path(fit_dir, "nmf_programs_sig_filtered.rds")
dir.create(dirname(path), recursive = TRUE, showWarnings = FALSE)
saveRDS(nmf_programs_sig, file = path)

```

``` {r jaccard_heatmaps, message=FALSE, warning=FALSE}
# Overlap counts between program gene-sets (genes_per_program-length signatures)
sim_mat <- apply(
  nmf_programs_sig, 2,
  function(a) apply(nmf_programs_sig, 2, function(b) length(intersect(a, b)))
)
saveRDS(sim_mat, file = file.path(fit_dir, "nmf_intersect.rds"))

hc   <- hclust(as.dist(genes_per_program - sim_mat), method = "average")
dend <- reorder(as.dendrogram(hc), colMeans(sim_mat))
ord  <- order.dendrogram(dend)

# Heatmap #1 (ggplot)
long <- reshape2::melt(sim_mat[ord, ord])
pal  <- c(
  colorRampPalette(c("white", rev(magma(323, begin = 0.15))[1]))(10),
  rev(magma(323, begin = 0.18))
)
p_heat <- ggplot(long, aes(Var1, Var2, fill = 100 * value / pmax(1, 100 - value))) +
  geom_tile() +
  scale_fill_gradient2(
    low = pal[1], mid = pal[112], high = pal[223],
    midpoint = 13.5, oob = scales::squish, name = "Jaccard %"
  ) +
  theme_void() + theme(legend.position = "right")
print(p_heat)
ggsave(file.path(plot_dir, "jaccard_heatmap_dendrogram22.pdf"), p_heat, width = 7, height = 6)

# Heatmap #2 (pheatmap) with sample bar
J      <- sim_mat[ord, ord, drop = FALSE]
J_perc <- 100 * J / pmax(1, 100 - J)
J_perc[!is.finite(J_perc)] <- NA

progs   <- colnames(J_perc)
samples <- sub("\\..*$", "", progs)
annot   <- data.frame(Sample = samples, row.names = progs)

sample_levels <- unique(samples)
ann_colors <- list(
  Sample = setNames(scales::hue_pal()(length(sample_levels)), sample_levels)
)

# Aligned with the pipeline-wide sequential ramp (one hue, light->dark)
# so this heatmap sits in the same visual family as every other figure.
custom_magma <- if (exists("scratch_heat_colours")) scratch_heat_colours(256) else c(
  colorRampPalette(c("white", rev(viridis::magma(323, begin = 0.15))[1]))(10),
  rev(viridis::magma(323, begin = 0.18))
)

# pheat2 <- pheatmap::pheatmap(
#   J_perc,
#   cluster_rows   = FALSE,
#   cluster_cols   = FALSE,
#   show_rownames  = FALSE,
#   show_colnames  = FALSE,
#   color          = custom_magma,
#   annotation_col = annot,
#   annotation_row = annot,
#   annotation_colors = ann_colors,
#   legend         = TRUE,
#   border_color   = NA,
#   annotation_names_col = TRUE,
#   annotation_names_row = FALSE
# )
# 
# pdf(file.path(plot_dir, "jaccard_heatmap_dendrogram3.pdf"), width = 7, height = 6)
# grid::grid.newpage()
# grid::grid.draw(pheat2$gtable)
# grid::grid.text("Jaccard index (%)",
#                 x = grid::unit(0.92, "npc"), y = grid::unit(0.95, "npc"),
#                 rot = 0, gp = grid::gpar(cex = 0.8))
# dev.off()
```

``` {r MP list}
min_intersect_initial <- params$min_intersect_initial
min_intersect_cluster <- params$min_intersect_cluster
min_group_size        <- params$min_group_size


nmf_intersect      <- sim_mat
nmf_programs_sig_K <- nmf_programs_sig

# Basis matrices per sample (for tie-breaks); ensure uppercase genes
Genes_nmf_w_basis <- lapply(prog_genes_ls, function(M) {
  rownames(M) <- toupper(rownames(M))
  M
})

Sorted_intersection <- sort(
  apply(nmf_intersect, 2, function(x) length(which(x >= min_intersect_initial)) - 1L),
  decreasing = TRUE
)

Cluster_list <- list()
MP_list      <- list()
k <- 1L

while (length(Sorted_intersection) && Sorted_intersection[1] >= min_group_size) {

  Curr_cluster <- c(names(Sorted_intersection)[1])

  Genes_MP <- nmf_programs_sig_K[, Curr_cluster]
  nmf_programs_sig_K <- nmf_programs_sig_K[, setdiff(colnames(nmf_programs_sig_K), Curr_cluster), drop = FALSE]

  Intersection_with_Genes_MP <- if (ncol(nmf_programs_sig_K)) {
    sort(apply(nmf_programs_sig_K, 2, function(x) length(intersect(Genes_MP, x))), decreasing = TRUE)
  } else numeric(0)

  NMF_history <- Genes_MP

  while (length(Intersection_with_Genes_MP) && Intersection_with_Genes_MP[1] >= min_intersect_cluster) {
    best <- names(Intersection_with_Genes_MP)[1]
    Curr_cluster <- c(Curr_cluster, best)

    Genes_MP_temp   <- sort(table(c(NMF_history, nmf_programs_sig_K[, best])), decreasing = TRUE)
    Genes_at_border <- Genes_MP_temp[which(Genes_MP_temp == Genes_MP_temp[genes_per_program])]

    if (length(Genes_at_border) > 1) {
      Genes_curr_NMF_score <- c()
      for (i_prog in Curr_cluster) {
        curr_sample <- strsplit(i_prog, "[.]")[[1]][[1]]
        B <- Genes_nmf_w_basis[[curr_sample]]
        if (is.null(B) || !nrow(B)) next
        idx <- match(names(Genes_at_border), rownames(B))
        ok  <- which(!is.na(idx))
        if (length(ok)) {
          vec <- B[idx[ok], i_prog, drop = TRUE]
          names(vec) <- rownames(B)[idx[ok]]
          Genes_curr_NMF_score <- c(Genes_curr_NMF_score, vec)
        }
      }
      Genes_curr_NMF_score_sort <- sort(Genes_curr_NMF_score, decreasing = TRUE)
      Genes_curr_NMF_score_sort <- Genes_curr_NMF_score_sort[!duplicated(names(Genes_curr_NMF_score_sort))]
      Genes_MP_temp <- c(
        names(Genes_MP_temp[which(Genes_MP_temp > Genes_MP_temp[genes_per_program])]),
        names(Genes_curr_NMF_score_sort)
      )
    } else {
      Genes_MP_temp <- names(Genes_MP_temp)[1:genes_per_program]
    }

    NMF_history <- c(NMF_history, nmf_programs_sig_K[, best])
    Genes_MP    <- Genes_MP_temp[1:genes_per_program]

    nmf_programs_sig_K       <- nmf_programs_sig_K[, setdiff(colnames(nmf_programs_sig_K), best), drop = FALSE]
    Intersection_with_Genes_MP <- if (ncol(nmf_programs_sig_K)) {
      sort(apply(nmf_programs_sig_K, 2, function(x) length(intersect(Genes_MP, x))), decreasing = TRUE)
    } else numeric(0)
  }

  Cluster_list[[paste0("Cluster_", k)]] <- Curr_cluster
  MP_list[[paste0("MP_", k)]]           <- Genes_MP
  k <- k + 1L

  keep_cols <- setdiff(colnames(nmf_intersect), Curr_cluster)
  nmf_intersect <- nmf_intersect[keep_cols, keep_cols, drop = FALSE]
  Sorted_intersection <- if (ncol(nmf_intersect)) {
    sort(apply(nmf_intersect, 2, function(x) length(which(x >= min_intersect_initial)) - 1L), decreasing = TRUE)
  } else numeric(0)
}

# safety net
if (!length(Cluster_list) && ncol(sim_mat) > 0) {
  seed_prog <- names(sort(colSums(sim_mat >= min_intersect_initial), decreasing = TRUE))[1]
  Cluster_list[[ "Cluster_1" ]] <- seed_prog
  MP_list[[ "MP_1" ]] <- nmf_programs_sig[, seed_prog]
}

# Record HOW the metaprograms were obtained. The fallback above is not a normal
# result: it fires when NOTHING met the clustering criteria and seeds a single
# "metaprogram" from one NMF program. Downstream that is indistinguishable from a
# genuine recurrent programme unless the report says so.
mp_from_fallback <- length(Cluster_list) == 1 &&
                    identical(names(Cluster_list), "Cluster_1") &&
                    length(Cluster_list[["Cluster_1"]]) == 1

saveRDS(Cluster_list, file = file.path(fit_dir, "Cluster_list_final.rds"))
saveRDS(MP_list,      file = file.path(fit_dir, "MP_list_final.rds"))
write.csv(MP_list,    file = file.path(fit_dir, "MP_list_final.csv"))
```

# What the metaprograms are

An NMF metaprogram is only meaningful if it **recurs across samples** — that is
the whole claim. A programme found in one sample is that sample's expression
structure, and a programme made of mitochondrial and ribosomal genes is a
technical axis regardless of how many samples it appears in. Neither check was
reported, so this stage could return a single mitochondrial programme from a
single sample and describe itself as "completed with nothing flagged".

```{r mp-report-prepare}
#| code-fold: true
suppressPackageStartupMessages({library(data.table); library(ggplot2)})

mp_tbl <- NULL
if (length(MP_list)) {
  mp_tbl <- rbindlist(lapply(names(MP_list), function(nm) {
    g <- as.character(MP_list[[nm]])
    g <- g[!is.na(g) & nzchar(g)]
    if (!length(g)) return(NULL)
    data.table(MP = nm, gene = g, rank = seq_along(g))
  }), fill = TRUE)
  if (!is.null(mp_tbl) && !nrow(mp_tbl)) mp_tbl <- NULL
}

# Per-sample cell counts drive the recurrence claim.
sc <- NULL
for (f in c(file.path(dirname(fit_dir), "_sample_cell_counts.tsv"),
            file.path("data", "per_sample_mat", "_sample_cell_counts.tsv"))) {
  if (is.null(sc) && file.exists(f)) sc <- tryCatch(fread(f), error = function(e) NULL)
}

# Gene-class composition: the check that separates biology from technical signal.
classify_gene <- function(g) {
  fifelse(grepl("^MT-",  g), "mitochondrial",
  fifelse(grepl("^RP[LS]", g), "ribosomal",
  fifelse(grepl("^(MALAT1|NEAT1|XIST|TALAM1)$", g) | grepl("^(LINC|ENSG)", g), "lncRNA / unannotated",
          "protein-coding / other")))
}
if (!is.null(mp_tbl)) mp_tbl[, class := classify_gene(toupper(gene))]
cat("metaprograms:", if (is.null(mp_tbl)) 0 else uniqueN(mp_tbl$MP),
    "| genes:", if (is.null(mp_tbl)) 0 else nrow(mp_tbl), "\n")
```

```{r mp-provenance-finding, results = 'asis'}
n_mp <- if (is.null(mp_tbl)) 0 else uniqueN(mp_tbl$MP)
n_samples_used <- if (!is.null(sc)) nrow(sc) else NA_integer_

if (exists("mp_from_fallback") && isTRUE(mp_from_fallback)) {
  scratch_finding(paste(
    "**No metaprogram met the clustering criteria.** The single programme below was",
    "seeded from one NMF program by the fallback path — it is one sample's",
    "expression structure, not a recurrent metaprogram. Treat it as a placeholder:",
    "the recurrence claim that makes a metaprogram meaningful has not been made here."),
    type = "caution", title = "Metaprograms came from the fallback path")
} else if (n_mp > 0) {
  scratch_finding(sprintf("**%d metaprogram(s)** derived by iterative clustering of the per-sample NMF programs.",
                          n_mp), type = "note", title = "Metaprograms")
}
```

```{r mp-sample-finding, results = 'asis'}
if (!is.null(sc) && nrow(sc)) {
  cc <- intersect(c("cells", "n_cells", "epithelial_cells"), names(sc))[1]
  scn <- intersect(c("sample", "sample_id"), names(sc))[1]
  if (!is.na(cc) && !is.na(scn)) {
    d <- sc[order(-get(cc))]
    thr <- tryCatch(as.integer(params$min_cells_per_sample_epith), error = function(e) NA_integer_)
    if (is.na(thr)) thr <- 150L
    kept <- d[get(cc) >= thr]; dropped <- d[get(cc) < thr]

    scratch_finding(sprintf(paste(
      "NMF ran on **%d of %d sample(s)** (%s). %s%s"),
      nrow(kept), nrow(d),
      paste(sprintf("%s: %s cells", kept[[scn]], format(kept[[cc]], big.mark = ",")), collapse = "; "),
      if (nrow(dropped))
        sprintf("**%s contributed too few cells (%s) and was excluded at the %d-cell threshold.** ",
                paste(dropped[[scn]], collapse = ", "),
                paste(format(dropped[[cc]], big.mark = ","), collapse = ", "), thr)
      else "",
      if (nrow(kept) < 2)
        paste("**With one sample there is no recurrence to measure** — a metaprogram is",
              "defined by appearing in several samples independently, so nothing here",
              "distinguishes a shared programme from this sample's own structure.")
      else ""),
      type = if (nrow(kept) < 2) "caution" else "note",
      title = "Samples entering the NMF")
  }
}
```

```{r mp-composition-finding, results = 'asis'}
if (!is.null(mp_tbl)) {
  comp <- mp_tbl[, .N, by = .(MP, class)][, frac := N / sum(N), by = MP]
  tech <- comp[class %in% c("mitochondrial", "ribosomal", "lncRNA / unannotated"),
               .(tech_frac = sum(frac)), by = MP][order(-tech_frac)]
  worst <- tech[1]

  if (nrow(tech) && worst$tech_frac >= 0.3) {
    scratch_finding(sprintf(paste(
      "**%s is %.0f%% mitochondrial, ribosomal or unannotated transcripts.** Those",
      "track cell quality and library composition, not a biological programme —",
      "a metaprogram dominated by them usually means the input was not filtered on",
      "percent-mito, or that the NMF is fitting a depth axis. Check the QC",
      "thresholds before interpreting it as tumour biology."),
      worst$MP, 100 * worst$tech_frac),
      type = "caution", title = "Metaprogram is technical, not biological")
  } else if (nrow(tech)) {
    scratch_finding(sprintf(
      "Highest technical-gene content is **%s** at %.0f%% (mitochondrial, ribosomal or unannotated) — below the 30%% concern threshold.",
      worst$MP, 100 * worst$tech_frac), type = "note", title = "Metaprogram composition")
  }
}
```

```{r}
#| label: fig-mp-composition
#| fig-cap: "Gene-class composition of each metaprogram"
#| fig-width: 9
#| fig-height: 4.5
if (!is.null(mp_tbl)) {
  comp <- mp_tbl[, .N, by = .(MP, class)][, frac := N / sum(N), by = MP]
  ggplot(comp, aes(x = MP, y = frac, fill = class,
                   text = sprintf("%s\n%s: %d genes (%.0f%%)", MP, class, N, 100 * frac))) +
    geom_col(width = .7) +
    coord_flip() +
    scale_fill_scratch(name = NULL) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(title = "What each metaprogram is made of",
         subtitle = "Mitochondrial, ribosomal and unannotated transcripts are technical signal, not programme biology",
         x = NULL, y = "share of genes") +
    theme_scratch()
}
```

```{r mp-genes-table}
if (!is.null(mp_tbl)) {
  scratch_table(mp_tbl[order(MP, rank), .(metaprogram = MP, rank, gene, class)],
                caption = "Metaprogram gene membership")
}
```

``` {r Iterative clustering to derive Meta-Programs (MPs)}

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

params

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```
