---
title: "Module - inferCNV"
author: "SCRATCH"
execute:
  freeze: auto
  cache: false
  warning: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: "Test"
  seurat_object: "data/project_Azimuth_annotation_object.RDS"
  # reference_table: "data/Ovarian_reference_table.csv"
  organism: "human"
  input_analysis_mode: 'samples'
  bayes_max_p_normal: 0
  thr_cluster_size: 1000
  n_threads: 8
  n_memory: 16
  work_directory: !expr here::here()
  auto_save: true
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE, message = FALSE, warning = FALSE, echo = FALSE}

# Project parameters 
project_name <- params$project_name
seurat_object <- params$seurat_object
reference_table <- params$reference_table

# Inputs and thresholds
input_model <- params$input_model
input_analysis_mode <- params$input_analysis_mode
bayes_max_p_normal <- as.numeric(params$bayes_max_p_normal)
thr_cluster_size <- params$thr_cluster_size

# Dataflow/Computational parameters
n_threads <- params$n_threads
n_memory <- params$n_memory

# Output parameters
work_directory <- params$work_directory
auto_save <- params$auto_save

# Changing root directory
knitr::opts_knit$set(
  root.dir = work_directory
)

```

## Loading library

```{r library}

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(infercnv)
library(annotables)
library(SCP)

```

## Folder structure

```{r project_directory}

if(!dir.exists(work_directory)) {
  dir.create(work_directory, recursive = T)
}

for(sub_directory in c('data', 'figures')) {
  dir.create(paste0(work_directory, '/', sub_directory))
}

```

```{r loading seurat object}
# Reading seurat object
seurat_object <- readRDS(file = seurat_object)
# seurat_object <- readRDS(file = 'Documents/SCRATCH-Annotation-dev/data-Final/SCRATCH_ANNOTATION:AZIMUTH_ANNOTATION/data/project_Azimuth_annotation_object.RDS')
# seurat_object <- readRDS(file = 'Documents/SCRATCH-Annotation-dev/data-Final/SCRATCH_ANNOTATION:SCYTPE_AGGREGATE_ANNOTATION/data/project_sctype_agg_annotation_object.RDS')

```

```{r filtering celltypes by tool type}
# Reference = the normal baseline inferCNV measures against. EVERY other cell is
# an observation. The previous version also carried a `cell_types_to_keep` list
# and subset the object to it, which is where this stage died:
#
#   sctype branch kept c("B_Plasma_Cells", "Epithelial", "Fibroblast", "T_Cells")
#   and scType assigns ZERO cells to "Epithelial" -- its labels are T_Cells,
#   B_Plasma_Cells, Myeloid, Unknown, NK_Cells, Fibroblast, Endothelial_Cells.
#
# So all 12,871 surviving cells were reference cells, there were no observations,
# and inferCNV crashed inside .plot_cnv_observations() with
# `obs_dendrogram[[1]]: subscript out of bounds`. The tumour cells were in the
# 4,308 "Unknown" that the keep-list discarded.
#
# Whitelisting observations is the wrong shape anyway: you cannot know in advance
# what the malignant cells were labelled, and that is the entire question. Keep
# every cell, name only the REFERENCE, and let everything else be observed.
if ("azimuth_labels" %in% colnames(seurat_object@meta.data)) {
  reference_label <- "azimuth_labels"
  reference_grp   <- c("B cell", "T cell", "Fibroblast")
} else if ("sctype" %in% colnames(seurat_object@meta.data)) {
  reference_label <- "sctype"
  reference_grp   <- c("B_Plasma_Cells", "T_Cells", "Fibroblast")
} else if ("cell_label" %in% colnames(seurat_object@meta.data)) {
  # Was `else if ("sctype" %in% ...)` -- a copy of the branch above, so this arm
  # was unreachable and a cell_label-only object hit the stop() instead.
  reference_label <- "cell_label"
  reference_grp   <- c("B cell", "T cell", "Fibroblast")
} else {
  stop("No supported annotation column found. The object must have one of the following columns: azimuth_labels, sctype, cell_label")
}

```

```{r loading_rds}



# # Reference population
# reference_table <- readr::read_csv(file = reference_table)
# reference_table <- readr::read_csv(file = 'Documents/SCRATCH-CNV-main/assets/OV_reference_table.csv')
# reference_table <- reference_table %>%
#   tibble::column_to_rownames(var = "barcode") %>%
#   as.data.frame()
# 
# # Add metadata
# seurat_object <- Seurat::AddMetaData(
#   object = seurat_object, 
#   metadata = reference_table
#   )

seurat_object$reference_label <- as.character(seurat_object@meta.data[[reference_label]])

# inferCNV needs a group label for every cell; NA is not a group.
seurat_object$reference_label[is.na(seurat_object$reference_label) |
                              !nzchar(seurat_object$reference_label)] <- "Unassigned"

# No keep-list: every cell is retained. Reference cells form the baseline, all
# others are observations.
.present   <- unique(seurat_object$reference_label)
.ref_found <- intersect(reference_grp, .present)
.n_ref     <- sum(seurat_object$reference_label %in% .ref_found)
.n_obs     <- sum(!seurat_object$reference_label %in% .ref_found)

# Fail with the actual label vocabulary rather than deep inside inferCNV's
# plotting code. Both of these produced `subscript out of bounds` before.
if (!length(.ref_found)) {
  stop("inferCNV: none of the reference groups (", paste(reference_grp, collapse = ", "),
       ") occur in `", reference_label, "`. Labels present: ",
       paste(sort(.present), collapse = ", "),
       ". Without reference cells there is no baseline to call CNV against.")
}
if (.n_obs == 0L) {
  stop("inferCNV: every cell falls in a reference group, so there is nothing to ",
       "observe. Labels present: ", paste(sort(.present), collapse = ", "))
}

message(sprintf(paste("[inferCNV] reference: %s cells across %d group(s) [%s];",
                      "observations: %s cells across %d group(s)"),
                format(.n_ref, big.mark = ","), length(.ref_found),
                paste(.ref_found, collapse = ", "),
                format(.n_obs, big.mark = ","),
                length(setdiff(.present, .ref_found))))

# Reference groups named but absent are dropped silently by inferCNV otherwise.
if (length(setdiff(reference_grp, .ref_found))) {
  warning("inferCNV: reference group(s) not present and therefore unused: ",
          paste(setdiff(reference_grp, .ref_found), collapse = ", "),
          call. = FALSE)
}

## yibo recommendations: add Endothelial  cells and smooth muscle  cellls and parasite to reference, p

# Setting random seed
random_seed <- 022003
set.seed(random_seed)

```

#| label: fig-clustering
#| fig-cap: 'UMAP by reference_label'
#| column: screen-inset-shaded
#| layout-nrow: 1

tryCatch({
  umap_plot <- SCP::CellDimPlot(
    srt = seurat_object, 
    group.by = "reference_label",
    # split.by = "reference_label",
    reduction = "umap",
    label = TRUE, 
    label_repel = TRUE,
    show_stat = TRUE,
    theme_use = "theme_blank"
  )
  ggsave("umap_reference_groups.png", plot = umap_plot, width = 6, height = 5, dpi = 300)
  print(umap_plot)
}, error = function(e) {
  message("⚠️ Skipping UMAP plot due to rendering error: ", e$message)
})
```
## InferCNV Analysis

:::{.callout-tip}
The @fig-clustering shows highly variable genes on the dataset.
:::

```{r infer_cnv_prepare}

seurat_counts <- GetAssayData(
  seurat_object,
  layer = "counts"
  )

```

```{r infercnv_genes}

# Human only
infercnv_genes <- annotables::grch38 %>%
  filter(biotype == "protein_coding") %>%
  filter(chr %in% 1:22) %>%
  arrange(chr, start) %>%
  distinct(symbol, chr, start, end) %>%
  mutate(
    chr = paste0('chr', chr)
  )

infercnv_genes <- infercnv_genes[!duplicated(infercnv_genes$symbol),]
infercnv_genes <- infercnv_genes %>%
  tibble::column_to_rownames(var = "symbol")

```

```{r infercnv_reference}

reference_table <- seurat_object@meta.data %>%
  dplyr::select(reference_label)

```

```{r infer_cnv_run}


infercnv_object <- infercnv::CreateInfercnvObject(
          raw_counts_matrix = seurat_counts,
          annotations_file  = reference_table,
          delim             = "\t",
          gene_order_file   = infercnv_genes,
          # .ref_found, not reference_grp: naming a group with no cells makes
          # inferCNV build an empty reference and fail obscurely later.
          ref_group_names = .ref_found
          )
# ref_group_names   = c("B cell", "T cell", "Fibroblast") -->
# ref_group_names   = c("Fibroblast", "Endothelial", "B cell", "T cell")
```

```{r infer_cnv_run part 2}
# BayesMaxPNormal gates inferCNV's Bayesian mixture step, which re-scores each
# predicted CNV region and drops the ones likely to be normal. On this data it
# aborts inside inferCNV's own MCMC code:
#
#   Error in `do.call()`: second argument must be a list
#     infercnv::run -> inferCNVBayesNet -> runMCMC -> getProbabilities
#       -> do.call(rbind, mcmc[[j]])
#
# getProbabilities assumes every cell group accumulated MCMC draws; a group that
# produced none leaves NULL there, and rbind refuses it. It is a property of the
# grouping, not of the call here, so it cannot be retried around. Setting the
# threshold to 0 skips that step and keeps the HMM's own state calls, which is
# what the downstream `add_to_seurat` and the STRATIFY consensus actually read.
#
# Restore inferCNV's upstream default with -P bayes_max_p_normal:0.5 once the
# image carries a version where the MCMC step tolerates an empty group.
infercnv_object <- infercnv::run(
          infercnv_object,
          cutoff = 0.25,
          out_dir='./data/infercnv',
          cluster_by_groups = TRUE,
          analysis_mode = input_analysis_mode,
          denoise = TRUE,
          HMM = TRUE,
          BayesMaxPNormal = bayes_max_p_normal,
          num_threads = n_threads
          )
```

```{r infercnv_combine_meta}

# Loading infercnv object
seurat_infercnv_object <- seurat_object[, colnames(infercnv_object@expr.data)]
seurat_infercnv_object <- infercnv::add_to_seurat(
    infercnv_output_path = './data/infercnv',
    seurat_obj = seurat_infercnv_object,
    top_n = 10
  )

# Summarizing CNV score
infercnv_summarise_cnv <- seurat_infercnv_object@meta.data %>%
    rowwise() %>% 
    mutate(
      genomwide_cnv_proportion = sum(c_across(starts_with("proportion_scaled_cnv")), na.rm = TRUE),
      n_affected_chromosomes = sum(c_across(starts_with("has_cnv")), na.rm = TRUE)
    )

# 
genomwide_cnv_proportion <- infercnv_summarise_cnv$genomwide_cnv_proportion
n_affected_chromosomes <- infercnv_summarise_cnv$n_affected_chromosomes

# 
seurat_infercnv_object[['genomwide_cnv_proportion']] <- log1p(genomwide_cnv_proportion)
seurat_infercnv_object[['n_affected_chromosomes']] <- n_affected_chromosomes

```

### inferCNV UMAP

```{r infercnv_visualization}
#| label: fig-feat-cnv
#| fig-cap: 'CNV Load'
#| column: screen-inset-shaded
#| layout-nrow: 1

FeaturePlot(seurat_infercnv_object, features = 'genomwide_cnv_proportion') +
  labs(title = "CNV - HMM score")

```

:::{.callout-tip}
The @fig-feat-cnv displays CNV load on the UMAP dimensions
:::

## Copy-number landscape

inferCNV writes far more than the heatmap it is known for: a per-cell,
per-chromosome table of how much of each chromosome is altered and in which
direction, and an HMM segmentation naming the regions themselves. None of it was
reported here — the stage produced one UMAP and no computed statement at all.
The sections below read those outputs directly.

```{r cnv-landscape-prepare}
#| code-fold: true
suppressPackageStartupMessages({library(data.table); library(ggplot2)})

.md <- as.data.table(seurat_infercnv_object@meta.data, keep.rownames = "cell_id")

# Resolve the annotation column with the same candidate order the rest of the
# pipeline uses, so "cell type" means one thing across every stage.
.ct_cand <- c("celltype", "azimuth_labels", "cell_type", "annotation",
              "sctype", "sctype_agg", "cell_label")
.col_ct  <- .ct_cand[.ct_cand %in% names(.md)][1]
.col_smp <- c("sample_id", "orig.ident", "sample", "patient_id")[
              c("sample_id", "orig.ident", "sample", "patient_id") %in% names(.md)][1]

# Long form: one row per cell x chromosome. The wide table carries ~220 columns
# (has_/proportion_/proportion_scaled_ x loss/dupli/cnv x 22 autosomes), which is
# unreadable as a table and unplottable as-is.
.melt_metric <- function(prefix, value_name) {
  cols <- grep(paste0("^", prefix, "_chr[0-9XY]+$"), names(.md), value = TRUE)
  if (!length(cols)) return(NULL)
  d <- melt(.md[, c("cell_id", cols), with = FALSE], id.vars = "cell_id",
            variable.name = "col", value.name = value_name)
  d[, chr := sub(paste0("^", prefix, "_chr"), "", as.character(col))][, col := NULL]
  d
}

.cnv   <- .melt_metric("proportion_cnv",   "prop_cnv")
.loss  <- .melt_metric("proportion_loss",  "prop_loss")
.dupli <- .melt_metric("proportion_dupli", "prop_dupli")

cnv_long <- NULL
if (!is.null(.cnv)) {
  cnv_long <- .cnv
  if (!is.null(.loss))  cnv_long <- merge(cnv_long, .loss,  by = c("cell_id", "chr"), all.x = TRUE)
  if (!is.null(.dupli)) cnv_long <- merge(cnv_long, .dupli, by = c("cell_id", "chr"), all.x = TRUE)
  cnv_long <- merge(cnv_long,
                    .md[, .(cell_id,
                            celltype = if (!is.na(.col_ct)) as.character(get(.col_ct)) else "all cells",
                            sample   = if (!is.na(.col_smp)) as.character(get(.col_smp)) else "all")],
                    by = "cell_id", all.x = TRUE)
  cnv_long[, chr := factor(chr, levels = c(as.character(1:22), "X", "Y"))]
  cnv_long <- cnv_long[!is.na(chr)]
}
cat("Cells:", nrow(.md), "| cell-type column:", .col_ct,
    "| chromosomes:", if (!is.null(cnv_long)) uniqueN(cnv_long$chr) else 0, "\n")
```

### Which populations carry copy-number burden?

```{r fig.width = 10, fig.height = 5.5}
#| label: fig-cnv-burden
#| fig-cap: "Genome-wide CNV burden by annotated cell type"
if (!is.na(.col_ct) && "genomwide_cnv_proportion" %in% names(.md)) {
  d <- .md[, .(celltype = as.character(get(.col_ct)),
               burden = as.numeric(genomwide_cnv_proportion))][!is.na(burden)]
  ord <- d[, .(m = median(burden, na.rm = TRUE), n = .N), by = celltype][order(-m)]

  # Order AFTER every join. data.table's merge() on a factor key returns a
  # CHARACTER column with the levels dropped, so setting the order first and
  # merging afterwards silently reverted this panel to alphabetical while the
  # subtitle still claimed it was ranked by median.
  d[, celltype := factor(celltype, levels = ord$celltype)]

  # A lollipop, not a violin. inferCNV was run with cluster_by_groups = TRUE, so
  # it estimates copy number PER GROUP and add_to_seurat() writes that single
  # value onto every cell in the group -- 9 distinct values across 24,415 cells
  # on this data, sd = 0 within every population. A violin or boxplot of it is
  # not merely ugly, it is a lie about the resolution: there is no distribution
  # to draw. One mark per group is the honest encoding.
  ord[, celltype := factor(celltype, levels = rev(ord$celltype))]
  ord[, is_ref := grepl("T cell|B cell|NK|Macrophage|Fibroblast|Endothel|Dendritic|Mast|pDC|Plasma",
                        celltype, ignore.case = TRUE)]

  ggplot(ord, aes(x = m, y = celltype,
                  text = sprintf("%s\n%s cells\nCNV proportion %.3f",
                                 celltype, format(n, big.mark = ","), m))) +
    geom_segment(aes(x = 0, xend = m, yend = celltype), colour = "grey80", linewidth = .8) +
    geom_point(aes(size = n, colour = is_ref)) +
    scale_colour_manual(values = c(`TRUE` = "#4a7fb5", `FALSE` = "#b32020"),
                        labels = c(`TRUE` = "reference lineage", `FALSE` = "tumour candidate"),
                        name = NULL) +
    scale_size_continuous(name = "cells", range = c(2.5, 8), labels = scales::comma) +
    scale_x_continuous(expand = expansion(mult = c(0, .12))) +
    labs(title = "Genome-wide CNV burden by cell type",
         subtitle = "One value per group: inferCNV ran per group, so every cell in a population shares its score",
         x = "genome-wide CNV proportion (log1p)", y = NULL) +
    theme_scratch() +
    theme(panel.grid.major.y = element_line(colour = "grey93", linewidth = .3))
}
```

```{r cnv-burden-finding, results = 'asis'}
if (!is.na(.col_ct) && "genomwide_cnv_proportion" %in% names(.md)) {
  d <- .md[, .(ct = as.character(get(.col_ct)), b = as.numeric(genomwide_cnv_proportion))][!is.na(b)]
  st <- d[, .(n = .N, med = median(b), q75 = quantile(b, .75)), by = ct][order(-med)]
  top <- st[1]; bot <- st[.N]

  # Reference lineages are supposed to be diploid. If one of them carries burden
  # comparable to the top population, the baseline itself is contaminated and
  # every downstream call inherits that.
  ref_like <- st[grepl("T cell|B cell|NK|Macrophage|Fibroblast|Endothel|Dendritic|Mast|pDC|Plasma",
                       ct, ignore.case = TRUE)]
  worst_ref <- if (nrow(ref_like)) ref_like[order(-med)][1] else NULL
  ratio <- if (!is.null(worst_ref) && bot$med > 0) worst_ref$med / bot$med else NA_real_

  # A zero-median population makes the ratio infinite, and sprintf renders that
  # as "NAx". State the separation in absolute terms when the floor is zero --
  # which is the healthy case: a reference lineage with no detectable CNV.
  spread <- if (bot$med > 0) sprintf("**%.1fx**", top$med / bot$med)
            else sprintf("unbounded — **%s has no detectable CNV at all** (median 0)", bot$ct)

  scratch_finding(sprintf(paste(
    "Highest CNV burden is **%s** (median %.3f over %s cells); lowest is **%s**",
    "(median %.3f). The separation between them is %s.%s"),
    top$ct, top$med, format(top$n, big.mark = ","), bot$ct, bot$med, spread,
    if (!is.null(worst_ref) && !is.na(ratio) && ratio > 3)
      sprintf(paste(" **%s is a reference-type lineage yet carries %.1fx the lowest burden** —",
                    "check that the reference set is genuinely diploid before trusting the calls."),
              worst_ref$ct, ratio) else ""),
    type = if (!is.null(worst_ref) && !is.na(ratio) && ratio > 3) "caution" else "note",
    title = "CNV burden by population")

  # Resolution matters more than the ranking. A reader seeing a per-cell column
  # reasonably assumes a per-cell measurement.
  n_uniq <- uniqueN(d$b); n_cells <- nrow(d)
  if (n_uniq <= 2 * uniqueN(d$ct)) {
    scratch_finding(sprintf(paste(
      "This score has **no per-cell resolution**: %d distinct values across %s cells",
      "(%d population groups). inferCNV was run with `cluster_by_groups = TRUE`, so",
      "it estimates copy number per GROUP and every cell in a group is assigned the",
      "same value. Consequences worth carrying forward: a per-cell threshold on this",
      "score can only ever split whole populations, it cannot identify malignant",
      "cells *within* a population, and any figure implying a within-group",
      "distribution is misleading. An allele-aware caller such as Numbat is",
      "genuinely per cell — which is why it, not this, drives the compartment split."),
      n_uniq, format(n_cells, big.mark = ","), uniqueN(d$ct)),
      type = "caution", title = "inferCNV burden is per group, not per cell")
  }
}
```

```{r cnv-burden-table}
# scratch_table() returns an htmlwidget, and a `results='asis'` chunk suppresses
# auto-printing -- so a table called from inside a findings chunk renders as
# NOTHING. Both tables in this section vanished that way. Findings and tables
# belong in separate chunks for exactly this reason.
if (!is.na(.col_ct) && "genomwide_cnv_proportion" %in% names(.md)) {
  d <- .md[, .(ct = as.character(get(.col_ct)), b = as.numeric(genomwide_cnv_proportion))][!is.na(b)]
  st <- d[, .(n = .N, med = median(b), q75 = quantile(b, .75)), by = ct][order(-med)]
  scratch_table(st[, .(`cell type` = ct, cells = n,
                       `median burden` = round(med, 4), `75th pct` = round(q75, 4))],
                caption = "Genome-wide CNV burden per annotated population")
}
```

### Where in the genome, and in which direction?

A chromosome carrying amplification in one population and deletion in another is
a different statement from one that is simply noisy, so gains and losses are kept
apart rather than summed into a single "altered" score.

```{r fig.width = 11, fig.height = 6}
#| label: fig-cnv-chr-heatmap
#| fig-cap: "Mean CNV proportion per chromosome and cell type"
if (!is.null(cnv_long)) {
  hm <- cnv_long[, .(prop = mean(prop_cnv, na.rm = TRUE), n = .N), by = .(celltype, chr)]
  ord <- hm[, .(m = mean(prop, na.rm = TRUE)), by = celltype][order(-m)]
  hm[, celltype := factor(celltype, levels = rev(ord$celltype))]

  ggplot(hm, aes(x = chr, y = celltype, fill = prop,
                 text = sprintf("chr%s — %s\nmean altered: %.1f%%\ncells: %s",
                                chr, celltype, 100 * prop, format(n, big.mark = ",")))) +
    geom_tile(colour = "white", linewidth = .4) +
    scale_fill_gradient(low = "#f2f6fa", high = "#8c1d18", name = "fraction\naltered",
                        labels = scales::percent_format(accuracy = 1)) +
    labs(title = "Copy-number alteration by chromosome and population",
         subtitle = "Fraction of each chromosome called altered, averaged over cells",
         x = "chromosome", y = NULL) +
    theme_scratch() +
    theme(panel.grid = element_blank())
}
```

```{r fig.width = 10, fig.height = 5}
#| label: fig-cnv-ampdel
#| fig-cap: "Amplification versus deletion per chromosome"
if (!is.null(cnv_long) && all(c("prop_dupli", "prop_loss") %in% names(cnv_long))) {
  ad <- cnv_long[, .(amp = mean(prop_dupli, na.rm = TRUE),
                     del = mean(prop_loss,  na.rm = TRUE)), by = chr]
  adl <- rbind(ad[, .(chr, direction = "amplification", value =  amp)],
               ad[, .(chr, direction = "deletion",      value = -del)])

  ggplot(adl, aes(x = chr, y = value, fill = direction,
                  text = sprintf("chr%s\n%s: %.1f%% of the chromosome",
                                 chr, direction, 100 * abs(value)))) +
    geom_col(width = .74) +
    geom_hline(yintercept = 0, colour = "grey40", linewidth = .4) +
    scale_fill_scratch(name = NULL) +
    scale_y_continuous(labels = function(x) scales::percent(abs(x), accuracy = 1)) +
    labs(title = "Gains and losses per chromosome",
         subtitle = "Above the line: amplification. Below: deletion. Mean fraction of the chromosome affected.",
         x = "chromosome", y = "fraction of chromosome") +
    theme_scratch()
}
```

```{r cnv-chr-finding, results = 'asis'}
if (!is.null(cnv_long)) {
  bychr <- cnv_long[, .(prop = mean(prop_cnv, na.rm = TRUE),
                        amp  = mean(prop_dupli, na.rm = TRUE),
                        del  = mean(prop_loss,  na.rm = TRUE)), by = chr][order(-prop)]
  top3 <- bychr[1:min(3, .N)]
  amp_tot <- sum(bychr$amp, na.rm = TRUE); del_tot <- sum(bychr$del, na.rm = TRUE)
  skew <- if (del_tot > 0) amp_tot / del_tot else NA_real_

  scratch_finding(sprintf(paste(
    "Most-altered chromosomes: %s. Across the genome the calls are **%.1fx more",
    "%s than %s** (%.1f%% vs %.1f%% of chromosome content on average).%s"),
    paste(sprintf("**chr%s** (%.1f%%)", top3$chr, 100 * top3$prop), collapse = ", "),
    if (!is.na(skew)) max(skew, 1 / skew) else NA_real_,
    if (!is.na(skew) && skew >= 1) "amplified" else "deleted",
    if (!is.na(skew) && skew >= 1) "deleted" else "amplified",
    100 * amp_tot / nrow(bychr), 100 * del_tot / nrow(bychr),
    if (!is.na(skew) && (skew > 5 || skew < 0.2))
      " A strong one-sided skew usually means the reference baseline is offset rather than that the tumour is uniformly gained or lost."
    else ""),
    type = if (!is.na(skew) && (skew > 5 || skew < 0.2)) "caution" else "note",
    title = "Genomic distribution of alterations")
}
```

```{r cnv-chr-table}
if (!is.null(cnv_long)) {
  bychr <- cnv_long[, .(prop = mean(prop_cnv, na.rm = TRUE),
                        amp  = mean(prop_dupli, na.rm = TRUE),
                        del  = mean(prop_loss,  na.rm = TRUE)), by = chr][order(-prop)]
  scratch_table(bychr[, .(chromosome = paste0("chr", chr),
                          `altered %` = round(100 * prop, 2),
                          `amplified %` = round(100 * amp, 2),
                          `deleted %` = round(100 * del, 2))],
                caption = "Per-chromosome alteration summary")
}
```

### HMM-segmented regions

The heatmap shows intensity; the HMM says where the boundaries are. These are the
discrete calls, comparable to Numbat's segments in the caller-concordance report.

```{r cnv-hmm-regions}
#| code-fold: true
hmm <- NULL
.rf <- list.files("data/infercnv", pattern = "pred_cnv_regions\\.dat$",
                  full.names = TRUE, recursive = TRUE)
if (length(.rf)) {
  hmm <- tryCatch(fread(.rf[1]), error = function(e) NULL)
  if (!is.null(hmm) && nrow(hmm)) {
    hmm[, size_Mb := (end - start) / 1e6]
    # inferCNV i6 states: 1-2 loss, 3 neutral, 4-6 gain.
    hmm[, direction := fifelse(state < 3, "loss", fifelse(state > 3, "gain", "neutral"))]
    hmm[, population := sub("\\..*$", "", cell_group_name)]
  }
}
```

```{r fig.width = 10.5, fig.height = 5.5}
#| label: fig-cnv-hmm
#| fig-cap: "HMM-predicted CNV regions by chromosome"
if (!is.null(hmm) && nrow(hmm)) {
  d <- copy(hmm)
  d[, chr_n := factor(sub("^chr", "", chr), levels = c(as.character(1:22), "X", "Y"))]
  ggplot(d[!is.na(chr_n)], aes(x = chr_n, y = size_Mb, fill = direction,
             text = sprintf("%s\n%s %s\n%.1f Mb (%.1f–%.1f Mb)\nstate %s",
                            population, chr, direction, size_Mb,
                            start / 1e6, end / 1e6, state))) +
    geom_col(width = .74) +
    scale_fill_scratch(name = NULL) +
    labs(title = "HMM-predicted copy-number regions",
         subtitle = "Total segment length called per chromosome, split by direction",
         x = "chromosome", y = "Mb called") +
    theme_scratch()
}
```

```{r cnv-hmm-finding, results = 'asis'}
if (!is.null(hmm) && nrow(hmm)) {
  bydir <- hmm[, .N, by = direction][order(-N)]
  scratch_finding(sprintf(paste(
    "The HMM resolved **%s regions** across %s population group(s): %s.",
    "Median segment **%.1f Mb** (range %.1f–%.1f). These are discrete calls and",
    "are directly comparable with Numbat's segments in the caller-concordance report."),
    format(nrow(hmm), big.mark = ","), uniqueN(hmm$population),
    paste(sprintf("%s %s", bydir$N, bydir$direction), collapse = ", "),
    median(hmm$size_Mb), min(hmm$size_Mb), max(hmm$size_Mb)),
    type = "note", title = "HMM segmentation")
}
```

```{r cnv-hmm-table}
if (!is.null(hmm) && nrow(hmm)) {
  # Side effects FIRST. scratch_table() returns a widget, and a chunk renders its
  # LAST value — with fwrite() after it, the table was computed, the CSV was
  # written, and nothing appeared in the report. Same rule that governs ggplot
  # objects, and just as silent when broken.
  fwrite(hmm, file.path("data", paste0(project_name, "_infercnv_hmm_regions.csv")))

  scratch_table(hmm[order(-size_Mb)][1:min(50, .N),
                    .(population, chromosome = chr, state, direction,
                      `start Mb` = round(start / 1e6, 2),
                      `end Mb` = round(end / 1e6, 2),
                      `size Mb` = round(size_Mb, 2))],
                caption = "Largest HMM-predicted CNV regions")
}
```

## Saving Seurat object

```{r object_dump}

# Saving full Seurat object
if(auto_save) {

  saveRDS(seurat_infercnv_object, file = paste0('./data/', project_name, '_infercnv_meta_object.RDS'))

}

```

------------------------------------------------------------------------

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

params

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```
