---
title: "SCRATCH CellComm - NicheNet"
author: "Syed Shujaat Ali Zaidi"
execute:
  freeze: auto
  cache: false
  warning: false
  message: true
  daemon: false
format:
  html:
    embed-resources: true  # Change this to true
    self-contained: true    # Keep for compatibility
    keep-md: false
    toc: true
    toc-location: right
    toc-depth: 2
    code-fold: show
    code-tools: true
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
  match: auto
  effect: zoom
  loop: true
params:
  # -----------------
  # Required inputs
  # -----------------
  seurat_object: null

  # NicheNet resources (MANDATORY; user provides these RDS files)
  # lr_network_rds: null
  # ligand_target_matrix_rds: null
  # weighted_networks_rds: null
  nichenet_assets_dir: "assets/nichenet_resources"

  # Expected upstream artifact (available in SCRATCH; used if present; fallback if missing)
  liana_csv: "project/cellcommunication/liana/data/liana_results.csv"
  

  # Optional upstream artifact (ignored by default; keep param for future)
  cellchat_rds: "project/cellcommunication/cellchat/data/cellchat_object.rds"

  # -----------------
  # Project
  # -----------------
  project_name: "project"
  work_directory: "."

  # -----------------
  # Data conventions
  # -----------------
  assay: "RNA"
  slot: null              # Seurat v4 slot OR Seurat v5 layer
  layer: "data"
  celltype_col: "celltype"
  condition_col: "batch"
  condition_ref: "1"
  condition_test: "2"

  # -----------------
  # Robustness filters
  # -----------------
  min_cells_per_group: 30
  min_cells_per_condition_in_receiver: 15
  expressed_pct: 0.05

  # -----------------
  # Receiver program definition
  # -----------------
  receiver_program_mode: "de"   # de|geneset
  # condition_ref: "Control"
  # condition_test: "Case"

  # If receiver_program_mode == "geneset"
  genes_of_interest: ""         # comma/semicolon separated list; if provided, overrides DE
  geneset_name: "GOI"

  # If receiver_program_mode == "de"
  logfc_min: 0.25
  padj_max: 0.05
  max_geneset: 300

  # -----------------
  # Auto receiver/sender selection (DEFAULT)
  # -----------------
  auto_k_receivers: 3
  auto_m_senders: 5

  # If you want manual override (optional)
  receiver_mode: "auto"         # auto|manual
  sender_mode: "auto"           # auto|manual|all_other
  manual_receivers: ""
  manual_senders: ""

  # -----------------
  # NicheNet outputs
  # -----------------
  top_ligands: 30
  heatmap_ligands: 15
  heatmap_targets_per_ligand: 40
  lr_support_top_ligands: 30
  lr_support_top_pairs: 60

  # -----------------
  # Visualizations
  # -----------------
  show_gene_set_summary: true
  plot_lr_support: true
  plot_ligand_target_heatmap: true
  plot_ligand_receptor_heatmap: true
  plot_sender_ligand_expression: true
  plot_receiver_receptor_expression: true
  plot_network_graph: true
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r}
suppressPackageStartupMessages({
library(Seurat)
library(SeuratObject)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(ggplot2)
library(Matrix)
library(ComplexHeatmap)
library(circlize)
library(scales)
library(igraph)
library(ggraph)
library(rlang)
})
# if (!requireNamespace("rlang", quietly = TRUE)) stop("Missing package: rlang")
# SeuratObject::PackageCheck("presto", error = FALSE)
# SeuratObject::GetAssayData
# getS3method("GetAssayData", "StdAssay")



seurat_v  <- as.character(packageVersion("Seurat"))
so_v      <- as.character(packageVersion("SeuratObject"))
# ---- Patch: override SeuratObject::PackageCheck (defunct in SeuratObject >= 5) ----
if (!requireNamespace("rlang", quietly = TRUE)) stop("Missing package: rlang")

slot_or_layer <- params$layer %||% slot_or_layer %||% "data"




message("Seurat=", seurat_v, " | SeuratObject=", so_v)

# ---- HARD PATCH: make SeuratObject::GetAssayData accept slot= again ----
if (packageVersion("SeuratObject") >= "5.0.0") {

  nsSO <- asNamespace("SeuratObject")

  patch_one <- function(fun_name) {
    if (!exists(fun_name, envir = nsSO, inherits = FALSE)) return(FALSE)

    orig <- get(fun_name, envir = nsSO, inherits = FALSE)

    shim <- function(object, layer = NULL, slot = NULL, ...) {
      # if old code calls slot=, map it to layer=
      if (is.null(layer) && !is.null(slot)) layer <- slot

      # call original using layer=
      # (some originals take layer explicitly; some take ...)
      if ("layer" %in% names(formals(orig))) {
        return(orig(object = object, layer = layer, ...))
      } else {
        # fallback: pass as first arg after object; many methods accept via ...
        return(orig(object = object, layer = layer, ...))
      }
    }

    # overwrite in namespace (bindings may be locked)
    try(unlockBinding(fun_name, nsSO), silent = TRUE)
    assign(fun_name, shim, envir = nsSO)
    try(lockBinding(fun_name, nsSO), silent = TRUE)

    TRUE
  }

  # Patch the ones Seurat commonly dispatches to in v5 objects
  patched <- c()
  for (nm in c("GetAssayData.StdAssay", "GetAssayData.Assay5", "GetAssayData.Assay")) {
    if (patch_one(nm)) patched <- c(patched, nm)
  }

  if (length(patched) == 0) {
    message("WARNING: No GetAssayData.* methods found to patch in SeuratObject namespace.")
  } else {
    message("Patched SeuratObject methods to map slot -> layer: ", paste(patched, collapse = ", "))
  }
}


# If SeuratObject is v5+, Seurat must also be v5+ (otherwise FindMarkers will call GetAssayData(slot=...))
# if (packageVersion("SeuratObject") >= "5.0.0" && packageVersion("Seurat") < "5.0.0") {
#   stop(
#     "Incompatible package versions detected:\n",
#     "  SeuratObject >= 5.0.0 but Seurat < 5.0.0.\n",
#     "Fix: upgrade Seurat to v5+ OR downgrade SeuratObject to <5.\n",
#     "Current: Seurat=", seurat_v, " | SeuratObject=", so_v
#   )
# }


`%||%` <- function(a, b) if (!is.null(a)) a else b
# slot_or_layer <- params$layer %||% slot_or_layer %||% "data"


if (!requireNamespace("nichenetr", quietly = TRUE)) stop("Missing package: nichenetr")
suppressPackageStartupMessages(library(nichenetr))

# ----------------------------

# Working dir & output dirs

# ----------------------------

if (!is.null(params$work_directory) && dir.exists(params$work_directory)) {
setwd(params$work_directory)
}

dir.create("figures/cellcomm/nichenet", recursive = TRUE, showWarnings = FALSE)
dir.create("data/nichenet",            recursive = TRUE, showWarnings = FALSE)
dir.create("data",                    recursive = TRUE, showWarnings = FALSE)

theme_set(theme_bw(base_size = 12))
set.seed(1)

# ----------------------------

# Utils

# ----------------------------

parse_list <- function(x) {
if (is.null(x) || length(x) == 0 || is.na(x) || !nzchar(x)) return(character(0))
v <- unlist(strsplit(x, "[,;|]"))
v <- trimws(v)
v[nzchar(v)]
}

norm01 <- function(x) {
if (all(is.na(x))) return(rep(0, length(x)))
rng <- max(x, na.rm = TRUE) - min(x, na.rm = TRUE)
if (rng == 0) return(rep(0, length(x)))
(x - min(x, na.rm = TRUE)) / rng
}

get_expressed <- function(mat, cells, min_pct = 0.05) {
if (length(cells) == 0) return(character(0))
sub <- mat[, cells, drop = FALSE]
frac <- Matrix::rowMeans(sub > 0)
names(frac)[frac >= min_pct]
}

avg_expr <- function(mat, cells) {
if (length(cells) == 0) return(setNames(numeric(0), character(0)))
sub <- mat[, cells, drop = FALSE]
Matrix::rowMeans(sub)
}

# ----------------------------

# Seurat v4/v5 compatibility

# ----------------------------
join_layers_if_needed <- function(obj, assay) {
  a <- obj[[assay]]
  if (is.null(a)) stop("Assay not found: ", assay)

  # Only relevant for Seurat v5 / Assay5 style objects
  has_layers <- "Layers" %in% getNamespaceExports("SeuratObject")
  has_join   <- "JoinLayers" %in% getNamespaceExports("SeuratObject")

  if (has_layers && has_join) {
    lays <- tryCatch(SeuratObject::Layers(a), error = function(e) NULL)
    if (!is.null(lays) && length(lays) > 1) {
      message("JoinLayers: assay '", assay, "' has ", length(lays), " layers -> joining")
      obj[[assay]] <- SeuratObject::JoinLayers(a)
    }
  }
  obj
}

get_assay_mat <- function(obj, assay="RNA", slot_or_layer="data") {
  obj <- join_layers_if_needed(obj, assay)

  # inspect the Seurat S3 method (this is the one that matters)
  m <- getS3method("GetAssayData", "Seurat")
  fm <- names(formals(m))

  if ("layer" %in% fm) {
    return(SeuratObject::GetAssayData(obj, assay = assay, layer = slot_or_layer))
  } else {
    return(SeuratObject::GetAssayData(obj, assay = assay, slot  = slot_or_layer))
  }
}

# get_assay_mat <- function(obj, assay = "RNA", slot_or_layer = "data") {
#   obj <- join_layers_if_needed(obj, assay)
# 
#   so_v5 <- utils::packageVersion("SeuratObject") >= "5.0.0"
# 
#   if (so_v5) {
#     return(SeuratObject::GetAssayData(obj, assay = assay, layer = slot_or_layer))
#   } else {
#     return(SeuratObject::GetAssayData(obj, assay = assay, slot  = slot_or_layer))
#   }
# }


subset_by_meta <- function(obj, col, value) {
  md <- obj@meta.data
  if (!col %in% colnames(md)) stop("Metadata column not found: ", col)
  cells <- rownames(md)[md[[col]] %in% value]
  obj[, cells, drop = FALSE]   # <-- works across Seurat v4/v5 without Seurat::subset
}


# subset_by_meta <- function(obj, col, value) {
# md <- obj@meta.data
# if (!col %in% colnames(md)) stop("Metadata column not found: ", col)
# cells <- rownames(md)[md[[col]] %in% value]
# SeuratObject::subset(obj, cells = cells)
# }

find_markers_compat <- function(obj, ident.1, ident.2,
                               assay = "RNA", slot_or_layer = "data",
                               logfc.threshold = 0.1, min.pct = 0.1, ...) {

  obj <- join_layers_if_needed(obj, assay)

  # fm_formals <- names(formals(Seurat::FindMarkers))
  fm_formals <- names(formals(getS3method("FindMarkers", "Seurat")))

  args <- list(
    object = obj,
    ident.1 = ident.1,
    ident.2 = ident.2,
    assay = assay,
    logfc.threshold = logfc.threshold,
    min.pct = min.pct,
    ...
  )

  if ("layer" %in% fm_formals) {
    args$layer <- slot_or_layer
  } else {
    args$slot  <- slot_or_layer
  }

  do.call(Seurat::FindMarkers, args)
}


# ----------------------------

# Robust metadata column resolver

# ----------------------------

resolve_col <- function(md, preferred, candidates, label) {
if (!is.null(preferred) && nzchar(preferred) && preferred %in% colnames(md)) return(preferred)
hit <- candidates[candidates %in% colnames(md)][1]
if (length(hit) == 0 || is.na(hit) || !nzchar(hit)) {
stop("No valid ", label, " column found. Tried preferred='", preferred,
"' and candidates: ", paste(candidates, collapse = ", "))
}
hit
}

CELLTYPE_CANDIDATES <- c(
"celltype","cell_type","CellType","celltype_final","celltype_refined",
"azimuth_labels","azimuth_label","predicted.celltype.l2","predicted.celltype.l1",
"annotation","annot","label","labels","final_annotation","seurat_clusters"
)

CONDITION_CANDIDATES <- c(
"condition","Condition","group","Group","case_control","status",
"DiseaseStatus","disease_status","phenotype","response","treatment","batch"
)

# ----------------------------

# LIANA schema resolver

# ----------------------------

resolve_liana_cols <- function(df) {
cands <- list(
source   = c("source","sender","cell_source","from"),
target   = c("target","receiver","cell_target","to"),
ligand   = c("ligand","ligand_symbol","ligand_name","ligand.complex","ligand_complex"),
receptor = c("receptor","receptor_symbol","receptor_name","receptor.complex","receptor_complex")
)
out <- list()
for (k in names(cands)) {
hit <- cands[[k]][cands[[k]] %in% colnames(df)][1]
out[[k]] <- ifelse(length(hit)==0 || is.na(hit), NA_character_, hit)
}
out
}

# ----------------------------

# NicheNet resources loader (NO URLs; user-provided)

# ----------------------------
resolve_nichenet_assets <- function(dir) {
  stopifnot(!is.null(dir), dir.exists(dir))

  # Prefer standard names if present; else match common Zenodo filenames
  candidates <- list(
    lr = c("lr_network.rds", "lr_network_human_21122021.rds", "lr_network_mouse_*.rds", "lr_network_*human*.rds"),
    lt = c("ligand_target_matrix.rds", "ligand_target_matrix_nsga2r_final.rds", "ligand_target_matrix_*.rds"),
    wn = c("weighted_networks.rds", "weighted_networks_nsga2r_final.rds", "weighted_networks_*.rds")
  )

  pick_one <- function(patterns) {
    for (p in patterns) {
      hits <- Sys.glob(file.path(dir, p))
      if (length(hits) > 0) return(hits[[1]])
    }
    NA_character_
  }

  lr_path <- pick_one(candidates$lr)
  lt_path <- pick_one(candidates$lt)
  wn_path <- pick_one(candidates$wn)

  if (is.na(lr_path)) stop("Could not find lr_network RDS under: ", dir)
  if (is.na(lt_path)) stop("Could not find ligand_target_matrix RDS under: ", dir)
  if (is.na(wn_path)) stop("Could not find weighted_networks RDS under: ", dir)

  list(lr = lr_path, lt = lt_path, wn = wn_path)
}

assets_dir <- params$nichenet_assets_dir
lr_path <- file.path(assets_dir, "lr_network_human.rds")
lt_path <- file.path(assets_dir, "ligand_target_matrix.rds")
wn_path <- file.path(assets_dir, "weighted_networks.rds")

stopifnot(dir.exists(assets_dir))
stopifnot(file.exists(lr_path), file.exists(lt_path), file.exists(wn_path))

lr_network <- readRDS(lr_path) %>% dplyr::distinct(from, to)
ligand_target_matrix <- readRDS(lt_path)
weighted_networks <- readRDS(wn_path)

# sanity checks
stopifnot(is.data.frame(lr_network), all(c("from","to") %in% colnames(lr_network)))
stopifnot(is.matrix(ligand_target_matrix) || inherits(ligand_target_matrix, "Matrix"))
stopifnot(is.list(weighted_networks))

all_ligands   <- unique(lr_network$from)
all_receptors <- unique(lr_network$to)



# Sanity checks (fast)
stopifnot(is.data.frame(lr_network), all(c("from","to") %in% colnames(lr_network)))
stopifnot(is.matrix(ligand_target_matrix) || inherits(ligand_target_matrix, "Matrix"))
stopifnot(is.list(weighted_networks))


load_nichenet_resources <- function(lr_network_rds, ligand_target_matrix_rds, weighted_networks_rds) {
stopifnot(!is.null(lr_network_rds), file.exists(lr_network_rds))
stopifnot(!is.null(ligand_target_matrix_rds), file.exists(ligand_target_matrix_rds))
stopifnot(!is.null(weighted_networks_rds), file.exists(weighted_networks_rds))

lr_network <- readRDS(lr_network_rds)
ligand_target_matrix <- readRDS(ligand_target_matrix_rds)
weighted_networks <- readRDS(weighted_networks_rds)

lr_network <- lr_network %>% dplyr::distinct(from, to)

list(lr_network = lr_network,
ligand_target_matrix = ligand_target_matrix,
weighted_networks = weighted_networks)
}
```

##What this notebook does
##Takes an analysis-ready Seurat object (batch corrected, annotated).
##Uses NicheNet to rank ligands that best explain a receiver’s response gene program.
##Uses LIANA (expected upstream) to restrict candidate ligands to LR-supported ligands (sender→receiver). If LIANA is missing or too restrictive, it falls back to expression+LR prior only.
##Automatically selects receivers and senders unless the user provides a genes of interest program or manual receivers/senders.

```{r}
# ---- required ----

stopifnot(!is.null(params$seurat_object), file.exists(params$seurat_object))

# ---- expected upstream ----

has_liana <- !is.null(params$liana_csv) && file.exists(params$liana_csv)
if (!has_liana) message("LIANA CSV not found -> running NicheNet without LIANA-guided restriction (fallback).")

# Load Seurat

seu <- readRDS(params$seurat_object)
stopifnot(inherits(seu, "Seurat"))
DefaultAssay(seu) <- params$assay
seu <- join_layers_if_needed(seu, params$assay)

# Resolve columns

md <- seu@meta.data
celltype_col  <- resolve_col(md, params$celltype_col,  CELLTYPE_CANDIDATES,  "celltype")
condition_col <- resolve_col(md, params$condition_col, CONDITION_CANDIDATES, "condition")

# if condition_ref/test are not present in the data, auto-pick the two most frequent levels
cond_vals <- seu@meta.data[[condition_col]]
lvl_counts <- sort(table(cond_vals), decreasing = TRUE)

# ---- pick condition_ref / condition_test WITHOUT modifying params ----
condition_ref  <- as.character(params$condition_ref)
condition_test <- as.character(params$condition_test)

if (!(condition_ref %in% names(lvl_counts)) || !(condition_test %in% names(lvl_counts))) {
  if (length(lvl_counts) < 2) stop("Need at least 2 levels in condition_col='", condition_col, "' to run DE mode.")
  condition_ref  <- names(lvl_counts)[1]
  condition_test <- names(lvl_counts)[2]
  message("Auto-selected condition_ref='", condition_ref, "' and condition_test='", condition_test, "' from condition_col='", condition_col, "'.")
} else {
  message("Using provided condition_ref='", condition_ref, "' and condition_test='", condition_test, "'.")
}
##################################################################################
# sanity
stopifnot(condition_ref %in% names(lvl_counts), condition_test %in% names(lvl_counts))



message("Resolved columns:\n  celltype_col  = ", celltype_col, "\n  condition_col = ", condition_col)
write_csv(tibble(celltype_col = celltype_col, condition_col = condition_col),
file.path("data/nichenet", "resolved_columns_nichenet.csv"))

# Filter rare cell types

ct_counts <- seu@meta.data %>%
count(.data[[celltype_col]], name = "n_cells") %>%
arrange(desc(n_cells))

keep_ct <- ct_counts %>% filter(n_cells >= params$min_cells_per_group) %>% pull(1)
keep_cells <- rownames(seu@meta.data)[seu@meta.data[[celltype_col]] %in% keep_ct]
stopifnot(length(keep_cells) > 0)

seu <- subset(seu, cells = keep_cells)

ct_counts2 <- seu@meta.data %>% count(.data[[celltype_col]], name="n_cells") %>% arrange(desc(n_cells))
write_csv(ct_counts2, file.path("data/nichenet", "celltype_counts_filtered.csv"))

celltypes <- sort(unique(seu@meta.data[[celltype_col]]))

# Expression matrix for "expressed genes"

# exprs_all <- get_assay_mat(seu, assay = params$assay, slot_or_layer = slot_or_layer)
exprs_all <- get_assay_mat(seu, assay = params$assay, slot_or_layer = slot_or_layer)


meta_all <- seu@meta.data %>%
mutate(celltype = .data[[celltype_col]],
condition = .data[[condition_col]])

resolve_path <- function(p) {
  if (is.null(p) || !nzchar(p)) return(NA_character_)
  if (grepl("^/", p)) return(p)
  file.path(params$work_directory %||% ".", p)
}

liana_csv_path    <- resolve_path(params$liana_csv)
cellchat_rds_path <- resolve_path(params$cellchat_rds)

has_liana   <- !is.na(liana_csv_path)   && file.exists(liana_csv_path)
has_cellchat<- !is.na(cellchat_rds_path)&& file.exists(cellchat_rds_path)

if (!has_liana)   message("LIANA CSV missing -> running without LIANA-guided restriction.")
if (!has_cellchat)message("CellChat RDS missing -> skipping CellChat-based additions.")

liana_res <- NULL; liana_cols <- NULL
if (has_liana) {
  liana_res  <- readr::read_csv(liana_csv_path, show_col_types = FALSE)
  liana_cols <- resolve_liana_cols(liana_res)

  if (any(is.na(unlist(liana_cols[c("source","target","ligand","receptor")])))) {
    warning("LIANA loaded but required columns not found; disabling LIANA-guided restriction.")
    has_liana <- FALSE
    liana_res <- NULL
    liana_cols <- NULL
  }
}


# Save a quick manifest

write_csv(
tibble(
seurat_object = params$seurat_object,
lr_network_rds = params$lr_network_rds,
ligand_target_matrix_rds = params$ligand_target_matrix_rds,
weighted_networks_rds = params$weighted_networks_rds,
liana_csv = params$liana_csv %||% "",
assay = params$assay,
slot = slot_or_layer,
celltype_col = celltype_col,
condition_col = condition_col
),
file.path("data/nichenet", "run_manifest.csv")
)

# Convenience sets (tutorial-like)

all_ligands   <- unique(lr_network$from)
all_receptors <- unique(lr_network$to)
```


##Receiver program mode
##If genes_of_interest is provided (non-empty), we use it as the receiver program and switch to "geneset" mode automatically. Otherwise we use receiver-specific DE (Case vs Control) like the tutorial.
```{r}
goi <- parse_list(params$genes_of_interest)

receiver_program_mode <- params$receiver_program_mode
if (length(goi) > 0) receiver_program_mode <- "geneset"

message("receiver_program_mode = ", receiver_program_mode)
if (receiver_program_mode == "geneset") {
message("Using genes_of_interest (n=", length(goi), "): ", paste(head(goi, 10), collapse = ", "),
if (length(goi) > 10) " ..." else "")
}

```


##Automatic receiver selection (default)
##If receiver_program_mode == "geneset": receivers are selected by GOI expression enrichment (average GOI expression per celltype).
##If receiver_program_mode == "de": receivers are selected by DE burden (number of up genes in Case vs Control) — same idea as the vignette where you define a receiver response.

```{r problem}
# Score receiver candidates

score_receivers_by_goi <- function(ct) {
cells <- rownames(meta_all)[meta_all$celltype == ct]
if (length(cells) < params$min_cells_per_group) return(NA_real_)
goi_present <- intersect(goi, rownames(exprs_all))
if (length(goi_present) < 3) return(NA_real_)
sub <- exprs_all[goi_present, cells, drop = FALSE]
mean(as.numeric(Matrix::colMeans(sub)))
}

compute_de_burden <- function(ct) {
  ct_cells <- rownames(seu@meta.data)[seu@meta.data[[celltype_col]] == ct]
  if (length(ct_cells) < params$min_cells_per_group) return(0)

  grp_vec <- seu@meta.data[[condition_col]]
  n_ref  <- sum(grp_vec[ct_cells] == condition_ref, na.rm = TRUE)
  n_test <- sum(grp_vec[ct_cells] == condition_test, na.rm = TRUE)

  if (n_ref < params$min_cells_per_condition_in_receiver ||
      n_test < params$min_cells_per_condition_in_receiver) return(0)

  de_tbl <- de_wilcox_layer(
    seu       = seu,
    cells_use = ct_cells,
    group_vec = grp_vec,
    group1    = condition_test,
    group2    = condition_ref,
    assay     = params$assay,
    layer     = slot_or_layer,
    min_pct   = 0.1,
    logfc_min = params$logfc_min
  )

  if (is.null(de_tbl) || nrow(de_tbl) == 0) return(0)
  nrow(de_tbl)
}


receiver_metrics <- tibble(celltype = celltypes)
```

```{r problem 1}
# Always use layer if present; fallback to slot for older objects
`%||%` <- function(a, b) if (!is.null(a)) a else b
slot_or_layer <- params$layer %||% params$slot %||% "data"

if (!requireNamespace("presto", quietly = TRUE)) {
  stop("Package 'presto' is required for DE in SeuratObject v5 when Seurat::FindMarkers breaks. Install it: install.packages('presto')")
}

# DE using assay matrix (layer) + presto (no Seurat::FindMarkers)
de_wilcox_layer <- function(seu, cells_use, group_vec, group1, group2,
                            assay = "RNA", layer = "data",
                            min_pct = 0.1, logfc_min = 0.25) {

  # subset metadata to cells
  group_vec <- as.character(group_vec[cells_use])
  keep <- group_vec %in% c(group1, group2)
  cells_use <- cells_use[keep]
  group_vec <- group_vec[keep]

  if (length(cells_use) == 0) return(NULL)

  # pull expression from layer (your get_assay_mat already uses layer when available)
  mat <- get_assay_mat(seu[, cells_use, drop = FALSE], assay = assay, slot_or_layer = layer)

  grp <- factor(group_vec, levels = c(group1, group2))
  n1 <- sum(grp == group1); n2 <- sum(grp == group2)
  if (n1 == 0 || n2 == 0) return(NULL)

  # gene filter by expression pct in either group
  idx1 <- which(grp == group1)
  idx2 <- which(grp == group2)
  pct1 <- Matrix::rowMeans(mat[, idx1, drop = FALSE] > 0)
  pct2 <- Matrix::rowMeans(mat[, idx2, drop = FALSE] > 0)
  keep_genes <- (pmax(pct1, pct2) >= min_pct)
  if (!any(keep_genes)) return(NULL)

  mat2 <- mat[keep_genes, , drop = FALSE]

  # presto Wilcoxon
  res <- presto::wilcoxauc(mat2, y = grp)

  # presto returns one row per feature per group; keep group1 rows and compute log2FC
  res1 <- res[res$group == group1, , drop = FALSE]

  # average expression per group for log2FC
  avg1 <- Matrix::rowMeans(mat2[, idx1, drop = FALSE])
  avg2 <- Matrix::rowMeans(mat2[, idx2, drop = FALSE])
  log2fc <- log2((avg1 + 1e-9) / (avg2 + 1e-9))

  out <- data.frame(
    gene       = res1$feature,
    p_val      = res1$pval,
    p_val_adj  = res1$padj,
    avg_log2FC = as.numeric(log2fc[match(res1$feature, rownames(mat2))]),
    stringsAsFactors = FALSE
  )

  # apply thresholds
  out <- out[!is.na(out$p_val_adj) & out$p_val_adj < params$padj_max &
               out$avg_log2FC > logfc_min, , drop = FALSE]
  out
}


if (receiver_program_mode == "geneset") {
receiver_metrics <- receiver_metrics %>%
mutate(goi_score = map_dbl(celltype, score_receivers_by_goi)) %>%
mutate(receiver_score = norm01(tidyr::replace_na(goi_score, 0))) %>%
arrange(desc(receiver_score))

} else {
receiver_metrics <- receiver_metrics %>%
mutate(de_burden = map_dbl(celltype, compute_de_burden)) %>%
mutate(receiver_score = norm01(de_burden)) %>%
arrange(desc(receiver_score))
}

write_csv(receiver_metrics, file.path("data/nichenet", "auto_receiver_metrics.csv"))
```

```{r problem2}
p_recv <- receiver_metrics %>%
mutate(celltype = factor(celltype, levels = rev(celltype))) %>%
ggplot(aes(x = celltype, y = receiver_score)) +
geom_col() + coord_flip() +
labs(
title = "Auto receiver ranking",
subtitle = if (receiver_program_mode == "geneset")
paste0("Based on GOI mean expression per celltype (n GOI = ", length(goi), ")")
else
"Based on DE burden (Case vs Control) within each celltype",
x = "Celltype", y = "Receiver score (0-1)"
)

ggsave(file.path("figures/cellcomm/nichenet", "auto_receiver_ranking.png"),
p_recv, width = 8.5, height = 6, dpi = 300)
p_recv

# Pick receivers (default auto)

manual_receivers <- parse_list(params$manual_receivers)
if (params$receiver_mode == "manual") {
receivers <- intersect(manual_receivers, celltypes)
if (length(receivers) == 0) stop("receiver_mode=manual but manual_receivers not found in object.")
} else {
receivers <- receiver_metrics %>% slice_head(n = as.integer(params$auto_k_receivers)) %>% pull(celltype)
}

receivers
```


##Automatic sender selection (default)
##Senders are chosen as the top celltypes by outgoing LR count from LIANA (preferred) or by abundance if LIANA is missing.

```{r}
manual_senders <- parse_list(params$manual_senders)

sender_score_table <- tibble(celltype = celltypes) %>%
left_join(ct_counts2 %>% rename(celltype = .data[[celltype_col]]), by="celltype") %>%
mutate(n_cells = replace_na(n_cells, 0))

if (has_liana) {
liana_out <- liana_res %>%
count(.data[[liana_cols$source]], name = "liana_out_lr") %>%
rename(celltype = .data[[liana_cols$source]])

sender_score_table <- sender_score_table %>%
left_join(liana_out, by="celltype") %>%
mutate(liana_out_lr = replace_na(liana_out_lr, 0))
} else {
sender_score_table <- sender_score_table %>% mutate(liana_out_lr = 0)
}

select_senders <- function(receiver_ct) {
all_others <- setdiff(celltypes, receiver_ct)

if (params$sender_mode == "manual") {
s <- intersect(manual_senders, all_others)
if (length(s) == 0) stop("sender_mode=manual but manual_senders invalid for receiver: ", receiver_ct)
return(s)
}
if (params$sender_mode == "all_other") return(all_others)

# auto: LIANA outgoing first, else n_cells

tbl <- sender_score_table %>%
filter(celltype %in% all_others)

if (has_liana) {
out_sorted <- tbl %>% arrange(desc(liana_out_lr), desc(n_cells)) %>%
slice_head(n = as.integer(params$auto_m_senders)) %>% pull(celltype)
} else {
out_sorted <- tbl %>% arrange(desc(n_cells)) %>%
slice_head(n = as.integer(params$auto_m_senders)) %>% pull(celltype)
}

if (length(out_sorted) == 0) all_others else out_sorted
}
```

##NicheNet per receiver (tutorial-aligned core steps)
##For each receiver:
##define response program (DE or GOI)
##define expressed ligands in senders and expressed receptors in receiver
##compute candidate ligands (expression + LR prior), then restrict using LIANA (directional), fallback if too few
##compute ligand activities (AUPR)
##visualize: activity ranking, ligand-target heatmap, ligand-receptor heatmap, LR support plot, network graph

```{r}
niche_summary <- list()

for (receiver_ct in receivers) {

senders <- select_senders(receiver_ct)

rec_tag <- paste0("receiver_", make.names(receiver_ct))
rec_dir_fig <- file.path("figures/cellcomm/nichenet", rec_tag)
rec_dir_dat <- file.path("data/nichenet", rec_tag)
dir.create(rec_dir_fig, recursive = TRUE, showWarnings = FALSE)
dir.create(rec_dir_dat, recursive = TRUE, showWarnings = FALSE)

message("\n---- Receiver: ", receiver_ct, " | Senders: ", paste(senders, collapse = ", "), " ----")

receiver_cells <- rownames(meta_all)[meta_all$celltype == receiver_ct]
sender_cells   <- rownames(meta_all)[meta_all$celltype %in% senders]

# -----------------------------

# Receiver response program

# -----------------------------

if (receiver_program_mode == "geneset") {
geneset_oi <- intersect(goi, rownames(ligand_target_matrix))
de_tbl <- tibble(gene = geneset_oi, source = "GOI")
} else {
sub <- subset_by_meta(seu, celltype_col, receiver_ct)
Idents(sub) <- as.factor(sub@meta.data[[condition_col]])


tab_cond <- table(Idents(sub))
if (!(condition_ref %in% names(tab_cond)) ||
    !(condition_test %in% names(tab_cond)) ||
    tab_cond[[condition_ref]] < params$min_cells_per_condition_in_receiver ||
    tab_cond[[condition_test]] < params$min_cells_per_condition_in_receiver) {
  warning("Skipping receiver ", receiver_ct, ": insufficient cells per condition.")
  next
}

de <- de_wilcox_layer(
  seu       = sub,  # or seu, but sub is fine if already subset
  cells_use = colnames(sub),
  group_vec = sub@meta.data[[condition_col]],
  group1    = condition_test,
  group2    = condition_ref,
  assay     = params$assay,
  layer     = slot_or_layer,
  min_pct   = 0.1,
  logfc_min = params$logfc_min
)

if (is.null(de)) {
  de <- data.frame(gene=character(), p_val_adj=numeric(), avg_log2FC=numeric())
}

# de already has a 'gene' column -> DO NOT do rownames_to_column()
de <- dplyr::arrange(de, p_val_adj)

write_csv(de, file.path(rec_dir_dat, "receiver_DE.csv"))

# geneset_oi <- de %>%
#   filter(!is.na(p_val_adj),
#          p_val_adj < params$padj_max,
#          avg_log2FC > params$logfc_min) %>%
#   slice_head(n = as.integer(params$max_geneset)) %>%
#   pull(gene)

fc_col <- intersect(c("avg_log2FC","avg_logFC","log2FC"), colnames(de))[1]
if (is.na(fc_col) || !nzchar(fc_col)) stop("No fold-change column found in DE table.")

geneset_oi <- de %>%
  filter(!is.na(p_val_adj),
         p_val_adj < params$padj_max,
         .data[[fc_col]] > params$logfc_min) %>%
  slice_head(n = as.integer(params$max_geneset)) %>%
  pull(gene)


geneset_oi <- intersect(geneset_oi, rownames(ligand_target_matrix))
de_tbl <- de


}

if (length(geneset_oi) < 10) {
warning("Skipping receiver ", receiver_ct, ": geneset_oi too small after filtering (", length(geneset_oi), ").")
next
}

# -----------------------------

# Expressed genes (sender/receiver)

# -----------------------------

expressed_senders  <- get_expressed(exprs_all, sender_cells,   min_pct = params$expressed_pct)
expressed_receiver <- get_expressed(exprs_all, receiver_cells, min_pct = params$expressed_pct)

expressed_ligands_senders <- intersect(all_ligands, expressed_senders)
expressed_receptors <- intersect(all_receptors, expressed_receiver)

# Candidate ligands: sender-expressed ligands that have at least one receptor expressed in receiver

potential_ligands_base <- lr_network %>%
filter(from %in% expressed_ligands_senders,
to %in% expressed_receptors) %>%
pull(from) %>%
unique()

# -----------------------------

# LIANA-guided restriction (directional), fallback if too few

# -----------------------------

potential_ligands <- potential_ligands_base
liana_ligs_dir <- character(0)

if (has_liana) {
liana_ligs_dir <- liana_res %>%
filter(.data[[liana_cols$target]] == receiver_ct,
.data[[liana_cols$source]] %in% senders) %>%
pull(.data[[liana_cols$ligand]]) %>%
unique()


liana_ligs_dir <- intersect(liana_ligs_dir, all_ligands)

guided <- intersect(potential_ligands_base, liana_ligs_dir)
if (length(guided) >= 5) {
  potential_ligands <- guided
  message("Using LIANA-guided potential ligands: ", length(potential_ligands),
          " (base=", length(potential_ligands_base), ")")
} else {
  message("LIANA-guided too few (", length(guided),
          ") -> fallback to base (", length(potential_ligands_base), ")")
}

write_csv(
  tibble(receiver = receiver_ct,
         senders = paste(senders, collapse = "; "),
         n_candidates_base = length(potential_ligands_base),
         n_liana_ligands_dir = length(liana_ligs_dir),
         n_candidates_final = length(potential_ligands)),
  file.path(rec_dir_dat, "ligand_candidate_counts.csv")
)


}

if (length(potential_ligands) < 5) {
warning("Skipping receiver ", receiver_ct, ": too few potential ligands (", length(potential_ligands), ").")
next
}

# Background expressed genes (must be in ligand_target_matrix)

background_expressed_genes <- intersect(expressed_receiver, rownames(ligand_target_matrix))
if (length(background_expressed_genes) < 500) {
warning("Receiver ", receiver_ct, ": background_expressed_genes small (", length(background_expressed_genes), ").")
}

# -----------------------------

# Ligand activity inference (core step)

# -----------------------------

ligand_activities <- nichenetr::predict_ligand_activities(
geneset                    = geneset_oi,
background_expressed_genes = background_expressed_genes,
ligand_target_matrix       = ligand_target_matrix,
potential_ligands          = potential_ligands
) %>% arrange(desc(aupr_corrected))

write_csv(ligand_activities, file.path(rec_dir_dat, "ligand_activities.csv"))

# -----------------------------

# Summary table

# -----------------------------

gene_set_summary <- tibble(
receiver = receiver_ct,
program_mode = receiver_program_mode,
geneset_name = if (receiver_program_mode == "geneset") params$geneset_name else "DE_up",
condition_ref = condition_ref,
condition_test = condition_test,
senders = paste(senders, collapse = "; "),
n_receiver_cells = length(receiver_cells),
n_sender_cells = length(sender_cells),
n_geneset_oi = length(geneset_oi),
n_background_expressed = length(background_expressed_genes),
n_receptors_expressed_receiver = length(expressed_receptors),
n_candidate_ligands_base = length(potential_ligands_base),
n_candidate_ligands_final = length(potential_ligands),
top_ligands = paste(head(ligand_activities$test_ligand, 10), collapse = "; ")
)
if (isTRUE(params$show_gene_set_summary)) {
write_csv(gene_set_summary, file.path(rec_dir_dat, "gene_set_summary.csv"))
}

# ==========================================================

# Advanced visualizations

# ==========================================================

# -----------------------------

# 1) Ligand activity barplot

# -----------------------------

plot_lig <- ligand_activities %>%
slice_head(n = as.integer(params$top_ligands)) %>%
mutate(test_ligand = factor(test_ligand, levels = rev(test_ligand)))

p_lig <- ggplot(plot_lig, aes(x = test_ligand, y = aupr_corrected)) +
geom_col() +
coord_flip() +
labs(
title = paste0("NicheNet ligand activity (receiver: ", receiver_ct, ")"),
subtitle = paste0("Program: ", gene_set_summary$geneset_name,
" | Senders: ", paste(senders, collapse = ", ")),
x = "Ligand",
y = "AUPR (corrected)"
)

ggsave(file.path(rec_dir_fig, "top_ligands_aupr.png"), p_lig, width = 9, height = 6, dpi = 300)
print(p_lig)

# -----------------------------

# 2) LR support dotplot (expression + activity)

# -----------------------------

if (isTRUE(params$plot_lr_support)) {
top_ligs <- ligand_activities %>%
slice_head(n = as.integer(params$lr_support_top_ligands)) %>%
pull(test_ligand)


lr_supported <- lr_network %>%
  filter(from %in% top_ligs, to %in% expressed_receptors) %>%
  distinct(from, to) %>%
  rename(ligand = from, receptor = to)

write_csv(lr_supported, file.path(rec_dir_dat, "ligand_receptor_supported_pairs.csv"))

if (nrow(lr_supported) > 0) {
  sender_means   <- avg_expr(exprs_all, sender_cells)
  receiver_means <- avg_expr(exprs_all, receiver_cells)

  lr_df <- lr_supported %>%
    mutate(
      ligand_avg_sender     = as.numeric(sender_means[ligand]),
      receptor_avg_receiver = as.numeric(receiver_means[receptor])
    ) %>%
    mutate(
      ligand_avg_sender     = if_else(is.na(ligand_avg_sender), 0, ligand_avg_sender),
      receptor_avg_receiver = if_else(is.na(receptor_avg_receiver), 0, receptor_avg_receiver)
    )

  lig_act_tbl <- ligand_activities %>%
    select(test_ligand, aupr_corrected) %>%
    rename(ligand = test_ligand)

  lr_plot <- lr_df %>%
    left_join(lig_act_tbl, by = "ligand") %>%
    mutate(score = rescale(aupr_corrected, to = c(0, 1)) * 0.7 + rescale(receptor_avg_receiver, to = c(0, 1)) * 0.3) %>%
    arrange(desc(score)) %>%
    slice_head(n = as.integer(params$lr_support_top_pairs))

  write_csv(lr_plot, file.path(rec_dir_dat, "ligand_receptor_support_top_pairs.csv"))

  lr_plot <- lr_plot %>%
    mutate(ligand = factor(ligand, levels = rev(unique(ligand))),
           receptor = factor(receptor, levels = unique(receptor)))

  p_lr <- ggplot(lr_plot, aes(x = receptor, y = ligand)) +
    geom_point(aes(size = receptor_avg_receiver, alpha = aupr_corrected)) +
    coord_flip() +
    labs(
      title = paste0("LR support (receiver: ", receiver_ct, ")"),
      subtitle = "Dot size: avg receptor expr in receiver; alpha: ligand activity (AUPR)",
      x = "Receptor (expressed in receiver)",
      y = "Ligand (top activity)",
      size = "Avg expr (receiver)",
      alpha = "AUPR"
    ) +
    theme(axis.text.y = element_text(size = 8),
          axis.text.x = element_text(size = 8))

  ggsave(file.path(rec_dir_fig, "ligand_receptor_support_dotplot.png"),
         p_lr, width = 10, height = 7, dpi = 300)
  print(p_lr)
}


}

# -----------------------------

# 3) Ligand-target heatmap (regulatory potential)

# -----------------------------

if (isTRUE(params$plot_ligand_target_heatmap)) {
top_ligs_hm <- ligand_activities %>% slice_head(n = as.integer(params$heatmap_ligands)) %>% pull(test_ligand)
top_ligs_hm <- intersect(top_ligs_hm, rownames(ligand_target_matrix))


top_targets <- function(lig, n = 40) {
  v <- ligand_target_matrix[lig, , drop = TRUE]
  names(sort(v, decreasing = TRUE))[1:min(n, length(v))]
}

targets_union <- unique(unlist(map(top_ligs_hm, top_targets, n = as.integer(params$heatmap_targets_per_ligand))))

lt_mat <- ligand_target_matrix[top_ligs_hm, targets_union, drop = FALSE]

# Prefer targets that are in the receiver program if enough overlap
lt_in_prog <- lt_mat[, colnames(lt_mat) %in% geneset_oi, drop = FALSE]
if (ncol(lt_in_prog) >= 10) lt_mat <- lt_in_prog

# Row-scale for visualization
lt_scaled <- t(scale(t(as.matrix(lt_mat))))
lt_scaled[is.na(lt_scaled)] <- 0

ht <- ComplexHeatmap::Heatmap(
  lt_scaled,
  name = "Z(RegPot)",
  column_title = paste0("Predicted targets (receiver: ", receiver_ct, ")"),
  row_title = "Top ligands",
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_column_names = FALSE
)

png(file.path(rec_dir_fig, "ligand_target_heatmap.png"), width = 2800, height = 1800, res = 300)
ComplexHeatmap::draw(ht)
dev.off()


}

# -----------------------------

# 4) Ligand-receptor heatmap (prior interaction potential)

# -----------------------------

# ---- Ligand-receptor heatmap (version-safe for nichenetr 2.2.1.1) ----
if (isTRUE(params$plot_ligand_receptor_heatmap)) {

  best_ligs <- ligand_activities %>%
    slice_head(n = as.integer(params$heatmap_ligands)) %>%
    pull(test_ligand)

  # This is the correct call for your version:
  lr_links_df <- nichenetr::get_weighted_ligand_receptor_links(
    best_upstream_ligands     = best_ligs,
    expressed_receptors       = expressed_receptors,
    lr_network                = lr_network,
    weighted_networks_lr_sig  = weighted_networks$lr_sig
  )

  # Ensure columns are exactly what prepare_* expects:
  lr_links_df <- lr_links_df %>% dplyr::select(from, to, weight)

  write_csv(lr_links_df, file.path(rec_dir_dat, "weighted_ligand_receptor_links.csv"))

  if (nrow(lr_links_df) > 0) {

    vis_lr <- nichenetr::prepare_ligand_receptor_visualization(
      lr_network_top_df_long = lr_links_df,
      best_upstream_ligands  = best_ligs,
      order_hclust           = "both"
    )

    mat_lr <- t(vis_lr)

    ht_lr <- ComplexHeatmap::Heatmap(
      mat_lr,
      name = "Prior\npotential",
      column_title = paste0("Receptors (receiver: ", receiver_ct, ")"),
      row_title = "Ligands",
      cluster_rows = TRUE,
      cluster_columns = TRUE
    )

    png(file.path(rec_dir_fig, "ligand_receptor_heatmap.png"), width = 2600, height = 1600, res = 300)
    ComplexHeatmap::draw(ht_lr)
    dev.off()
  }
}

# -----------------------------

# 5) Sender ligand expression dotplot (Seurat)

# -----------------------------

if (isTRUE(params$plot_sender_ligand_expression)) {
top_ligs <- ligand_activities %>%
slice_head(n = as.integer(params$heatmap_ligands)) %>%
pull(test_ligand) %>%
intersect(rownames(exprs_all))


if (length(top_ligs) >= 3) {
  obj_senders <- subset(seu, cells = sender_cells)
  Idents(obj_senders) <- obj_senders@meta.data[[celltype_col]]
  p <- DotPlot(obj_senders, features = rev(top_ligs)) + coord_flip() +
    labs(title = paste0("Sender ligand expression (receiver: ", receiver_ct, ")"),
         subtitle = paste0("Senders: ", paste(senders, collapse = ", ")))
  ggsave(file.path(rec_dir_fig, "sender_ligand_expression_dotplot.png"), p, width = 10, height = 7, dpi = 300)
  print(p)
}


}

# -----------------------------

# 6) Receiver receptor expression dotplot (Seurat)

# -----------------------------

if (isTRUE(params$plot_receiver_receptor_expression)) {
best_ligs <- ligand_activities %>% slice_head(n = as.integer(params$heatmap_ligands)) %>% pull(test_ligand)


# receptors paired with best ligands (supported by prior + expression)
rec_pairs <- lr_network %>%
  filter(from %in% best_ligs, to %in% expressed_receptors) %>%
  distinct(to) %>%
  pull(to) %>%
  intersect(rownames(exprs_all))

rec_pairs <- head(rec_pairs, 30)

if (length(rec_pairs) >= 3) {
  obj_rec <- subset(seu, cells = receiver_cells)
  Idents(obj_rec) <- obj_rec@meta.data[[condition_col]]
  p <- DotPlot(obj_rec, features = rev(rec_pairs)) + coord_flip() +
    labs(title = paste0("Receiver receptor expression (", receiver_ct, ")"),
         subtitle = paste0("Grouped by ", condition_col))
  ggsave(file.path(rec_dir_fig, "receiver_receptor_expression_dotplot.png"), p, width = 10, height = 7, dpi = 300)
  print(p)
}


}

# -----------------------------

# 7) Network graph (ligand -> receptor -> targets)

# -----------------------------

if (isTRUE(params$plot_network_graph)) {
best_ligs <- ligand_activities %>% slice_head(n = min(10, params$heatmap_ligands)) %>% pull(test_ligand)


# ligand->receptor edges (prior + expression)
lr_edges <- lr_network %>%
  filter(from %in% best_ligs, to %in% expressed_receptors) %>%
  distinct(from, to) %>%
  rename(from_node = from, to_node = to) %>%
  mutate(type = "ligand_receptor")

# ligand->targets (top targets per ligand, intersect program if possible)
top_targets_per_lig <- function(lig, n = 20) {
  if (!lig %in% rownames(ligand_target_matrix)) return(character(0))
  v <- ligand_target_matrix[lig, , drop = TRUE]
  tg <- names(sort(v, decreasing = TRUE))[1:min(n, length(v))]
  tg <- intersect(tg, geneset_oi)
  if (length(tg) < 5) tg <- names(sort(v, decreasing = TRUE))[1:min(n, length(v))]
  tg
}

lt_edges <- map_dfr(best_ligs, function(lig) {
  tg <- top_targets_per_lig(lig, n = 20)
  tibble(from_node = lig, to_node = tg, type = "ligand_target")
})

edges <- bind_rows(lr_edges, lt_edges) %>% filter(!is.na(to_node), nzchar(to_node))
if (nrow(edges) > 0) {
  g <- igraph::graph_from_data_frame(edges, directed = TRUE)

  V(g)$role <- case_when(
    V(g)$name %in% best_ligs ~ "ligand",
    V(g)$name %in% expressed_receptors ~ "receptor",
    TRUE ~ "target"
  )

  p_net <- ggraph(g, layout = "fr") +
    geom_edge_link(alpha = 0.4) +
    geom_node_point(aes(shape = role), size = 3) +
    geom_node_text(aes(label = name), repel = TRUE, size = 3) +
    labs(title = paste0("NicheNet network (receiver: ", receiver_ct, ")"),
         subtitle = "Ligands connect to expressed receptors and predicted targets") +
    theme_void()

  ggsave(file.path(rec_dir_fig, "nichenet_network_graph.png"), p_net, width = 10, height = 8, dpi = 300)
  print(p_net)

  write_csv(as_tibble(edges), file.path(rec_dir_dat, "network_edges.csv"))
}


}

# -----------------------------

# Store final summary row

# -----------------------------

niche_summary[[receiver_ct]] <- gene_set_summary
}
```

```{r summary}
niche_summary_df <- bind_rows(niche_summary)
write_csv(niche_summary_df, file.path("data/nichenet", "nichenet_summary.csv"))
write_csv(niche_summary_df, file.path("data", "nichenet_summary.csv"))

niche_summary_df

```

```{r}
out_rds <- paste0(params$project_name, "_nichenet_done.rds")
saveRDS(list(done=TRUE,
receivers=receivers,
receiver_program_mode=receiver_program_mode,
timestamp=Sys.time()),
file = out_rds)
out_rds
```

```{r}
sessionInfo()
```

