---
title: "Module - Azimuth cell typing"
description: | 
  Description
execute:
  freeze: auto
  cache: false
  fig-path: figures/
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: 'Project'
  seurat_object: 'SCRATCH-Annotation/data/Yost_main_cluster_object.RDS'
  reference_object: 'SCRATCH-Annotation/assets/ref.Rds'
  n_threads: 8
  n_memory: 16
  work_directory: "." 
  auto_save: TRUE
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup}
#| include: false

# Project parameters 
project_name <- params$project_name
seurat_object <- params$seurat_object
reference_object <- params$reference_object
work_directory <- if (!is.null(params$work_directory)) params$work_directory else "."
  
# Optional parameters

# Dataflow/Computational parameters
n_threads <- params$n_threads
n_memory <- params$n_memory

# Output parameters
timestamp <- params$timestamp
auto_save <- params$auto_save

# Rmarkdown options
knitr::opts_knit$set(
  root.dir = work_directory
  )

knitr::opts_chunk$set(
  fig.path = "figures/",
  dev = "png",
  dpi = 300
)
```

:::{.callout-important collapse="true"}
Here we will print the parameters
:::

## Loading library

```{r}
#| output: false
options(lifecycle_verbosity = "quiet")
library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(Azimuth)
library(SCP)
library(BiocParallel)
register(SerialParam()) 
```

## Folder structure

```{r project_directory}
#| output: false

if(!dir.exists(work_directory)) {
  dir.create(work_directory, recursive = T)
}

for(sub_directory in c('data', 'figures')) {
  dir.create(paste0(work_directory, '/', sub_directory))
}

```

```
## Loading Seurat object

```{r LoadingDta}

# Reading seurat object
seurat_object <- readRDS(file = seurat_object)

# Reading reference object
reference_object <- readRDS(file = reference_object)

# Ensure reference has a 'seurat_annotation' column
meta_cols <- colnames(reference_object@meta.data)

if (!"seurat_annotation" %in% meta_cols) {
  if ("seurat_annotations" %in% meta_cols) {
    # create an alias so downstream code works unchanged
    reference_object$seurat_annotation <- reference_object$seurat_annotations
  } else {
    stop(
      "Reference object has no 'seurat_annotation' or 'seurat_annotations' column. ",
      "Available metadata columns are: ",
      paste(meta_cols, collapse = ", ")
    )
  }
}



# Setting random seed
random_seed <- 2203
set.seed(random_seed)

```
## Projecting annotation

```{r Anchors}
#| output: false

# find anchors
anchors <- FindTransferAnchors(reference = reference_object, query = seurat_object)

# transfer labels
predictions <- TransferData(
  anchorset = anchors,
  refdata = reference_object$seurat_annotations
)

```

## Adding metadata

```{r RenamingAnnotation}

# Renaming annotation column
predictions <- predictions %>%
  dplyr::rename(
    azimuth_labels = predicted.id
  )

# Adding labels
seurat_object <- AddMetaData(
  object = seurat_object, metadata = predictions)


```

## Data Visualization

```{r AnnotationUMAP}
#| label: fig-azimuth-annotation-umap
#| fig-cap: "Azimuth annotation UMAP"
#| fig-width: 12
#| fig-height: 9

annotation_UMAP <- SCP::CellDimPlot(
  srt = seurat_object,
  group.by = "azimuth_labels",
  reduction = "umap",
  label = TRUE,
  label_repel = TRUE,
  show_stat = TRUE,
  theme_use = "theme_blank"
) +
  theme(legend.position = "right")

annotation_UMAP 

```


```{r CellDimPlot}
#| label: fig-azimuth-celldimplot-split
#| fig-cap: "Azimuth split UMAP (one panel per label)"
#| fig-width: 14
#| fig-height: 9

p_split <- SCP::CellDimPlot(
  srt = seurat_object,
  group.by = "azimuth_labels",
  split.by = "azimuth_labels",
  reduction = "umap",
  label = FALSE,
  label_repel = FALSE,
  show_stat = TRUE,
  legend.position = "none",
  theme_use = "theme_blank"
) +
  theme(legend.position = "right")

p_split
```


```{r FeatureDimPlot}
#| label: fig-azimuth-featuredimplot-predscore
#| fig-cap: "Azimuth mapping score (prediction.score.max)"
#| fig-width: 12
#| fig-height: 9

p_score <- SCP::FeatureDimPlot(
  srt = seurat_object,
  features = "prediction.score.max",
  reduction = "umap",
  label = FALSE,
  label_repel = FALSE,
  show_stat = TRUE,
  theme_use = "theme_blank"
)

p_score 
```


```{r Assay}

#

seurat_v4 <- seurat_object
seurat_v4[["RNA"]] <- as(object = seurat_v4[["RNA"]], Class = "Assay")

#

#SCP::FeatureStatPlot(
#  srt = seurat_v4, 
#  group.by = "azimuth_labels",
#  bg.by = "azimuth_labels",
#  stat.by = "prediction.score.max",
#  add_box = TRUE,
#  add_line = 0.2, 
#  line_type = 2
#) +
#  labs(
#    y = "Prediction score"
#  )
Cairo::CairoPNG(filename = "figures/azimuth_prediction_scores.png", width = 1800, height = 1400, res = 300)
ht <- SCP::FeatureStatPlot(
  srt = seurat_v4, 
  group.by = "azimuth_labels",
  bg.by = "azimuth_labels",
  stat.by = "prediction.score.max",
  add_box = TRUE,
  add_line = 0.2, 
  line_type = 2
) +
  labs(y = "Prediction score")
print(ht)
dev.off()

```

## Azimuth dendrogram

```{r annotation_tree}
#| warning: false

# Aggregate data by annotations
cell_average <- AverageExpression(
  seurat_v4, 
  assay = 'RNA',
  layer = 'scale.data',
  return.seurat = FALSE, 
  group.by = "azimuth_labels"
  )

cell_average <- cell_average$RNA

# Compute distance matrix on aggregated data
cell_dist_matrix <- as.matrix(cell_average)
cell_dist_matrix <- stats::cor(cell_dist_matrix, method = "pearson")

# Convert correlation matrix to a distance matrix
cell_dist_matrix <- stats::as.dist(1 - cell_dist_matrix)

# Perform hierarchical clustering
cell_hierarchical <- stats::hclust(cell_dist_matrix)

```

```{r figDendrogram}
#| fig-cap: 'Dendrogram'
#| fig-width: 10
#| fig-height: 12
#| fig-align: center

# Printing dendrogram
Cairo::CairoPNG(filename = "figures/FigDendogram.png", width = 1800, height = 1400, res = 300)
FigDendogram.png <- plot(cell_hierarchical, main = "Dendrogram", xlab = "Index", ylab = "Height", hang = -1)
print(FigDendogram.png)
dev.off()

```

```{r DE1}
message(">>> [DE] sanity checks at: ", Sys.time())

if (!exists("seurat_v4")) {
stop(">>> [DE] ERROR: object 'seurat_v4' not found. Did the previous chunk defining it run?")
}

message("Object size (seurat_v4): ", format(object.size(seurat_v4), units = "GB"))
message("Number of cells: ", ncol(seurat_v4))
message("Number of genes: ", nrow(seurat_v4))
message("Meta.data columns: ", paste(colnames(seurat_v4@meta.data), collapse = ", "))

if (!"azimuth_labels" %in% colnames(seurat_v4@meta.data)) {
stop(
">>> [DE] ERROR: column 'azimuth_labels' not found in meta.data. ",
"Available columns: ", paste(colnames(seurat_v4@meta.data), collapse = ", ")
)
}



```


```{r FindAllMarkers}
message(">>> [DE] running FindAllMarkers for azimuth_labels at: ", Sys.time())

seurat_annotation <- seurat_v4
Idents(seurat_annotation) <- "azimuth_labels"

all_markers_azimuth <- tryCatch(
{
FindAllMarkers(
object = seurat_annotation,
assay = "RNA",
only.pos = FALSE,
test.use = "wilcox",
logfc.threshold = 0.25,
min.pct = 0.1
)
},
error = function(e) {
message(">>> [DE] FindAllMarkers ERROR: ", conditionMessage(e))
return(NULL)
}
)

if (is.null(all_markers_azimuth)) {
message(">>> [DE] FindAllMarkers returned NULL; downstream DE plots will be skipped.")
} else {
message(">>> [DE] Total markers (azimuth_labels): ", nrow(all_markers_azimuth))



saveRDS(
all_markers_azimuth,
file = file.path("data", paste0(project_name, "_DE_azimuth_all.rds"))
)

DE_azimuth_sig <- subset(
all_markers_azimuth,
avg_log2FC > 1 & p_val_adj < 0.05
)
message(">>> [DE] Significant markers (log2FC > 1 & padj < 0.05): ", nrow(DE_azimuth_sig))

saveRDS(
DE_azimuth_sig,
file = file.path("data", paste0(project_name, "_DE_azimuth_sig.rds"))
)
}
```

```{r azimuth_de_tables}
#| message: true

if (exists("all_markers_azimuth") &&
    is.data.frame(all_markers_azimuth) &&
    nrow(all_markers_azimuth) > 0) {

  top_markers <- all_markers_azimuth %>%
    dplyr::group_by(cluster) %>%
    dplyr::slice_max(order_by = avg_log2FC, n = 5, with_ties = FALSE) %>%
    dplyr::ungroup()

  # Save tables
  readr::write_csv(top_markers, file.path("data", paste0(project_name, "_azimuth_top_markers_fc.csv")))
  saveRDS(top_markers, file.path("data", paste0(project_name, "_azimuth_top_markers_fc.rds")))

  message("Top markers per azimuth label:")
  print(table(top_markers$cluster))

  message("Number of unique top marker genes:")
  print(length(unique(top_markers$gene)))

  # Print a preview in HTML
  print(utils::head(top_markers, 20))

} else {
  message(">>> [DE_top_markers_summary] No markers available; skipping.")
}

if (exists("DE_azimuth_sig") &&
    is.data.frame(DE_azimuth_sig) &&
    nrow(DE_azimuth_sig) > 0) {

  top_markers_heat <- DE_azimuth_sig %>%
    dplyr::group_by(cluster) %>%
    dplyr::slice_max(order_by = avg_log2FC, n = 10, with_ties = FALSE) %>%
    dplyr::ungroup()

  # Save tables
  readr::write_csv(top_markers_heat, file.path("data", paste0(project_name, "_azimuth_sig_top10_fc.csv")))
  saveRDS(top_markers_heat, file.path("data", paste0(project_name, "_azimuth_sig_top10_fc.rds")))

  message("Saved top markers for heatmap: ", nrow(top_markers_heat))

} else {
  message(">>> [DE_azimuth_heatmap] No significant DE markers; skipping tables.")
}
```

```{r fig-azimuth-de-heatmap}
#| fig-cap: "Azimuth DE heatmap (top markers per label)"
#| fig-width: 14
#| fig-height: 10
#| fig-path: "figures/azimuth_de_heatmap_"
#| dev: pdf

if (exists("DE_azimuth_sig") &&
    is.data.frame(DE_azimuth_sig) &&
    nrow(DE_azimuth_sig) > 0 &&
    exists("top_markers_heat") &&
    nrow(top_markers_heat) > 0) {

  p_heat <- DoHeatmap(
    object  = seurat_annotation,
    features = unique(top_markers_heat$gene),
    group.by = "azimuth_labels"
  ) +
    scale_fill_viridis_c(option = "B") +
    theme(
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 6),
      axis.text.y = element_text(size = 6)
    )

  p_heat
} else {
  plot.new()
  text(0.5, 0.5, "No significant DE markers available for heatmap.")
}

```


```{r fig-azimuth-de-dotplot}
#| fig-cap: "Azimuth DE dotplot (top markers per label)"
#| fig-width: 14
#| fig-height: 8
#| fig-path: "figures/azimuth_de_dotplot_"
#| dev: pdf

if (exists("all_markers_azimuth") &&
    is.data.frame(all_markers_azimuth) &&
    nrow(all_markers_azimuth) > 0) {

  top_markers_dot <- all_markers_azimuth %>%
    dplyr::group_by(cluster) %>%
    dplyr::slice_max(order_by = avg_log2FC, n = 5, with_ties = FALSE) %>%
    dplyr::ungroup()

  # Save table backing the plot
  readr::write_csv(top_markers_dot, file.path("data", paste0(project_name, "_azimuth_dotplot_top5_fc.csv")))
  saveRDS(top_markers_dot, file.path("data", paste0(project_name, "_azimuth_dotplot_top5_fc.rds")))

  p_dot <- DotPlot(
    seurat_annotation,
    features = unique(top_markers_dot$gene),
    group.by = "azimuth_labels"
  ) +
    RotatedAxis() +
    theme(axis.text.x = element_text(size = 6),
          axis.text.y = element_text(size = 6))

  p_dot
} else {
  plot.new()
  text(0.5, 0.5, "No marker table available for dotplot.")
}

```


```{r MarkerBasedVisualization}
if (exists("all_markers_azimuth") &&
is.data.frame(all_markers_azimuth) &&
nrow(all_markers_azimuth) > 0) {

top_markers <- all_markers_azimuth %>%
dplyr::group_by(cluster) %>%  # 'cluster' here is azimuth_labels
dplyr::slice_max(order_by = avg_log2FC, n = 5, with_ties = FALSE)

message("Top markers per azimuth label:")
print(table(top_markers$cluster))

message("Number of unique top marker genes:")
print(length(unique(top_markers$gene)))
} else {
message(">>> [DE_top_markers_summary] No markers available; skipping summary.")
}

if (exists("DE_azimuth_sig") &&
is.data.frame(DE_azimuth_sig) &&
nrow(DE_azimuth_sig) > 0) {

message(">>> [DE_azimuth_heatmap] building heatmap")

top_markers_heat <- DE_azimuth_sig %>%
dplyr::group_by(cluster) %>% # 'cluster' = azimuth_labels
dplyr::slice_max(order_by = avg_log2FC, n = 10, with_ties = FALSE)

DoHeatmap(
object = seurat_annotation,
features = unique(top_markers_heat$gene),
group.by = "azimuth_labels"
) +
scale_fill_viridis_c(option = "B") +
theme(
axis.text.x = element_text(
angle = 90,
vjust = 0.5,
hjust = 1,
size = 6
),
axis.text.y = element_text(size = 6)
)
} else {
message(">>> [DE_azimuth_heatmap] No significant DE markers; skipping heatmap.")
}

if (exists("all_markers_azimuth") &&
is.data.frame(all_markers_azimuth) &&
nrow(all_markers_azimuth) > 0) {

message(">>> [DE_azimuth_dotplot] building DotPlot")

top_markers_dot <- all_markers_azimuth %>%
dplyr::group_by(cluster) %>%
dplyr::slice_max(order_by = avg_log2FC, n = 5, with_ties = FALSE)

DotPlot(
seurat_annotation,
features = unique(top_markers_dot$gene),
group.by = "azimuth_labels"
) +
RotatedAxis() +
theme(
axis.text.x = element_text(size = 6),
axis.text.y = element_text(size = 6)
)
} else {
message(">>> [DE_azimuth_dotplot] No markers table available; skipping DotPlot.")
}

```

```{r CellComposition}
#| fig-cap: 'Cell composition by Azimuth labels and Seurat clusters'
#| fig-width: 18
#| fig-height: 10
#| fig-align: center

Cairo::CairoPNG(filename = "figures/CellComposition.png", width = 1800, height = 1400, res = 300)

CP <-SCP::CellStatPlot(
seurat_object,
stat.by  = "seurat_clusters",
group.by = "azimuth_labels",
label    = TRUE
)
print(CP)
dev.off()

```

```{r Savinf AzumuthAnnotaitonObject}
#| cache: false

if (isTRUE(auto_save)) {
saveRDS(
seurat_object,
file = file.path("data", paste0(project_name, "_Azimuth_annotation_object.RDS"))
)
}

```


## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

print(params)

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```
