---
title: "Module - Sctype Subtype/State annotation"
description: | 
  Description
execute:
  freeze: auto
  cache: false
  fig-path: figures/
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-format: png
    fig-dpi: 300
    fig-responsive: true
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: 'Project'
  seurat_object: '/Users/affaustino/Projects/SCRATCH-Annotation/data/Project_major_annotation_object.RDS'
  input_cell_markers_db: 'assets/cell_markers_database.csv'
  input_parent_level: 'B_Plasma_Cells'
  thr_n_features: 2000
  input_deg_method: 'wilcox'
  thr_npc: 25
  thr_resolution: 0.25
  thr_min_percentage: 0.1
  thr_fold_change: 0.25
  n_threads: 8
  n_memory: 16
  work_directory: !expr here::here()
  auto_save: TRUE
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Applied globally, so
# the plotting code below needs no changes. Guarded: if the library is not
# staged, the notebook renders exactly as it did before.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup}
#| include: false

# Project parameters 
project_name <- params$project_name
seurat_object <- params$seurat_object

# Inputs and thresholds
input_cell_markers_db <- params$input_cell_markers_db
input_parent_level <- strsplit(
  params$input_parent_level, split = ';')[[1]]

input_deg_method <- params$input_deg_method
thr_n_features <- params$thr_n_features
thr_npc <- params$thr_npc
thr_resolution <- params$thr_resolution
thr_min_percentage <- params$thr_min_percentage
thr_fold_change <- params$thr_fold_change

# Optional parameters

# Dataflow/Computational parameters
n_threads <- params$n_threads
n_memory <- params$n_memory

# Output parameters
work_directory <- params$work_directory
timestamp <- params$timestamp
auto_save <- params$auto_save


if(!dir.exists(work_directory)) {
  dir.create(work_directory, recursive = T)
}

# for(sub_directory in c('data', 'figures')) {
#   dir.create(paste0(work_directory, '/', sub_directory))
# }
dir.create(file.path(work_directory, "figures"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(work_directory, "data"), recursive = TRUE, showWarnings = FALSE)


# Rmarkdown options
knitr::opts_knit$set(
  root.dir = work_directory
  )

knitr::opts_chunk$set(
  fig.path = "figures/",
  dev = "png",
  dpi = 300
)
```

:::{.callout-important collapse="true"}
Here we will print the parameters
:::

## Loading library

```{r}
#| output: false

library(readr)
library(dplyr)
library(ggplot2)
library(Seurat)
library(HGNChelper)

```

```{r}
#| include: false

gene_sets_prepare <- function(path_to_db_file, cell_type){

  cell_markers = openxlsx::read.xlsx(path_to_db_file)
  cell_markers = cell_markers[cell_markers$tissueType == cell_type,]
  cell_markers$geneSymbolmore1 = gsub(" ","",cell_markers$geneSymbolmore1); cell_markers$geneSymbolmore2 = gsub(" ","",cell_markers$geneSymbolmore2)

  # correct gene symbols from the given DB (up-genes)
  cell_markers$geneSymbolmore1 = sapply(1:nrow(cell_markers), function(i){

    markers_all = gsub(" ", "", unlist(strsplit(cell_markers$geneSymbolmore1[i],",")))
    markers_all = toupper(markers_all[markers_all != "NA" & markers_all != ""])
    markers_all = sort(markers_all)

    if(length(markers_all) > 0){
      suppressMessages({markers_all = unique(na.omit(checkGeneSymbols(markers_all)$Suggested.Symbol))})
      paste0(markers_all, collapse=",")
    } else {
      ""
    }
  })

  # correct gene symbols from the given DB (down-genes)
  cell_markers$geneSymbolmore2 = sapply(1:nrow(cell_markers), function(i){

    markers_all = gsub(" ", "", unlist(strsplit(cell_markers$geneSymbolmore2[i],",")))
    markers_all = toupper(markers_all[markers_all != "NA" & markers_all != ""])
    markers_all = sort(markers_all)

    if(length(markers_all) > 0){
      suppressMessages({markers_all = unique(na.omit(checkGeneSymbols(markers_all)$Suggested.Symbol))})
      paste0(markers_all, collapse=",")
    } else {
      ""
    }
  })

  cell_markers$geneSymbolmore1 = gsub("///",",",cell_markers$geneSymbolmore1)
  cell_markers$geneSymbolmore1 = gsub(" ","",cell_markers$geneSymbolmore)
  
  cell_markers$geneSymbolmore2 = gsub("///",",",cell_markers$geneSymbolmore2)
  cell_markers$geneSymbolmore2 = gsub(" ","",cell_markers$geneSymbolmore)

  gs = lapply(1:nrow(cell_markers), function(j) gsub(" ","",unlist(strsplit(toString(cell_markers$geneSymbolmore1[j]),",")))) 
  names(gs) = cell_markers$cellName
  
  gs2 = lapply(1:nrow(cell_markers), function(j) gsub(" ","",unlist(strsplit(toString(cell_markers$geneSymbolmore2[j]),",")))) 
  names(gs2) = cell_markers$cellName

  list(gs_positive = gs, gs_negative = gs2)
}

sctype_score <- function(scRNAseqData, scaled = !0, gs, gs2 = NULL, gene_names_to_uppercase = !0, ...){

  # check input matrix
  if(!is.matrix(scRNAseqData)){
    warning("scRNAseqData doesn't seem to be a matrix")
  } else {
    if(sum(dim(scRNAseqData)) == 0){
      warning("The dimension of input scRNAseqData matrix equals to 0, is it an empty matrix?")
    }
  }

  # marker sensitivity
  marker_stat = sort(table(unlist(gs)), decreasing = T);
  marker_sensitivity = data.frame(score_marker_sensitivity = scales::rescale(as.numeric(marker_stat), to = c(0,1), from = c(length(gs),1)),
                                      gene_ = names(marker_stat), stringsAsFactors = !1)

  # convert gene names to Uppercase
  if(gene_names_to_uppercase){
    rownames(scRNAseqData) = toupper(rownames(scRNAseqData));
  }

  # subselect genes only found in data
  names_gs_cp = names(gs); names_gs_2_cp = names(gs2);
  gs = lapply(1:length(gs), function(d_){
    GeneIndToKeep = rownames(scRNAseqData) %in% as.character(gs[[d_]]); rownames(scRNAseqData)[GeneIndToKeep]})
  
  gs2 = lapply(1:length(gs2), function(d_){
    GeneIndToKeep = rownames(scRNAseqData) %in% as.character(gs2[[d_]]); rownames(scRNAseqData)[GeneIndToKeep]})
  
  names(gs) = names_gs_cp; names(gs2) = names_gs_2_cp;
  cell_markers_genes_score = marker_sensitivity[marker_sensitivity$gene_ %in% unique(unlist(gs)),]

  # z-scale if not
  if(!scaled) Z <- t(scale(t(scRNAseqData))) else Z <- scRNAseqData

  # multiple by marker sensitivity
  for(jj in 1:nrow(cell_markers_genes_score)){
    Z[cell_markers_genes_score[jj,"gene_"], ] = Z[cell_markers_genes_score[jj,"gene_"], ] * cell_markers_genes_score[jj, "score_marker_sensitivity"]
  }

  # subselect only with marker genes
  Z = Z[unique(c(unlist(gs),unlist(gs2))), ]

  # combine scores
  es = do.call("rbind", lapply(names(gs), function(gss_){
    sapply(1:ncol(Z), function(j) {
      gs_z = Z[gs[[gss_]], j]; gz_2 = Z[gs2[[gss_]], j] * -1
      sum_t1 = (sum(gs_z) / sqrt(length(gs_z))); sum_t2 = sum(gz_2) / sqrt(length(gz_2));
      if(is.na(sum_t2)){
        sum_t2 = 0;
      }
      sum_t1 + sum_t2
    })
  }))

  dimnames(es) = list(names(gs), colnames(Z))
  es.max <- es[!apply(is.na(es) | es == "", 1, all),] # remove na rows

  es.max
}

```

## Folder structure



## Loading Seurat object

```{r loading_rds}

# Reading seurat object
seurat_object <- readRDS(file = seurat_object)

# Setting random seed
random_seed <- 2203
set.seed(random_seed)

```

```{r}

# Extracting cells for downstream annotation
seurat_object <- subset(
  seurat_object,
  subset = sctype == input_parent_level
)

```

```{r loading_markers}

# Get cell-type-specific gene sets from our in-built database (DB)
cell_markers_database <- read_csv(file = input_cell_markers_db, show_col_types = FALSE)
cell_markers_database <- cell_markers_database %>%
  filter(parent_level == input_parent_level) %>%
  as.data.frame()

# Adding group@state label
cell_markers_database <- cell_markers_database %>%
  rowwise() %>%
  mutate(
    group_state = paste0(cell_group, "@", cell_annotation)
  )

# Formatting Cellmarkers database
gene_list_db <- list(
  'gs_positive' = split(cell_markers_database$markers, f = cell_markers_database$group_state)
)

```

## Counts normalization

```{r normalization}
#| output: false

seurat_object <- NormalizeData(
  object = seurat_object, 
  normalization.method = "LogNormalize",
  scale.factor = 10000)

# Retrieving HVG
seurat_object <- FindVariableFeatures(
  seurat_object, selection.method = "vst", nfeatures = thr_n_features)

```

## Filtering and scaling HVG

```{r}

# Moving clonotye expansion signal
filtering_vst_markers <- list(
  "T_Cells" = "^TR[ABDG][VDJ]",
  "B_Plasma_Cells" = "IGH[KL]V.*"
)

if(input_parent_level %in% names(filtering_vst_markers)) {
  
  VariableFeatures(seurat_object) <- grep(
    filtering_vst_markers[[input_parent_level]], VariableFeatures(seurat_object), value = TRUE, invert = TRUE)
  
}

# Scaling the dataset
seurat_object <- ScaleData(seurat_object)

```

## Dimensionality Reduction

```{r dimensionality_reduction}
#| output: false

# Running PCA and printing out dimensions
seurat_object <- RunPCA(seurat_object, verbose = FALSE, npcs = 50)
print(seurat_object[["pca"]], dims = 1:5, nfeatures = 5)

```

## Finding neighbors and clusters

```{r finding_neighbors}
#| output: false

# Finding clusters
seurat_object <- FindNeighbors(
  object = seurat_object, 
  reduction = 'pca', 
  dims = 1:thr_npc,
  graph.name = paste0('snn_npc', thr_npc)
)

seurat_object <- FindClusters(
  object = seurat_object,
  graph.name = paste0('snn_npc', thr_npc),
  resolution = thr_resolution
)

# Running UMAP reduction
seurat_object <- RunUMAP(
  object = seurat_object,
  reduction = 'pca', 
  dims = 1:thr_npc,
  seed.use = random_seed
)

```

## Cell annotation

```{r sctype_annotation}
#| code-fold: true

# Load example scRNA-seq matrix
seurat_scaled <- LayerData(
  seurat_object,
  layer = "scale.data"
)

# Calculating scType score
sctype_calculation <- sctype_score(
  scRNAseqData = seurat_scaled,
  scaled = TRUE,
  gs = gene_list_db$gs_positive,
  gs2 = NULL
  )

# Merge by cluster
sctype_results <- do.call("rbind", lapply(unique(seurat_object@meta.data$seurat_clusters), function(cl) {

    es.max.cl <- sort(
      rowSums(sctype_calculation[, rownames(seurat_object@meta.data[seurat_object@meta.data$seurat_clusters == cl, ])]), decreasing = !0)

    head(
      data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(seurat_object@meta.data$seurat_clusters==cl)), 10)
    
    }
  )
)

# Retrieving best-scored results per cluster
sctype_best_scored <- sctype_results %>%
  group_by(cluster) %>%
  top_n(n = 1, wt = scores)

# Set low-confident clusters to "unknown"
sctype_best_scored$type[
  as.numeric(as.character(sctype_best_scored$scores)) < sctype_best_scored$ncells/4] <- "Unknown"

# Adding labels to Seurat object
seurat_object@meta.data$sctype <- NA
for(cl in unique(sctype_best_scored$cluster)) {

  cl_type <- sctype_best_scored[sctype_best_scored$cluster == cl, ]
  seurat_object@meta.data$sctype[
    seurat_object@meta.data$seurat_clusters == cl] <- as.character(cl_type$type[1])

}

```

## Cell composition

```{r sctype_classification}
#| code-fold: true

# View results
sctype_best_scored_ordered <- sctype_best_scored %>%
  arrange(cluster)

# Adding the markers information
cell_list <- cell_markers_database[
  cell_markers_database$group_state %in% sctype_best_scored_ordered$type, c("cell_group", "cell_annotation", "group_state", "markers")]

cell_list <- cell_list %>%
  group_by(cell_group, cell_annotation, group_state) %>%
  summarize(signature = paste(markers, collapse = ","))

cell_list <- cell_list %>%
  dplyr::rename(
    type = group_state
  )

sctype_best_scored_ordered <- merge(
  sctype_best_scored_ordered,
  cell_list,
  by = "type"
)

sctype_best_scored_ordered <- sctype_best_scored_ordered %>%
  select(cell_group, type, cluster, scores, signature)

```

```{r}
#| column: screen-inset-shaded
# Printing output table

DT::datatable(
  sctype_best_scored_ordered,
  class = 'cell-border stripe',
  options = list(dom = 'tip')
)

```

## Data Visualization

```{r annotation_umap}
#| label: fig-umap-state
#| fig-width: 15
#| fig-height: 10
#| fig-align: center

p <- SCP::CellDimPlot(
  srt = seurat_object, 
  group.by = "sctype",
  reduction = "umap",
  label = TRUE, 
  label_repel = TRUE,
  show_stat = TRUE,
  theme_use = "theme_blank"
) +
  theme(
    legend.position = "bottom"
  )
print(p)
```

:::{.callout-tip}
The @fig-umap-state displays a substype/state cell type annotation. The accuracy might change depending on co-confounding factors, and clustering resolution. For inspecting the results, please check the Differential expression analysis.
:::

## Module-score

```{r calculating_module_score}

# Calculating ModuleScore
non_malignant_markers <- gene_list_db$gs_positive

# Avoiding redundant markers
non_malignant_markers_index <-
  names(non_malignant_markers) %in% colnames(seurat_object@meta.data)
non_malignant_markers_valid <- non_malignant_markers[!non_malignant_markers_index]

# UCell is a different package from Seurat and may or may not hit the same
# defunct GetAssayData(slot=) path under SeuratObject 5.x — it was not installed
# on the machine where this was checked, so it is UNVERIFIED. Rather than change
# the scoring method blind, keep UCell as the primary and fall back only if it
# actually errors. UCell's rank-based scores are not identical to the Tirosh
# method, so the fallback is a degraded path, not an equivalent one: it is
# announced loudly.
seurat_object <- tryCatch(
  UCell::AddModuleScore_UCell(
    seurat_object,
    ncores = n_threads,
    features = non_malignant_markers_valid
  ),
  error = function(e) {
    if (!exists("scratch_module_score")) stop(e)
    warning("UCell::AddModuleScore_UCell failed (", conditionMessage(e),
            "). Falling back to scratch_module_score — NOTE these are rank-based ",
            "vs mean-difference scores and are NOT numerically comparable to a ",
            "UCell run. Thresholds tuned on UCell output may not transfer.",
            immediate. = TRUE)
    scratch_module_score(seurat_object, features = non_malignant_markers_valid,
                         name = "UCell_")
  }
)

colnames(seurat_object@meta.data) <- 
  gsub('_UCell$', '', colnames(seurat_object@meta.data))

```

### Feature Plot

```{r feature_plot}
#| label: fig-feat-markers
#| fig-width: 10
#| fig-height: 12
#| fig-align: center

fp <- FeaturePlot(
  object = seurat_object,
  features = names(non_malignant_markers),
  label = TRUE,
  ncol = 2,
  min.cutoff = 'q1'
)

print (fp)
```

:::{.callout-tip}
The @fig-feat-markers represents a 'FeaturePlot' which allows coloring the cells based on a categorical variable, e.g., module score associated with cell lineage marker or state signatures.
:::

### Violin Plot

```{r violin_plot}
#| label: fig-violin
#| fig-width: 14
#| fig-height: 20
#| fig-align: center

vp <- VlnPlot(
  object = seurat_object,
  ncol = 2,
  features = unique(seurat_object$sctype)
)

print (vp)
```

:::{.callout-tip}
The @fig-violin displays the overall distribution associated with a specific categorical variable across cell groups, such as gene sets corresponding to cell types or labels.
:::

## BubblePlot

```{r}
#| label: fig-bubble
#| fig-width: 18
#| fig-height: 10
#| fig-align: center
#| column: screen-inset-shaded

non_malignant_markers_list <- unique(unlist(non_malignant_markers, use.names = FALSE))

annot_dotplot <- DotPlot(seurat_object, features = non_malignant_markers_list, group.by = "sctype") +
  theme(
    axis.text.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.9),
    legend.position = 'top'
  )

print (annot_dotplot)

```

:::{.callout-tip}
The @fig-bubble enables a visual inspection of gene expression levels across different clusters, highlighting both the percentage of cells expressing a gene and the average expression level within each cluster.
:::

## Differential expression

```{r differential}
#| warning: false

all_markers <- FindAllMarkers(
  object = seurat_object,
  only.pos = FALSE, 
  test.use = input_deg_method,
  logfc.threshold = thr_fold_change,
  random.seed = random_seed,
  min.pct = thr_min_percentage
  )

```

```{r deg_table, message = FALSE, warning = FALSE, echo = TRUE}
#| column: screen-inset-shaded

# Reordering all markers data.frame
all_markers <- all_markers %>%
  select(cluster, gene, pct.1, pct.2, p_val, avg_log2FC)

# Printing out all markers
DT::datatable(all_markers, 
    class = 'cell-border stripe',
    rownames = FALSE,
    filter = 'top',
    width = '100%', 
    options = list(scrollX = TRUE)
    )

```

## Cluster composition per cell type

```{r cluster_composition}

# Retrieving cluster composition
seurat_cluster_composition <- seurat_object@meta.data %>%
  mutate(
    total_cell_study = n()
  )

seurat_cluster_composition <- seurat_cluster_composition %>%
  group_by(seurat_clusters) %>%
  mutate(
    total_cell_per_cluster = n()
  )

seurat_cluster_composition <- seurat_cluster_composition %>%
  group_by(sctype) %>%
  mutate(
    total_cell_per_type = n()
  )

seurat_cluster_composition <- seurat_cluster_composition %>%
  distinct(seurat_clusters, sctype, total_cell_study, total_cell_per_cluster, total_cell_per_type)

# Cell percentage per cluster
seurat_cluster_composition <- seurat_cluster_composition %>%
  group_by(sctype) %>%
  mutate(
    percentage = round(((total_cell_per_cluster * 100) / total_cell_per_type), digits = 2)
  )

```

```{r barplot_composition}
#| label: fig-bar-composition
#| fig-width: 18
#| fig-height: 10
#| fig-align: center

# Boxplot
csp <- SCP::CellStatPlot(
  seurat_object,
  stat.by = "patient_id", 
  group.by = "sctype", 
  label = TRUE
)

print (csp)
```

:::{.callout-tip}
The @fig-bar-composition shows stacked violin plots illustrate gene expression distributions for multiple genes across cell types.
:::

## Saving Seurat object

```{r object_dump}
#| cache: false

if(auto_save) {

  saveRDS(seurat_object, file = paste0('./data/', project_name, '_', input_parent_level, '_annotation_object.RDS'))

  annotation_table <- seurat_object@meta.data %>%
    select(seurat_clusters, sctype) %>%
    tibble::rownames_to_column(var = "barcode")

  write_csv(annotation_table, file = paste0("data/", project_name, "_", input_parent_level, "_annotation.csv"))

}

```

## Parameters log

```{r params_log, message = FALSE, warning = FALSE, echo = FALSE}

print(params)

```

## Session info

```{r session_info, message = FALSE, warning = FALSE, echo = FALSE}

sessionInfo()

```
