---
title: "Module - CNV caller concordance"
description: |
  inferCNV and Numbat, cell by cell: where they agree, where they do not, and
  what the disagreement is made of.
execute:
  freeze: auto
  cache: false
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    embed-resources: true
    code-fold: show
    code-tools: true
    df-print: paged
    fig-align: center
lightbox:
    match: auto
    effect: zoom
    loop: true
params:
  project_name: 'Test'
  seurat_object: 'data/project_Azimuth_annotation_object.RDS'
  caller_inputs: 'caller_inputs'
  celltype_col: 'celltype'
  sample_col: 'sample_id'
  reference_labels: "T_Cells;NK_Cells;Myeloid;B_Plasma_Cells;Endothelial_Cells;Fibroblast"
  infercnv_sd_threshold: 3
  work_directory: !expr here::here()
  auto_save: TRUE
---

```{r scratch-viz-setup, include=FALSE}
# Publication figure system: house theme, validated colour-vision-safe palette,
# and vector (PDF) output alongside every raster figure. Guarded: if the library
# is not staged the notebook still renders.
local({
  roots <- c(".", "assets/viz", Sys.getenv("SCRATCH_VIZ_DIR", "."))
  for (lib in c("scratch_viz.R", "scratch_report.R")) {
    for (r in roots) {
      p <- file.path(r, lib)
      if (file.exists(p)) { source(p, local = FALSE); break }
    }
  }
})
```

```{r setup, include = FALSE, message = FALSE, warning = FALSE, echo = FALSE}
suppressPackageStartupMessages({
  library(data.table); library(ggplot2); library(Seurat); library(SeuratObject)
})
project_name  <- params$project_name
work_directory <- params$work_directory
knitr::opts_knit$set(root.dir = work_directory)
for (d in c("data", "figures", "report")) dir.create(d, showWarnings = FALSE)
```

## Why this exists

inferCNV and Numbat answer the same question by different physics. inferCNV
smooths **expression** along the genome, so a cell whose transcriptional
programme happens to run high across a chromosome arm can look aneuploid.
Numbat phases reads at **heterozygous SNPs**, so it reads allelic imbalance
directly and can see copy-neutral LOH — where allele ratios shift but dosage
does not, and expression-based callers are blind by construction.

Two callers that fail differently are only useful if somebody looks at both.
Until this module existed, they were merged straight into a consensus by
STRATIFY and the disagreement disappeared into it: a run where the two callers
split on thousands of cells was indistinguishable, in every published output,
from a run where they agreed completely.

This notebook is **read-only**. It resolves nothing and changes no call. That is
deliberate — the comparison is only trustworthy if it cannot quietly become the
answer.

```{r load}
#| code-fold: true

obj  <- readRDS(params$seurat_object)
meta <- as.data.table(obj@meta.data, keep.rownames = "cell_id")
meta[, cell_id := as.character(cell_id)]

# Annotation column, same candidate order as STRATIFY so the two agree on what
# "cell type" means.
ct_candidates <- c(params$celltype_col, "azimuth_labels", "celltype", "cell_type",
                   "annotation", "sctype", "sctype_agg", "sctype_labels", "cell_label")
col_ct  <- ct_candidates[ct_candidates %in% names(meta)][1]
smp_candidates <- c(params$sample_col, "sample_id", "orig.ident", "sample", "patient_id")
col_smp <- smp_candidates[smp_candidates %in% names(meta)][1]

cat("Cells:", nrow(meta), "| cell-type column:", col_ct, "| sample column:", col_smp, "\n")

root <- params$caller_inputs
stage_files <- function(sub) {
  d <- file.path(root, sub)
  if (!dir.exists(d)) return(character(0))
  list.files(d, recursive = TRUE, full.names = TRUE)
}
```

## Reading the callers

```{r read-callers}
#| code-fold: true

# ---- inferCNV: a continuous genome-wide CNV load --------------------------
ic_files <- stage_files("infercnv")
ic <- NULL
if (length(ic_files)) {
  f <- ic_files[grepl("infercnv_meta_object\\.RDS$", ic_files)]
  if (!length(f)) f <- ic_files[grepl("\\.RDS$", ic_files)]
  if (length(f)) {
    o <- tryCatch(readRDS(f[1]), error = function(e) NULL)
    m <- if (inherits(o, "Seurat")) as.data.table(o@meta.data, keep.rownames = "cell_id")
         else if (is.data.frame(o)) as.data.table(o, keep.rownames = "cell_id") else NULL
    if (!is.null(m)) {
      sc <- intersect(c("genomwide_cnv_proportion", "genomewide_cnv_proportion",
                        "cnv_score", "cnv_load"), names(m))[1]
      if (!is.na(sc)) {
        ic <- m[, .(cell_id = as.character(cell_id), infercnv_score = as.numeric(get(sc)))]
      }
    }
  }
}

# ---- Numbat: a per-cell call plus its posterior ---------------------------
nb_files <- stage_files("numbat")
nb <- NULL
if (length(nb_files)) {
  tsv <- nb_files[grepl("_numbat_calls\\.tsv$", nb_files)]
  if (length(tsv)) {
    parts <- lapply(tsv, function(p) tryCatch(fread(p), error = function(e) NULL))
    parts <- Filter(function(x) !is.null(x) && nrow(x), parts)
    if (length(parts)) {
      nb <- rbindlist(parts, fill = TRUE)
      if ("cell_id" %in% names(nb)) {
        keep <- intersect(c("cell_id", "numbat_call", "p_cnv", "clone_opt", "compartment_opt"),
                          names(nb))
        nb <- unique(nb[, ..keep], by = "cell_id")
        nb[, cell_id := as.character(cell_id)]
      } else nb <- NULL
    }
  }
}

have_ic <- !is.null(ic) && nrow(ic) > 0
have_nb <- !is.null(nb) && nrow(nb) > 0
cat("inferCNV per-cell scores:", if (have_ic) nrow(ic) else 0, "\n")
cat("Numbat per-cell calls   :", if (have_nb) nrow(nb) else 0, "\n")

cmp <- meta[, .(cell_id, celltype = if (!is.na(col_ct)) as.character(get(col_ct)) else NA_character_,
                sample = if (!is.na(col_smp)) as.character(get(col_smp)) else "ALL")]
if (have_ic) cmp <- merge(cmp, ic, by = "cell_id", all.x = TRUE, sort = FALSE)
if (have_nb) cmp <- merge(cmp, nb, by = "cell_id", all.x = TRUE, sort = FALSE)
```

```{r one-caller-guard, results = 'asis'}
if (!have_ic || !have_nb) {
  missing <- c(if (!have_ic) "inferCNV", if (!have_nb) "Numbat")
  scratch_finding(sprintf(paste(
    "**Only one caller produced per-cell output**, so there is nothing to compare.",
    "Missing: **%s**. The compartment split downstream will rest on a single",
    "caller with no cross-check — the condition under which a degenerate",
    "threshold in that caller cannot be caught by anything."),
    paste(missing, collapse = ", ")),
    type = "caution", title = "No caller comparison possible")
}
```

## Do the callers rank cells the same way?

The threshold-free question, and the one worth asking first: **ignoring where
anyone drew a cutoff, does inferCNV score the cells Numbat calls malignant more
highly than the ones it calls diploid?** If it does, the callers agree about the
biology and any disagreement is about where the line sits. If it does not, they
disagree about the cells themselves, and no choice of threshold will reconcile
them.

```{r rank-agreement}
#| code-fold: true

auc <- NA_real_; n_pos <- 0L; n_neg <- 0L
if (have_ic && have_nb) {
  d <- cmp[!is.na(infercnv_score) & !is.na(numbat_call)]
  pos <- d[numbat_call == "malignant", infercnv_score]
  neg <- d[numbat_call == "non_malignant", infercnv_score]
  n_pos <- length(pos); n_neg <- length(neg)
  if (n_pos > 0 && n_neg > 0) {
    # AUC via the Mann-Whitney U statistic: the probability that a randomly
    # chosen Numbat-malignant cell scores higher on inferCNV than a randomly
    # chosen Numbat-diploid one. 0.5 is coin-flip, 1.0 is perfect separation.
    r <- rank(c(pos, neg))
    auc <- (sum(r[seq_len(n_pos)]) - n_pos * (n_pos + 1) / 2) / (as.numeric(n_pos) * n_neg)
  }
}
```

```{r rank-finding, results = 'asis'}
if (!is.na(auc)) {
  verdict <- if (auc >= 0.9) "the two callers order cells almost identically" else
             if (auc >= 0.7) "the callers broadly agree on ordering, with real overlap" else
             if (auc >= 0.55) "the callers agree only weakly on which cells carry CNV" else
             "**the callers do not agree on which cells carry CNV at all**"
  scratch_finding(sprintf(paste(
    "inferCNV scores a random Numbat-malignant cell above a random",
    "Numbat-diploid one **%.1f%%** of the time (AUC, n = %s vs %s): %s.",
    "This comparison uses no threshold, so it is unaffected by where either",
    "caller's cutoff was drawn."),
    100 * auc, format(n_pos, big.mark = ","), format(n_neg, big.mark = ","), verdict),
    type = if (!is.na(auc) && auc < 0.7) "caution" else "note",
    title = "Threshold-free agreement")
}
```

```{r fig-score-by-numbat, fig.width = 9, fig.height = 5}
#| label: fig-score-by-numbat
#| fig-cap: "inferCNV CNV load, split by Numbat's call"
if (have_ic && have_nb) {
  d <- cmp[!is.na(infercnv_score) & !is.na(numbat_call)]
  if (nrow(d)) {
    p <- ggplot(d, aes(x = numbat_call, y = infercnv_score, fill = numbat_call)) +
      geom_violin(scale = "width", colour = NA, alpha = .85) +
      geom_boxplot(width = .16, outlier.shape = NA, fill = "white", colour = "grey25") +
      scale_fill_scratch(guide = "none") +
      labs(title = "inferCNV CNV load, split by Numbat's call",
           subtitle = sprintf("Separation here is the agreement; AUC = %.3f", auc),
           x = "Numbat call", y = "inferCNV genome-wide CNV load") +
      theme_scratch()
    # The plot is the chunk VALUE, deliberately not print()-wrapped: ggplot2 4.0
    # dispatches print() through S7, which bypasses knit_print.ggplot — so a
    # print()ed plot renders STATIC and never reaches the PNG/PDF export hooked
    # into that method. This notebook's first render produced 3 static images and
    # an empty figures/ directory for exactly that reason.
    p
  }
} else cat("Needs both callers; skipped.\n")
```

## Where they disagree, cell by cell

```{r binarise}
#| code-fold: true

# inferCNV emits a continuous load, so a comparison against Numbat's categorical
# call needs a cutoff. This MIRRORS the rule STRATIFY uses — the null built from
# reference (non-malignant) lineages, mean + k*SD — so the confusion matrix below
# describes the same decision the pipeline actually makes. It is duplicated
# rather than shared because this module runs INSIDE the CNV stage, before
# STRATIFY exists; keep the two rules in step if either changes.
# The matcher and the cutoff rule come from assets/viz/scratch_report.R and are
# the SAME functions stratify_compartments.qmd uses. That is the point: this
# confusion matrix claims to describe the decision STRATIFY makes, so it has to
# be built from the identical rule. It was duplicated here at first; two copies
# of one threshold diverge the moment either is touched.
match_labels <- scratch_match_labels

ref_labels <- trimws(unlist(strsplit(as.character(params$reference_labels), "[;,|]")))
k <- as.numeric(params$infercnv_sd_threshold)

if (have_ic) {
  cmp[, is_ref := !is.na(celltype) & match_labels(celltype, ref_labels)]
  cmp[, infercnv_call := NA_character_]
  cut_basis <- list()
  for (s_i in unique(cmp$sample)) {
    d <- cmp[sample == s_i]
    r <- scratch_cnv_cutoff(d$infercnv_score, d$is_ref, k = k)
    cut_basis[[s_i]] <- r
    cmp[sample == s_i & !is.na(infercnv_score),
        infercnv_call := fifelse(infercnv_score > r$cutoff, "malignant", "non_malignant")]
  }
  cat("Cutoffs used (same rule as STRATIFY):\n")
  for (nm in names(cut_basis))
    cat(sprintf("  %-14s cutoff %.4f  n_ref %d  basis %s\n", nm,
                cut_basis[[nm]]$cutoff, cut_basis[[nm]]$n_ref, cut_basis[[nm]]$basis))
}
```

```{r fig-confusion, fig.width = 7.5, fig.height = 5}
#| label: fig-confusion
#| fig-cap: "inferCNV vs Numbat, per cell"
conf <- NULL
if (have_ic && have_nb) {
  d <- cmp[!is.na(infercnv_call) & !is.na(numbat_call)]
  if (nrow(d)) {
    conf <- d[, .N, by = .(infercnv_call, numbat_call)]
    conf[, frac := N / sum(N)]
    p <- ggplot(conf, aes(x = numbat_call, y = infercnv_call, fill = N)) +
      geom_tile(colour = "white", linewidth = 1.5) +
      geom_text(aes(label = sprintf("%s\n%.1f%%", format(N, big.mark = ","), 100 * frac)),
                colour = "grey15", size = 4) +
      scale_fill_gradient(low = "#eef3f9", high = "#3b6ea5", guide = "none") +
      labs(title = "inferCNV vs Numbat, per cell",
           subtitle = "Off-diagonal cells are where the two callers disagree",
           x = "Numbat call", y = "inferCNV call") +
      theme_scratch() + theme(panel.grid = element_blank())
    p
  }
} else cat("Needs both callers; skipped.\n")
```

```{r agreement-finding, results = 'asis'}
if (!is.null(conf) && nrow(conf)) {
  tot   <- sum(conf$N)
  agree <- sum(conf[infercnv_call == numbat_call, N])
  pct   <- 100 * agree / tot

  # Cohen's kappa: agreement corrected for what chance alone would produce. With
  # unbalanced compartments raw agreement flatters — 90% of cells non-malignant
  # means two callers that both say "non-malignant" always look 90% concordant.
  m <- dcast(conf, infercnv_call ~ numbat_call, value.var = "N", fill = 0)
  mm <- as.matrix(m[, -1, with = FALSE]); rownames(mm) <- m$infercnv_call
  common <- intersect(rownames(mm), colnames(mm))
  po <- if (length(common)) sum(diag(mm[common, common, drop = FALSE])) / sum(mm) else 0
  # pe uses only the classes both callers can express; when one caller emits a
  # SINGLE class its marginal is 1, pe collapses to 1, and (po-pe)/(1-pe) divides
  # by zero — the first render reported "Cohen's kappa -Inf" because inferCNV
  # called nothing malignant among the cells Numbat scored. Kappa is genuinely
  # undefined there, so report NA and say why rather than printing an infinity.
  pe <- sum((rowSums(mm) / sum(mm))[common] * (colSums(mm) / sum(mm))[common])
  kappa <- if (is.finite(pe) && pe < 1) (po - pe) / (1 - pe) else NA_real_
  one_class <- length(unique(d$infercnv_call)) < 2 || length(unique(d$numbat_call)) < 2

  only_ic <- conf[infercnv_call == "malignant" & numbat_call == "non_malignant", sum(N)]
  only_nb <- conf[infercnv_call == "non_malignant" & numbat_call == "malignant", sum(N)]

  kappa_txt <- if (is.na(kappa))
      "Cohen's kappa is **undefined** here: one caller assigned every shared cell to a single class, so there is no chance-agreement baseline to correct against."
    else sprintf("Cohen's kappa **%.2f**.%s", kappa,
      if (kappa < 0.4) " **Below 0.4 is poor agreement** — the raw percentage is carried by the majority class, not by the callers finding the same cells." else "")

  scratch_finding(sprintf(paste(
    "The callers agree on **%.1f%%** of the %s cells both scored. %s",
    "They disagree on **%s**: **%s** called malignant by inferCNV only,",
    "**%s** by Numbat only.%s"),
    pct, format(tot, big.mark = ","), kappa_txt,
    format(tot - agree, big.mark = ","),
    format(only_ic, big.mark = ","), format(only_nb, big.mark = ","),
    if (one_class) paste(" One caller made a single call across every shared cell.",
                         "Combined with a high rank AUC that is a THRESHOLD problem, not a",
                         "disagreement about which cells carry CNV.") else ""),
    type = if (is.na(kappa) || kappa < 0.4) "caution" else "note",
    title = "Caller agreement")
}
```

## What threshold would agree with Numbat?

inferCNV's cutoff is `mean + k * SD` of the reference-lineage null, with `k` set
by `--stratify_infercnv_sd`. Numbat needs no such choice — it calls from allelic
evidence directly. So Numbat's calls can be used to ask what `k` *would* have
agreed, which turns an arbitrary constant into a measured one.

This is an **advisory**. It reports the number; it does not change the threshold,
because doing so silently would move every downstream result without anything
saying why.

```{r k-calibration}
#| code-fold: true
kcal <- NULL
if (have_ic && have_nb && "is_ref" %in% names(cmp)) {
  per <- lapply(unique(cmp$sample), function(s_i) {
    d <- cmp[sample == s_i & !is.na(infercnv_score) & !is.na(numbat_call)]
    nulls <- cmp[sample == s_i & is_ref & is.finite(infercnv_score), infercnv_score]
    if (nrow(d) < 50 || length(nulls) < 30 || stats::sd(nulls) == 0) return(NULL)
    mu <- mean(nulls); sdv <- stats::sd(nulls)
    P <- sum(d$numbat_call == "malignant"); N <- sum(d$numbat_call == "non_malignant")
    if (!P || !N) return(NULL)
    grid <- seq(0, 6, by = 0.1)
    J <- vapply(grid, function(k_i) {
      thr <- mu + k_i * sdv
      tp <- sum(d$infercnv_score > thr & d$numbat_call == "malignant")
      fp <- sum(d$infercnv_score > thr & d$numbat_call == "non_malignant")
      tp / P - fp / N
    }, numeric(1))
    # The best k is often a PLATEAU, not a point: scores cluster, so a range of
    # thresholds separates the same cells. Report the plateau and take its
    # stringent end — the highest k that still achieves the best separation.
    best <- max(J)
    plateau <- grid[J >= best - 1e-9]
    data.table(sample = s_i, k_lo = min(plateau), k_hi = max(plateau),
               J = best, n_cells = nrow(d))
  })
  kcal <- rbindlist(Filter(Negate(is.null), per))
}
```

```{r k-calibration-finding, results = 'asis'}
if (!is.null(kcal) && nrow(kcal)) {
  cfg <- as.numeric(params$infercnv_sd_threshold)
  rec <- max(kcal$k_hi)
  scratch_finding(sprintf(paste(
    "Calibrated against Numbat, the SD multiplier that best separates its calls is",
    "**k = %s**%s (Youden J %.3f). The pipeline is configured at **k = %s**.%s"),
    paste(sprintf("%.1f", unique(kcal$k_hi)), collapse = " / "),
    if (any(kcal$k_hi > kcal$k_lo))
      sprintf(" (a plateau from %.1f; the stringent end is quoted)", min(kcal$k_lo)) else "",
    max(kcal$J), format(cfg, digits = 2),
    if (abs(rec - cfg) >= 0.5)
      sprintf(paste(" **That is %.1f SD from the configured value.** At k = %s the",
                    "threshold and Numbat disagree materially; consider",
                    "--stratify_infercnv_sd %.1f."),
              abs(rec - cfg), format(cfg, digits = 2), rec)
    else " The configured value is close to the calibrated one."),
    type = if (abs(rec - cfg) >= 0.5) "caution" else "note",
    title = "Calibrated CNV threshold")

  scratch_table(kcal[, .(sample, `best k (low)` = k_lo, `best k (high)` = k_hi,
                         `Youden J` = round(J, 3), cells = n_cells)],
                caption = "SD multiplier calibrated against Numbat, per sample")

  if (nrow(kcal) == 1)
    scratch_finding(paste("Only **one sample** had both an inferCNV score and Numbat",
      "calls, so this calibration rests on that sample alone. Treat it as",
      "indicative until a second sample confirms it."),
      type = "caution", title = "Calibration is single-sample")
}
```

## Disagreement by cell type

Where the callers split is more informative than how often. Disagreement
concentrated in the tumour lineage is a threshold argument; disagreement spread
across immune lineages means one caller is calling normal cells aneuploid.

```{r fig-agreement-by-celltype, fig.width = 9.5, fig.height = 5.5}
#| label: fig-agreement-by-celltype
#| fig-cap: "Per-cell-type agreement between the callers"
bycell <- NULL
if (have_ic && have_nb && !is.na(col_ct)) {
  d <- cmp[!is.na(infercnv_call) & !is.na(numbat_call) & !is.na(celltype)]
  if (nrow(d)) {
    bycell <- d[, .(n = .N,
                    agree = sum(infercnv_call == numbat_call),
                    infercnv_only = sum(infercnv_call == "malignant" & numbat_call == "non_malignant"),
                    numbat_only   = sum(infercnv_call == "non_malignant" & numbat_call == "malignant")),
                by = celltype]
    bycell[, pct_agree := round(100 * agree / n, 1)]
    setorder(bycell, -n)

    pl <- melt(bycell[, .(celltype, agree, infercnv_only, numbat_only)],
               id.vars = "celltype", variable.name = "outcome", value.name = "cells")
    pl[, outcome := factor(outcome, levels = c("agree", "infercnv_only", "numbat_only"),
                           labels = c("agree", "inferCNV only", "Numbat only"))]
    p <- ggplot(pl, aes(x = reorder(celltype, cells, sum), y = cells, fill = outcome)) +
      geom_col(width = .72) + coord_flip() +
      scale_fill_scratch(name = NULL) +
      labs(title = "Per-cell-type agreement between the CNV callers",
           subtitle = "Bars are cells; colour is which caller called malignant when they differed",
           x = NULL, y = "cells") +
      theme_scratch()
    p
  }
} else cat("Needs both callers and a cell-type column; skipped.\n")
```

```{r bycell-finding, results = 'asis'}
if (!is.null(bycell) && nrow(bycell)) {
  worst <- bycell[order(pct_agree)][1]
  scratch_finding(sprintf(paste(
    "Lowest agreement is in **%s** (%.1f%% over %s cells: %s inferCNV-only,",
    "%s Numbat-only). Highest is **%s** at %.1f%%."),
    worst$celltype, worst$pct_agree, format(worst$n, big.mark = ","),
    format(worst$infercnv_only, big.mark = ","), format(worst$numbat_only, big.mark = ","),
    bycell[order(-pct_agree)][1]$celltype, bycell[order(-pct_agree)][1]$pct_agree),
    type = "note", title = "Where the callers diverge")
  scratch_table(bycell[, .(celltype, cells = n, agree, `inferCNV only` = infercnv_only,
                           `Numbat only` = numbat_only, `% agree` = pct_agree)],
                caption = "Caller agreement per annotated cell type")
}
```

## Coverage

A caller that scored no cells cannot disagree with anything, and a comparison
run over a small overlap says little about the rest.

```{r coverage, results = 'asis'}
n_all  <- nrow(cmp)
n_both <- if (have_ic && have_nb) cmp[!is.na(infercnv_score) & !is.na(numbat_call), .N] else 0L
scratch_finding(sprintf(paste(
  "inferCNV scored **%s** of %s cells; Numbat called **%s**. Both scored the",
  "same **%s** (%.1f%%) — the comparison above covers only those."),
  format(if (have_ic) cmp[!is.na(infercnv_score), .N] else 0L, big.mark = ","),
  format(n_all, big.mark = ","),
  format(if (have_nb) cmp[!is.na(numbat_call), .N] else 0L, big.mark = ","),
  format(n_both, big.mark = ","), 100 * n_both / max(1L, n_all)),
  type = if (n_both < 0.5 * n_all) "caution" else "note",
  title = "Comparison coverage")
```

## Saving

```{r save}
if (isTRUE(as.logical(params$auto_save))) {
  out_cols <- intersect(c("cell_id", "sample", "celltype", "infercnv_score",
                          "infercnv_call", "numbat_call", "p_cnv", "clone_opt"),
                        names(cmp))
  fwrite(cmp[, ..out_cols],
         file.path("data", paste0(project_name, "_cnv_caller_comparison.csv")))

  summary_dt <- data.table(
    metric = c("cells_total", "cells_infercnv", "cells_numbat", "cells_both",
               "pct_agree", "cohens_kappa", "auc_rank", "infercnv_only", "numbat_only"),
    value  = c(n_all,
               if (have_ic) cmp[!is.na(infercnv_score), .N] else 0L,
               if (have_nb) cmp[!is.na(numbat_call), .N] else 0L,
               n_both,
               if (exists("pct")) round(pct, 2) else NA_real_,
               if (exists("kappa")) round(kappa, 4) else NA_real_,
               if (!is.na(auc)) round(auc, 4) else NA_real_,
               if (exists("only_ic")) only_ic else NA_integer_,
               if (exists("only_nb")) only_nb else NA_integer_))
  fwrite(summary_dt, file.path("data", paste0(project_name, "_cnv_caller_summary.csv")))
  if (!is.null(bycell)) fwrite(bycell, file.path("data", paste0(project_name, "_cnv_agreement_by_celltype.csv")))

  # ---- The cumulative object ------------------------------------------------
  # Write both callers' per-cell results back onto the object before handing it
  # downstream. Without this the CNV stage is a dead end for the object: its
  # results live only in side files, STRATIFY re-reads them, and everything the
  # object gained upstream has to be re-joined at the very end by INTEGRATION.
  add <- cmp[match(colnames(obj), cmp$cell_id)]
  for (cn in intersect(c("infercnv_score", "infercnv_call", "numbat_call",
                         "p_cnv", "clone_opt"), names(add))) {
    obj[[cn]] <- add[[cn]]
  }
  saveRDS(obj, file.path("data", paste0(project_name, "_cnv_object.RDS")))
  cat("Saved CNV-enriched object with columns:",
      paste(intersect(c("infercnv_score","infercnv_call","numbat_call","p_cnv","clone_opt"),
                      colnames(obj@meta.data)), collapse = ", "), "\n")
}
```

## Session info

```{r session, echo = FALSE}
sessionInfo()
```
